---
title: "Sta 440 Case 4"
author: "Olivia Fu, Yura Heo, Jack Steel"
date: "November 2025"
format: 
  pdf:
    documentclass: article
    fontsize: 11pt
geometry: margin=1in
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
load("./rootData.RData")

library(ggplot2)
library(dplyr)
library(nlme)
library(broom)
library(knitr)
library(stringr)
library(tidyverse)
```


# Background

Root circumnutation, the helical or oscillatory growth movement of roots, has long been hypothesized to arise from differential cell elongation on opposite sides of the root tip. This phenomenon plays an essential role in how roots navigate the soil environment, influencing their ability to penetrate obstacles, anchor the plant, and optimize nutrient uptake. To investigate the biological mechanisms underlying circumnutation, we compare the spatial growth patterns between wild-type and mutant genotypes that differ in their circumnutation behavior. Specifically, we model root elongation along the midline across inner and outer sides of the root to determine whether growth asymmetry explains the observed movement and whether this asymmetry is reduced in mutants.

TODO: talk about EDA and main model

# Exploratory Data Analysis


```{r}
plot1 <- ggplot(rootData, aes(x = side, y = Length, fill = side)) +
  geom_boxplot() +
  facet_wrap(~ Genotype) +
  labs(title = "Cell Length Distribution by Inner vs Outer",
       y = "Cell Length") +
  theme_bw()
```

The roots are classified as mutant or wild type, and their cell lengths on the inner side and outer side of the roots are measured at various distances along the midline. The distribution for the cell length is right skewed for both sides and both root types, with the mutant type being slightly stronger right skewed (@fig-cell-length-dist). The inner and outer side for the mutant type have a nearly identical distribution, with a median cell length around 9. On the other hand, the distributions for the cell length for the wild type differ between the inner measurements and outer measurements, where the outer cell length measurements are slightly larger than the inner measurements. 

```{r}
rootData_WT <- rootData |> filter(Genotype == "WT")

plot2 <- ggplot(rootData_WT, aes(x = Midline, y = Length, 
                        color = side, group = interaction(root, side))) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "Cell Length Along Midline – Wild Type Roots",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

rootData_MU <- rootData |> filter(Genotype == "MU")
desired_order <- c("MU1", "MU2", "MU3", "MU4", "MU5", "MU6", "MU7", "MU8", "MU9", "MU10", "MU11")
rootData_MU <- rootData_MU |>
  mutate(root = factor(root, levels = desired_order))

plot3 <- ggplot(rootData_MU, aes(x = Midline, y = Length, 
                        color = side, group = interaction(root, side))) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "Cell Length Along Midline – Mutant Roots",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The trends for the wild type roots in @fig-cell-length-per-root-wt and @fig-cell-length-per-root-smooth-wt show a large gap between the lines for the inner and outer measurements, while the trends for the mutant type roots in @fig-cell-length-per-root-mu and @fig-cell-length-per-root-smooth-mu show a small gap between the lines for the inner and outer measurements. Wild type roots one, three, and eight particularly have higher inner measurements than outer measurements. On the other hand, mutant type roots have intersecting lines for inner and outer measurements as distance along the midline increases. This is further emphasized by the smooth line plots in @fig-cell-length-smooth-wt and @fig-cell-length-smooth-mu - the wild type has noticeably different inner and outer measurements for cell length. 


```{r}
plot4 <- ggplot(rootData_WT,
       aes(x = Midline, y = Length, color = side, group = interaction(root, side))) +
  geom_smooth(method = "loess", se = FALSE, span = 0.75, size = 0.8) +
  geom_point(alpha = 0.4, size = 0.5) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "WT: Inner vs Outer Smoothed Trends for Each Root",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot5 <- ggplot(rootData_MU,
       aes(x = Midline, y = Length, color = side, group = interaction(root, side))) +
  geom_smooth(method = "loess", se = FALSE, span = 0.75, size = 0.8) +
  geom_point(alpha = 0.4, size = 0.5) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "MT: Inner vs Outer Smoothed Trends for Each Root",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
plot6 <- ggplot(rootData_WT, aes(x = Midline, y = Length, color = side)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.8, size = 1.2) +
  labs(
    title = "WT: Inner vs Outer Overall Cell Length Trends",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw()

plot7 <- ggplot(rootData_MU, aes(x = Midline, y = Length, color = side)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.8, size = 1.2) +
  labs(
    title = "MU: Inner vs Outer Overall Cell Length Trends",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw()
```


```{r}
logis_smooth <- function(df) {
  ggplot(
    df,
    aes(x = Midline, y = Length, color = side, group = interaction(root, side))
  ) +
    geom_point(alpha = 0.4, size = 0.5) +
    geom_smooth(
      method = "nls",
      formula = y ~ SSlogis(x, Asym, xmid, scal),
      method.args = list(control = nls.control(maxiter = 200, warnOnly = TRUE)),
      se = FALSE,
      size = 0.8
    ) +
    facet_wrap(~ root, scales = "free_y") +
    labs(
      x = "Distance Along Midline",
      y = "Cell Length",
      color = "Side"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Wild type
plot8 <- logis_smooth(rootData_WT) + ggtitle("WT: Inner vs Outer Logistic Fits for Each Root")

# Mutant
plot9 <- logis_smooth(rootData_MU) + ggtitle("MU: Inner vs Outer Logistic Fits for Each Root")
```

#### Discuss Logistic 

```{r}
plot10 <- rootData_WT |>
  ggplot(aes(x=root, y=Length, fill=side)) +
  geom_bar(stat="summary", fun="mean", position="dodge") +
  labs(
    x = "Root", 
    y = "Mean Cell Length",
    title = "Mean Cell Length for Each Wild Type Root"
  )

plot11 <- rootData_MU |>
  ggplot(aes(x=root, y=Length, fill=side)) +
  geom_bar(stat="summary", fun="mean", position="dodge") +
  labs(
    x = "Root", 
    y = "Mean Cell Length",
    title = "Mean Cell Length for Each Mutant Type Root"
  )
```

The mean cell length across each root for mutant type has similar measurements for the inside and the outside cells, besides for mutant root nine, where the mean inner measurement seems to be larger than the mean outer cell length measurement (@fig-mean-cell-length-per-root-mu). On the other hand, the mean outer cell length is larger for every wild type root besides root seven, where the mean inner measurement is slightly larger than the mean outer measurement (@fig-mean-cell-length-per-root-wt). 

# Model Rationale

# Model Implementation and Evaluation

# Results

# Limitations and Conclusion

# Appendix 
```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-dist
#| fig-cap: "Distribution of cell length for inner and outer cells."
plot1
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-wt
#| fig-cap: "Distribution of cell length for each wild root."
plot2
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-mu
#| fig-cap: "Distribution of cell length for each mutant root."
plot3
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-smooth-wt
#| fig-cap: "Distribution of cell length for each wild root."
plot4
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-smooth-mu
#| fig-cap: "Distribution of cell length for each mutant root."
plot5
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-smooth-wt
#| fig-cap: "Distribution of cell length for wild type."
plot6
```
```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-smooth-wt
#| fig-cap: "Distribution of cell length for mutant type."
plot7
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-smooth-wt
#| fig-cap: "Logistic fit of cell length for wild type."
plot8
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-smooth-wu
#| fig-cap: "Logistic fit of cell length for mutant type."
plot9
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-mean-cell-length-per-root-wt
#| fig-cap: "Distribution of mean cell length for each wild root."
plot10
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-smooth-mu
#| fig-cap: "Distribution of mean cell length for each mutant root."
plot11
```

