---
title: "Sta 440 Case 4"
author: "Olivia Fu, Yura Heo, Jack Steel"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format: 
  pdf:
    documentclass: article
    fontsize: 11pt
geometry: margin=1in
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
load("./rootData.RData")

library(ggplot2)
library(patchwork)
library(dplyr)
library(nlme)
library(broom)
library(knitr)
library(stringr)
library(tidyverse)
library(purrr)
library(kableExtra)
library(tidyr)
```

# Background

Root circumnutation, the helical or oscillatory growth movement of roots, has been hypothesized to result from differential cell elongation on opposing sides of the root tip. This growth pattern enables roots to navigate the soil, penetrate obstacles, and optimize anchorage and nutrient uptake. To explore this mechanism, we compare spatial growth patterns between wild-type and mutant genotypes that differ in their circumnutation behavior. Specifically, we model root elongation along the midline across inner and outer sides to test whether growth asymmetry drives circumnutation and whether this asymmetry is reduced in mutants.

This analysis aims to quantify the extent of asymmetry in cell length along the inner and outer curves of individual roots and to assess how these patterns differ between genotypes. By fitting both frequentist and hierarchical Bayesian nonlinear regression models, we seek to: (1) identify within-root growth differences linked to circumnutation, (2) characterize variability across roots, and (3) evaluate whether the mutant genotype exhibits reduced differential growth consistent with its impaired movement phenotype.

# Exploratory Data Analysis

```{r}
plot1 <- ggplot(rootData, aes(x = side, y = Length, fill = side)) +
  geom_boxplot() +
  facet_wrap(~ Genotype) +
  labs(title = "Cell Length Distribution by Inner vs Outer",
       y = "Cell Length") +
  theme_bw()
```

The roots are classified as mutant or wild type, and their cell lengths on the inner side and outer side of the roots are measured at various distances along the midline. The distribution for the cell length is right skewed for both sides and both root types, with the mutant type being slightly stronger right skewed (@fig-cell-length-dist). The inner and outer side for the mutant type have a nearly identical distribution, with a median cell length around 9. On the other hand, the distributions for the cell length for the wild type differ between the inner measurements and outer measurements, where the outer cell length measurements are slightly larger than the inner measurements.

```{r}
rootData_WT <- rootData |> filter(Genotype == "WT")

plot2 <- ggplot(rootData_WT, aes(x = Midline, y = Length, 
                        color = side, group = interaction(root, side))) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "Cell Length Along Midline – Wild Type Roots",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

rootData_MU <- rootData |> filter(Genotype == "MU")
desired_order <- c("MU1", "MU2", "MU3", "MU4", "MU5", "MU6", "MU7", "MU8", "MU9", "MU10", "MU11")
rootData_MU <- rootData_MU |>
  mutate(root = factor(root, levels = desired_order))

plot3 <- ggplot(rootData_MU, aes(x = Midline, y = Length, 
                        color = side, group = interaction(root, side))) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "Cell Length Along Midline – Mutant Roots",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The trends for the wild type roots in @fig-cell-length-per-root-wt and @fig-cell-length-per-root-smooth-wt show a large gap between the lines for the inner and outer measurements, while the trends for the mutant type roots in @fig-cell-length-per-root-mu and @fig-cell-length-per-root-smooth-mu show a small gap between the lines for the inner and outer measurements. Wild type roots one, three, and eight particularly have higher inner measurements than outer measurements. On the other hand, mutant type roots have intersecting lines for inner and outer measurements as distance along the midline increases. This is further emphasized by the smooth line plots in @fig-cell-length-smooth-wt and @fig-cell-length-smooth-mu - the wild type has noticeably different inner and outer measurements for cell length.

```{r}
plot4 <- ggplot(rootData_WT,
       aes(x = Midline, y = Length, color = side, group = interaction(root, side))) +
  geom_smooth(method = "loess", se = FALSE, span = 0.75, size = 0.8) +
  geom_point(alpha = 0.4, size = 0.5) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "WT: Inner vs Outer Smoothed Trends for Each Root",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot5 <- ggplot(rootData_MU,
       aes(x = Midline, y = Length, color = side, group = interaction(root, side))) +
  geom_smooth(method = "loess", se = FALSE, span = 0.75, size = 0.8) +
  geom_point(alpha = 0.4, size = 0.5) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "MT: Inner vs Outer Smoothed Trends for Each Root",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
plot6 <- ggplot(rootData_WT, aes(x = Midline, y = Length, color = side)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.8, size = 1.2) +
  labs(
    title = "WT: Inner vs Outer Overall Cell Length Trends",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw()

plot7 <- ggplot(rootData_MU, aes(x = Midline, y = Length, color = side)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.8, size = 1.2) +
  labs(
    title = "MU: Inner vs Outer Overall Cell Length Trends",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw()
```

```{r}
plot10 <- rootData_WT |>
  ggplot(aes(x=root, y=Length, fill=side)) +
  geom_bar(stat="summary", fun="mean", position="dodge") +
  labs(
    x = "Root", 
    y = "Mean Cell Length",
    title = "Mean Cell Length for Each Wild Type Root"
  )

plot11 <- rootData_MU |>
  ggplot(aes(x=root, y=Length, fill=side)) +
  geom_bar(stat="summary", fun="mean", position="dodge") +
  labs(
    x = "Root", 
    y = "Mean Cell Length",
    title = "Mean Cell Length for Each Mutant Type Root"
  )
```

The mean cell length across each root for mutant type has similar measurements for the inside and the outside cells, besides for mutant root nine, where the mean inner measurement seems to be larger than the mean outer cell length measurement (@fig-mean-cell-length-per-root-mu). On the other hand, the mean outer cell length is larger for every wild type root besides root seven, where the mean inner measurement is slightly larger than the mean outer measurement (@fig-mean-cell-length-per-root-wt).

# Model Rationale and Implementation

## Frequentist Model

### Model Rationale

Our EDA showed that cell length increases nonlinearly from the root tip toward the base, consistent with a sigmoidal growth pattern. These findings motivated a restricted logistic model structure that captures both the saturating nature of growth and the side-specific shift in elongation patterns through the inclusion of a side indicator term ($\beta$) to quantify asymmetry in elongation between the inner and outer sides. Because the data were collected separately for each side of the root and for two genotypes (wild type and mutant), this formulation directly reflects the experimental design and allows side- and genotype-level comparisons. The logistic form also aligns with biological constraints: cell lengths cannot be negative and are expected to plateau as elongation slows near the root base. $$
\text{Length} \sim A + \frac{(B - A)}{1 + \exp\left(\frac{x_{\text{mid}} + \beta \cdot \text{side} - x}{\text{scal}}\right)}
$$Here, A and B are lower and upper asymptotes of cell length, $x_{mid}$ is the inflection point, $scal$ controls curve steepness, and $\beta$ shifts the midpoint between inner (0) and outer (1) sides, indicating differential growth.

### Model Implementation

Nonlinear least squares (NLS) models were fit separately for each root with a custom logistic function. Parameter bounds were set to ensure biologically realistic ranges, and starting values were derived from each root's observed cell length distribution. A total of 9 wild-type and 11 mutant roots were models independently to assess within-root and between-genotype differences in elongation patterns.

```{r}
fpl_customized <- function(x, A, B, xmid, scal, beta, side_ind){
  xmid_eff <- xmid + beta * side_ind
  A + (B - A) / (1 + exp((xmid_eff - x) / scal))
}
```

```{r}
WT_model_output <- list()

for (root_num in unique(rootData_WT$root)) {
  curr_root <- rootData_WT |> filter(root == root_num)
  curr_root$side_ind <- ifelse(curr_root$side == "Outer", 1, 0)

  y_min <- min(curr_root$Length)
  y_max <- max(curr_root$Length)

  start_vals <- list(
    A = y_min * 0.9,
    B = y_max * 1.1,
    xmid = median(curr_root$Midline),
    scal = diff(range(curr_root$Midline)) / 4,
    beta = -50
  )

  lower_bounds <- c(A = y_min * 0.5,
                    B = y_min,
                    xmid = min(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)) / 10,
                    beta = -500)

  upper_bounds <- c(A = y_max,
                    B = y_max * 2.0,
                    xmid = max(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)),
                    beta = 500)

  fit <- try(
    nls(
      Length ~ fpl_customized(Midline, A, B, xmid, scal, beta, side_ind),
      data = curr_root,
      start = start_vals,
      algorithm = "port",
      lower = lower_bounds,
      upper = upper_bounds,
      control = nls.control(maxiter = 1000, warnOnly = TRUE, minFactor = 1e-10)
    ),
    silent = TRUE
  )

  if (!inherits(fit, "try-error")) {
    WT_model_output[[paste0("root_", root_num)]] <- fit
  } else {
    message(paste("Model failed for root", root_num))
  }
}
```

```{r}
MU_model_output <- list()

for (root_num in unique(rootData_MU$root)) {
  curr_root <- rootData_MU |> filter(root == root_num)
  curr_root$side_ind <- ifelse(curr_root$side == "Outer", 1, 0)

  y_min <- min(curr_root$Length)
  y_max <- max(curr_root$Length)

  start_vals <- list(
    A = y_min * 0.9,
    B = y_max * 1.1,
    xmid = median(curr_root$Midline),
    scal = diff(range(curr_root$Midline)) / 4,
    beta = 0
  )

  lower_bounds <- c(A = y_min * 0.5,
                    B = y_min,
                    xmid = min(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)) / 10,
                    beta = -500)

  upper_bounds <- c(A = y_max,
                    B = y_max * 2.0,
                    xmid = max(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)),
                    beta = 500)

  fit <- try(
    nls(
      Length ~ fpl_customized(Midline, A, B, xmid, scal, beta, side_ind),
      data = curr_root,
      start = start_vals,
      algorithm = "port",
      lower = lower_bounds,
      upper = upper_bounds,
      control = nls.control(maxiter = 1000, warnOnly = TRUE, minFactor = 1e-10)
    ),
    silent = TRUE
  )

  if (!inherits(fit, "try-error")) {
    MU_model_output[[paste0("root_", root_num)]] <- fit
  } else {
    message(paste("Model failed for root", root_num))
  }
}
```

### Model Evaluation

```{r}
WT_residuals <- data.frame()

for (name in names(WT_model_output)) {
  model <- WT_model_output[[name]]
  root_label <- str_replace(name, "root_", "")
    curr_data <- rootData_WT |> filter(root == root_label)
  
  curr_data <- curr_data |>
    mutate(
      Fitted = fitted(model),
      Residuals = resid(model)
    )
  
  WT_residuals <- bind_rows(WT_residuals, curr_data)
}

plot12 <- ggplot(WT_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (WT)", x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

plot13 <- ggplot(WT_residuals, aes(x = Midline, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (WT)", x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()
```

```{r}
MU_residuals <- data.frame()

for (name in names(MU_model_output)) {
  model <- MU_model_output[[name]]
  root_label <- str_replace(name, "root_", "")
    curr_data <- rootData_MU |> filter(root == root_label)
  
  curr_data <- curr_data |>
    mutate(
      Fitted = fitted(model),
      Residuals = resid(model)
    )
  
  MU_residuals <- bind_rows(MU_residuals, curr_data)
}

plot15 <- ggplot(MU_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (MU)", x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

plot16 <- ggplot(MU_residuals, aes(x = Midline, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (MU)", x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()
```

```{r}
all_residuals <- bind_rows(WT_residuals, MU_residuals)

# Ensure proper root order: WT1..WT9, MU1..MU11
all_residuals <- all_residuals %>%
  mutate(root = factor(root, 
                       levels = c(paste0("WT", 1:9), paste0("MU", 1:11))))

# Plot combined residual boxplot
p_all_box <- ggplot(all_residuals, aes(x = root, y = Residuals, fill = side)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(title = "Residuals by Root (Frequentist Model)",
       x = "Root ID", y = "ResidualS", fill = "Side") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )
```

```{r}
# ACF plot (autocorr check)
m0 <- lm(Length ~ side * poly(Midline, 2), data = rootData)
```

The residual versus fitted plots (@fig-resid-freq) for the wild type and mutant type roots show a fanning pattern while the residuals vs. midline plots (@fig-resid-freq) for the wild type and mutant type roots are scattered and evenly distanced along the midline. This shows that our frequentist model was a good fit for both the mutant type root data and the wild type root data.

The ACF plot (@fig-freq-acf) suggests that there is autocorrelation that our model might not necessarily account for and that our model may have not captured the patterns in the data, as the bars exceed the bounds.

## Bayesian Model

### Model Rationale

We modeled cell length as normally distributed around a logistic growth curve, with root-specific random effects and covariate effects for genotype and side:

$$
\text{Length}_i \sim \mathcal{N}(\mu_i, \sigma^2),
$$

where $\mu_i$ represents the expected cell length for observation $i$. The distance along the root midline was normalized to $[0,1]$ to ensure comparability across roots of different lengths.

Guided by our exploratory data analysis and frequentist results, we modeled the mean cell length with a four-parameter logistic curve, where for each root $r$, $A_r$ denotes the lower asymptote (baseline cell length), $B_r$ the upper asymptote (maximum cell length), $x_{\text{mid},i}$ the effective inflection point (location of fastest growth), and $\text{scal}_r$ the steepness parameter (slope of transition):

$$
\mu_i = A_{r[i]} + \frac{B_{r[i]} - A_{r[i]}}{1 + \exp\left(\frac{x_{\text{mid},i} - x_{01,i}}{\text{scal}_{r[i]}}\right)}.
$$

To capture differences between wildtype vs mutant and inner vs outer root sides, we modeled the effective midpoint as a function of side and genotype effects, where $\alpha_r$ represents the baseline midpoint for wildtype inner and mutant inner sides, $\gamma_1$ captures the shift between outer and inner sides in wildtype roots, and $\gamma_2$ captures the additional shift for mutant outer roots. This structure allows side and genotype effects to interact, enabling direct evaluation of the difference between inner and outer sides in wildtype roots and whether this difference diminishes in mutant roots:

$$
x_{\text{mid},i} = \alpha_{r[i]} + \gamma_1 \cdot \text{side}_i + \gamma_2 \cdot \text{side}_i \cdot \text{geno}_i.
$$

To represent uncertainty about side and genotype effects, we placed normal priors centered at zero, $\gamma_1, \gamma_2 \sim \mathcal{N}(0, 0.5^2)$, corresponding to a standard deviation of 0.5 on the normalized $x$-axis (i.e., expected shifts less than half a root length). Residual variation was modeled via a half-$t$ prior, $\sigma \sim \text{Half-}t(0, 3, \nu = 3)$, which provides heavier tails than a uniform prior – allowing large residuals if supported but concentrating most mass around moderate values.

To account for biological variability in the four logistic parameters, we included root-level random effects, $a_j \sim \mathcal{N}(\mu_a, \sigma_a^2)$, $d_j \sim \mathcal{N}(\mu_d, \sigma_d^2)$, $z_{0,j} \sim \mathcal{N}(\mu_{z0}, \sigma_{z0}^2)$, and $\eta_{s,j} \sim \mathcal{N}(\mu_{\eta_s}, \sigma_{\eta_s}^2)$. These were transformed to ensure biologically meaningful constraints: $A_j = e^{a_j} > 0$, $B_j = A_j + e^{d_j} > A_j$, $\alpha_j = \text{logit}^{-1}(z_{0,j}) \in (0,1)$, and $\text{scal}_j = e^{\eta_{s,j}} > 0$.

We used weakly informative, data-anchored hyperpriors to guide estimation: $\mu_a \sim \mathcal{N}(\log L_{\text{lower}}, 1)$, $\mu_d \sim \mathcal{N}(\log L_{\text{span}}, 1)$, $\mu_{z0} \sim \mathcal{N}(0, 1)$, and $\mu_{\eta_s} \sim \mathcal{N}(\log 0.2, 0.5)$, where $L_{\text{lower}}$ and $L_{\text{span}}$ are empirical lower and span length scales from the data. To allow moderate between-root variability, we set the precisions following $\text{Gamma}(1, 1)$ priors.

### Model Implementation

We initialized the Bayesian model using estimates from the frequentist nonlinear least squares (NLS) fits. For each root, the NLS point estimates and standard errors for the four logistic parameters ($A$, $B$, $x_{\text{mid}}$, and $\text{scal}$) were used to generate dispersed starting values. Specifically, for each chain, we drew initial values from Normal distributions centered at the NLS estimates with standard deviations set to five times the corresponding NLS standard errors. This overdispersion ensured that the chains started from diverse points in the parameter space rather than being overly constrained by the frequentist results. The starting values were then transformed to the latent parameters ($a$, $d$, $z_0$, $\eta_s$) as described previously. The fixed effects ($\gamma_1$, $\gamma_2$) were initialized near zero with small random noise ($\text{sd} = 0.01$) to maintain variability without imposing strong directional bias. The residual scale parameter ($t_\sigma$) was initialized from a $\text{Normal}(0, 3)$ distribution, consistent with its half-$t$ prior specification. We ran four MCMC chains with 50,000 iterations each, discarding the first 10,000 as burn-in and using no thinning.

```{r}
library(rjags)
library(R2jags)
library(coda)
```

```{r}
# Normalize Midline to the [0,1] range
rootData <- rootData |>
  group_by(root) |>
  mutate(x01 = (Midline - min(Midline)) / (max(Midline) - min(Midline))) |>
  ungroup()
```

```{r}
L_lower <- min(rootData$Length, na.rm = TRUE)
L_upper <- max(rootData$Length, na.rm = TRUE)
L_span  <- L_upper - L_lower

bdat <- list(
  Length = rootData$Length,
  N = nrow(rootData),
  x01 = rootData$x01,
  side = as.numeric(rootData$side == "Outer"),
  geno = as.numeric(rootData$Genotype == "MU"),
  root = as.numeric(as.factor(rootData$root)),
  n.root = length(unique(rootData$root)),
  L_lower = L_lower,
  L_span  = L_span
)

fun.params <- c("A", "B", "alpha", "scal",
                "gamma1", "gamma2",
                "sigma",
                "mu_a", "mu_d", "mu_z0", "mu_eta_s",
                "prec_a", "prec_d", "prec_z0", "prec_eta_s")
```

```{r}
get_nls_summary <- function(model) {
  s <- summary(model)
  est <- s$coefficients[, 1]
  se  <- s$coefficients[, 2]
  tibble(A = est["A"], A_se = se["A"],
         B = est["B"], B_se = se["B"],
         scal = est["scal"], scal_se = se["scal"],
         xmid = est["xmid"], xmid_se = se["xmid"])
}

WT_results_full <- map_dfr(WT_model_output, get_nls_summary, .id = "root")
MU_results_full <- map_dfr(MU_model_output, get_nls_summary, .id = "root")
all_results <- bind_rows(WT_results_full, MU_results_full)
```

```{r}
root_levels <- levels(as.factor(rootData$root))
nroot <- length(root_levels)
A_init <- B_init <- scal_init <- alpha_init <- numeric(nroot)
A_sd <- B_sd <- scal_sd <- alpha_sd <- numeric(nroot)

for (i in seq_along(root_levels)) {
  this_name <- paste0("root_", root_levels[i])
  row_i <- all_results[all_results$root == this_name, ]
  rd_i <- filter(rootData, root == root_levels[i])

  # we have fits for every root
  A_init[i] <- row_i$A
  A_sd[i]   <- row_i$A_se
  B_init[i] <- row_i$B
  B_sd[i]   <- row_i$B_se

  # Get min and max for this root
  mid_min <- min(rd_i$Midline)
  mid_max <- max(rd_i$Midline)
  span    <- mid_max - mid_min

  # Normalize both xmid and scal
  alpha_init[i] <- (row_i$xmid - mid_min) / span
  alpha_sd[i]   <- row_i$xmid_se / span
  scal_init[i]  <- row_i$scal / span
  scal_sd[i]    <- row_i$scal_se / span
}
```

```{r}
fun.inits <- function(chain_id = 1) {
  t_sigma.start <- rnorm(1, mean = 0, sd = 3)

  # Overdisperse latent parameters around NLS estimates
  a.start     <- log(pmax(1e-3, rnorm(bdat$n.root, mean = A_init, sd = 5 * A_sd)))
  d.start     <- log(pmax(1e-3, rnorm(bdat$n.root, mean = B_init - A_init, sd = 5 * B_sd)))
  z0.start    <- qlogis(
    pmin(0.99, pmax(0.01, rnorm(bdat$n.root, mean = alpha_init, sd = 5 * alpha_sd))))
  eta_s.start <- log(pmax(1e-3, rnorm(bdat$n.root, mean = scal_init, sd = 5 * scal_sd)))

  # Fixed effects (gamma terms) — random small perturbations
  gamma1.start <- rnorm(1, mean = 0, sd = 0.01)
  gamma2.start <- rnorm(1, mean = 0, sd = 0.01)

  return(list(
    a = a.start,
    d = d.start,
    z0 = z0.start,
    eta_s = eta_s.start,
    gamma1 = gamma1.start,
    gamma2 = gamma2.start,
    t_sigma = t_sigma.start
  ))
}

init_list <- list(
  fun.inits(1),
  fun.inits(2),
  fun.inits(3),
  fun.inits(4)
)
```

```{r}
# fun.model.file <- "model_root.R" 
# mcmc.pars <- list(iter = 50000, burn = 10000, nchain=4, thin=1)
# 
# jags.out <- jags(
#   data = bdat,
#   inits = init_list,
#   parameters.to.save = fun.params,
#   model.file = fun.model.file,
#   n.chains = mcmc.pars$nchain,
#   n.iter = mcmc.pars$iter,
#   n.burnin = mcmc.pars$burn,
#   n.thin = mcmc.pars$thin,
#   DIC = FALSE,
#   progress.bar = "none"
# )
```

```{r}
# load the result instead of running jags everytime
load("./bayes_model_run_final.RData")
```

### Model Evaluation

```{r}
sim <- as.mcmc(jags.out)
```

#### Trace plots

```{r}
gelman_results <- gelman.diag(sim)
psrf_df <- as.data.frame(gelman_results$psrf)
psrf_df$Parameter <- rownames(psrf_df)
colnames(psrf_df) <- c("Point_Est", "Upper_CI", "Parameter")
```

```{r}
psrf_top20 <- psrf_df |>
  arrange(desc(Point_Est)) |>
  slice(1:20) |>
  mutate(Flag = ifelse(Point_Est > 1.01, "Above 1.01", "Below 1.01"))

# Plot
plot19 <-  ggplot(psrf_top20, aes(x = Point_Est, y = reorder(Parameter, Point_Est))) +
  geom_point(size = 1.8, color = "steelblue") +
  geom_errorbarh(aes(xmin = Point_Est, xmax = Upper_CI),
                 height = 0.1, color = "gray50") +
  geom_vline(xintercept = 1, linetype = "dotted", color = "gray40") +
  geom_vline(xintercept = 1.01, linetype = "dashed", color = "orange") +
  geom_vline(xintercept = 1.05, linetype = "dashed", color = "red") +
  coord_cartesian(xlim = c(0.999, 1.06)) +
  scale_color_manual(values = c("Below 1.01" = "steelblue",
                                "Above 1.01" = "firebrick")) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 7),
    panel.grid.major.y = element_blank()
  ) +
  labs(
    y = "Parameters (Top 20 by R-hat)",
    x = expression(hat(R)),
    title = "Gelman–Rubin Diagnostic (R-hat)",
    subtitle = "Dashed lines at 1.01 (modern) and 1.05 indicate potential non-convergence"
  )
```

```{r}
library(data.table)
ess <- effectiveSize(sim)

A_vals     <- ess[grep("^A\\[", names(ess))]
alpha_vals <- ess[grep("^alpha\\[", names(ess))]
B_vals     <- ess[grep("^B\\[", names(ess))]
scal_vals  <- ess[grep("^scal\\[", names(ess))]
gamma_vals <- ess[grep("^gamma", names(ess))]

sigma       <- ess["sigma"]
mu_a        <- ess["mu_a"]
mu_d        <- ess["mu_d"]
mu_eta_s    <- ess["mu_eta_s"]
mu_z0       <- ess["mu_z0"]
prec_a      <- ess["prec_a"]
prec_d      <- ess["prec_d"]
prec_eta_s  <- ess["prec_eta_s"]
prec_z0     <- ess["prec_z0"]

effective_dt <- data.table(
  x = c(1, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 2, 20, 3, 4, 5, 6, 7, 8, 9),
  type = c("WT", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU",
           "MU", "WT", "MU", "WT", "WT", "WT", "WT", "WT", "WT", "WT"),
  A = as.numeric(A_vals),
  alpha = as.numeric(alpha_vals),
  B = as.numeric(B_vals),
  scal = as.numeric(scal_vals)
)

hyper_dt <- data.table(
  parameter = c(
    "gamma1 (Outer effect)",
    "gamma2 (Interaction)",
    "sigma (Residual SD)",
    "mu_a",
    "mu_d",
    "mu_eta_s",
    "mu_z0",
    "prec_a",
    "prec_d",
    "prec_eta_s",
    "prec_z0"
  ),
  ESS = as.numeric(c(
    gamma_vals,
    sigma,
    mu_a, mu_d, mu_eta_s, mu_z0,
    prec_a, prec_d, prec_eta_s, prec_z0
  ))
)


hyper_dt[, category := fifelse(grepl("^gamma", parameter), "Fixed effects (γ)",
                        fifelse(grepl("^mu_", parameter), "Hyper means (μ)",
                        fifelse(grepl("^prec_", parameter), "Precisions (τ)",
                        "Residual variance")))]

setorder(hyper_dt, category)


table6 <- kable(hyper_dt, digits = 1, caption = "Effective Sample Size (ESS) for Hyperparameters")
```

```{r}
library(patchwork)

p1 <- ggplot(data=effective_dt, aes(x=x, y=A, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "ESS",
    title = "A Effective Sample Size",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 8000)) +
  annotate("label", x=7, y=8000, label=as.character(as.integer(max(effective_dt$A))), size=2)

p2 <- effective_dt |>
  ggplot(aes(x=x, y=B, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "ESS",
    title = "B Effective Sample Size",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 8000)) +
  annotate("label", x=7, y=8000, label=as.character(as.integer(max(effective_dt$B))), size=2)

p3 <- effective_dt |>
  ggplot(aes(x=x, y=alpha, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "ESS",
    title = "Alpha Effective Sample Size",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 8000)) +
  annotate("label", x=7, y=8000, label=as.character(as.integer(max(effective_dt$alpha))), size=2)

p4 <- effective_dt |>
  ggplot(aes(x=x, y=scal, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "ESS",
    title = "Scale Effective Sample Size",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 8000)) +
  annotate("label", x=7, y=8000, label=as.character(as.integer(max(effective_dt$scal))), size=2)

plot20 <- (p1 + p2) / (p3 + p4)
```

```{r}
rl <- raftery.diag(sim)
```

```{r}
# Combine the resmatrix data for all chains
df_list <- lapply(rl, function(x) as.data.frame(x$resmatrix))

# Get common parameter names
common_params <- Reduce(intersect, lapply(df_list, rownames))

# Average across chains
df_mean <- Reduce("+", lapply(df_list, function(df) df[common_params, ])) / length(df_list)

# Add parameter names
df_mean$Parameter <- common_params

# Clean column names
colnames(df_mean) <- c("Burn_in_M", "Total_N", "Lower_bound_Nmin", "Dependence_I", "Parameter")

# Drop Lower_bound_Nmin
df_mean <- subset(df_mean, select = -Lower_bound_Nmin)

# Sort and select top 10 by Dependence_I
df_top10 <- df_mean[order(-df_mean$Dependence_I), ][1:10, ]

# Nicely formatted table
table5 <- kable(df_top10, digits = 2, align = "c",
      caption = "Top 10 Parameters by Mean Dependence Factor (Averaged Across Chains)")
```

```{r}

mu_a_rl = mean(c(rl[[1]][2][[1]][63], rl[[2]][2][[1]][63], rl[[3]][2][[1]][63], rl[[4]][2][[1]][63]))
mu_a_df_i = mean(c(rl[[1]][2][[1]][63 + (91 * 3)], rl[[2]][2][[1]][63 + (91 * 3)], rl[[3]][2][[1]][63 + (91 * 3)], rl[[4]][2][[1]][63 + (91 * 3)]))

mu_d_rl = mean(c(rl[[1]][2][[1]][64], rl[[2]][2][[1]][64], rl[[3]][2][[1]][64], rl[[4]][2][[1]][64]))
mu_d_df_i = mean(c(rl[[1]][2][[1]][64 + (91 * 3)], rl[[2]][2][[1]][64 + (91 * 3)], rl[[3]][2][[1]][64 + (91 * 3)], rl[[4]][2][[1]][64 + (91 * 3)]))

mu_eta_s_rl = mean(c(rl[[1]][2][[1]][65], rl[[2]][2][[1]][65], rl[[3]][2][[1]][65], rl[[4]][2][[1]][65]))
mu_eta_s_df_i = mean(c(rl[[1]][2][[1]][65 + (91 * 3)], rl[[2]][2][[1]][65 + (91 * 3)], rl[[3]][2][[1]][65 + (91 * 3)], rl[[4]][2][[1]][65 + (91 * 3)]))

mu_z0_rl = mean(c(rl[[1]][2][[1]][66], rl[[2]][2][[1]][66], rl[[3]][2][[1]][66], rl[[4]][2][[1]][66]))
mu_z0_df_i = mean(c(rl[[1]][2][[1]][66 + (91 * 3)], rl[[2]][2][[1]][66 + (91 * 3)], rl[[3]][2][[1]][66 + (91 * 3)], rl[[4]][2][[1]][66 + (91 * 3)]))

prec_a_rl = mean(c(rl[[1]][2][[1]][67], rl[[2]][2][[1]][67], rl[[3]][2][[1]][67], rl[[4]][2][[1]][67]))
prec_a_df_i = mean(c(rl[[1]][2][[1]][67 + (91 * 3)], rl[[2]][2][[1]][67 + (91 * 3)], rl[[3]][2][[1]][67 + (91 * 3)], rl[[4]][2][[1]][67 + (91 * 3)]))

prec_d_rl = mean(c(rl[[1]][2][[1]][68], rl[[2]][2][[1]][68], rl[[3]][2][[1]][68], rl[[4]][2][[1]][68]))
prec_d_df_i = mean(c(rl[[1]][2][[1]][68 + (91 * 3)], rl[[2]][2][[1]][68 + (91 * 3)], rl[[3]][2][[1]][68 + (91 * 3)], rl[[4]][2][[1]][68 + (91 * 3)]))

prec_eta_s_rl = mean(c(rl[[1]][2][[1]][69], rl[[2]][2][[1]][69], rl[[3]][2][[1]][69], rl[[4]][2][[1]][69]))
prec_eta_s_df_i = mean(c(rl[[1]][2][[1]][69 + (91 * 3)], rl[[2]][2][[1]][69 + (91 * 3)], rl[[3]][2][[1]][69 + (91 * 3)], rl[[4]][2][[1]][69 + (91 * 3)]))

prec_z0_rl = mean(c(rl[[1]][2][[1]][70], rl[[2]][2][[1]][70], rl[[3]][2][[1]][70], rl[[4]][2][[1]][70]))
prec_z0_df_i = mean(c(rl[[1]][2][[1]][70 + (91 * 3)], rl[[2]][2][[1]][70 + (91 * 3)], rl[[3]][2][[1]][70 + (91 * 3)], rl[[4]][2][[1]][70 + (91 * 3)]))

scal_rl = c()
scal_df_i = c()
for(idx in 71:90){
  scal_rl = c(scal_rl, mean(c(rl[[1]][2][[1]][idx], rl[[2]][2][[1]][idx], rl[[3]][2][[1]][idx], rl[[4]][2][[1]][idx])))
  new_idx = idx + (85*3)
  scal_df_i = c(scal_df_i, mean(c(rl[[1]][2][[1]][new_idx], rl[[2]][2][[1]][new_idx], rl[[3]][2][[1]][new_idx], rl[[4]][2][[1]][new_idx])))
}

sigma_rl = mean(c(rl[[1]][2][[1]][91], rl[[2]][2][[1]][91], rl[[3]][2][[1]][91], rl[[4]][2][[1]][91]))
sigma_df_i = mean(c(rl[[1]][2][[1]][(91 * 4)], rl[[2]][2][[1]][(91 * 4)], rl[[3]][2][[1]][(91 * 4)], rl[[4]][2][[1]][(91 * 3)]))

rl_dt <- data.table(
  x = c(1, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 2, 20, 3, 4, 5, 6, 7, 8, 9),
  type = c("WT", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "WT", "MU", "WT", "WT", "WT", "WT", "WT", "WT", "WT"),
  A_rl = A_rl,
  alpha_rl = alpha_rl,
  B_rl = B_rl,
  scal_rl = scal_rl,
  A_df_i = A_df_i,
  alpha_df_i = alpha_df_i,
  B_df_i = B_df_i,
  scal_df_i = scal_df_i
)
```

```{r}
# Plots for Burn-In RL for Mean of Chains
p1 <- rl_dt |>
  ggplot(aes(x=x, y=A_rl, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "A Burn-In",
    title = "A Burn-In\nfor each Root",
    fill = "Root Type"
  ) 

p2 <- rl_dt |>
  ggplot(aes(x=x, y=B_rl, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "B Burn-In",
    title = "B Burn-In\nfor each Root",
    fill = "Root Type"
  ) 

p3 <- rl_dt |>
  ggplot(aes(x=x, y=alpha_rl, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Alpha Burn-In",
    title = "Alpha Burn-In\nfor each Root",
    fill = "Root Type"
  )  

p4 <- rl_dt |>
  ggplot(aes(x=x, y=scal_rl, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Scale Burn-In",
    title = "Scale Burn-In\nfor each Root",
    fill = "Root Type"
  ) 

plot20 <- (p1 + p2) / (p3 + p4)
```

```{r}
# Gamma, Sigma, Tau for Burn-In 
rl_dt2 <- data.table(
  gamma1 = gamma_rl[1],
  gamma2 = gamma_rl[2],
  mu_a = mu_a_rl,
  mu_d = mu_d_rl,
  mu_eta_s = mu_eta_s_rl,
  mu_z0 = mu_z0_rl,
  prec_a = prec_a_rl,
  prec_d = prec_d_rl,
  prec_eta_s = prec_eta_s_rl,
  prec_z0 = prec_z0_rl
)

table3 <- kable(rl_dt2)
```

```{r}
# Plots for Dependence Factor RL for Mean of Chains
p1 <- rl_dt |>
  ggplot(aes(x=x, y=A_df_i, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "A Dependence Factor",
    title = "A Dependence Factor\nfor each Root",
    fill = "Root Type"
  ) 

p2 <- rl_dt |>
  ggplot(aes(x=x, y=B_df_i, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "B Dependence Factor",
    title = "B Dependence Factor\nfor each Root",
    fill = "Root Type"
  ) 

p3 <- rl_dt |>
  ggplot(aes(x=x, y=alpha_df_i, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Alpha Dependence Factor",
    title = "Alpha Dependence Factor\nfor each Root",
    fill = "Root Type"
  )  

p4 <- rl_dt |>
  ggplot(aes(x=x, y=scal_df_i, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Scale Dependence Factor",
    title = "Scale Dependence Factor\nfor each Root",
    fill = "Root Type"
  ) 

plot21 <- (p1 + p2) / (p3 + p4)
```

```{r}
df_i_dt2 <- data.table(
  gamma1 = gamma_df_i[1],
  gamma2 = gamma_df_i[2],
  mu_a = mu_a_df_i,
  mu_d = mu_d_df_i,
  mu_eta_s = mu_eta_s_df_i,
  mu_z0 = mu_z0_df_i,
  prec_a = prec_a_df_i,
  prec_d = prec_d_df_i,
  prec_eta_s = prec_eta_s_df_i,
  prec_z0 = prec_z0_df_i
)

table4 <- kable(df_i_dt2)
```

The Gelman-Rubin Plot (@fig-gelman-rubin) indicates that all of our parameters converged with 95% confidence, as each parameter has an R-hat upper confidence interval boundary less than 1.03, which is less than potential non-convergence boundary of 1.05. The parameters that performed the worst in regards to R-hat value are scattered among mutant type and wild type parameters. The traceplots (@fig-trace-plots) also point towards convergence for each of the hyperparameters in our model. Each traceplot appears to have a “fuzzy caterpillar” shape and seem to oscillate around a value that the model converged on for the duration of the iterations. 

The effective sample sizes for A, B, Alpha, and Scale parameters (@fig-ess) appear to be high across each of the wild and mutant type root seven particularly having large effective sample sizes. It appears that the effective sample sizes for the wild type roots is slightly higher than those for the mutant type roots. The effective sample sizes for each of the hyperparameters are high in value (Table 6), particularly for the gamma1 parameter for outer effect and the gamma2 parameter for the interaction between genotype and side. 

In the Raftery-Lewis diagnostic table (Table 5), the parameters with the top ten burn-in and dependence factors are displayed. The dependence factor is highest for various wild and mutant type A and B parameters, indicating that high autocorrelation in the samples increases the sample size needed to converge. This is correlated with the burn in value, as a high burn means there must be a large number of discarded initial iterations before the chain reached stationarity. 

### Residual Plots
```{r}
# par(mfrow=c(3, 3), las=1)
# autocorr.plot(sim, lag.max=10, auto.layout=FALSE, ask=FALSE)
# dev.off()
```
@fig-resid-bayes

### Posterior versus Prior

```{r}
posterior <- as.data.frame(do.call(rbind, sim))
```

```{r}
# Number of draws for prior
n_samp <- nrow(posterior)

priors <- data.frame(
  gamma1 = rnorm(n_samp, 0, 1/sqrt(4)),
  gamma2 = rnorm(n_samp, 0, 1/sqrt(4)),
  mu_a = rnorm(n_samp, log(L_lower), 1),
  mu_d = rnorm(n_samp, log(L_span), 1),
  mu_z0 = rnorm(n_samp, 0, 1),
  mu_eta_s = rnorm(n_samp, log(0.2), 1/sqrt(2)),
  sigma = abs(rt(n_samp, df = 3)) * 3,
  prec_a = rgamma(n_samp, shape = 1, rate = 1),
  prec_d = rgamma(n_samp, shape = 1, rate = 1),
  prec_z0 = rgamma(n_samp, shape = 1, rate = 1),
  prec_eta_s = rgamma(n_samp, shape = 1, rate = 1)
)
```

```{r}
params_to_plot <- c("gamma1", "gamma2", 
                    "mu_a", "mu_d", "mu_z0","mu_eta_s", 
                    "prec_a", "prec_d", "prec_z0", "prec_eta_s", "sigma")

# Make sure we only use parameters actually present
params_to_plot <- intersect(params_to_plot, names(posterior))

# Convert wide → long format
priors_long <- priors %>%
  select(all_of(params_to_plot)) %>%
  pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") %>%
  mutate(type = "Prior")

posterior_long <- posterior %>%
  select(all_of(params_to_plot)) %>%
  pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") %>%
  mutate(type = "Posterior")

df_all <- bind_rows(priors_long, posterior_long)

# Limit sigma range for better visualization
df_all <- df_all |>
  mutate(value = ifelse(parameter == "sigma" & value > 10, 10, value))
```

```{r}
plot22 <- ggplot(df_all, aes(x = value, fill = type)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~parameter, scales = "free", ncol = 4) +
  theme_bw(base_size = 13) +
  labs(title = "Prior vs Posterior Distributions",
       x = "Parameter value", y = "Density") +
  scale_fill_manual(values = c("gray70", "steelblue"))
```

#### Discourse on fig-post-vs-prior-hyper

# Results

## Frequentist estimates

```{r}
WT_results <- data.frame(
  root = character(),
  A = numeric(),
  B = numeric(),
  xmid = numeric(),
  scal = numeric(),
  beta = numeric(),
  beta_p = numeric(),
  beta_sig = character(),
  stringsAsFactors = FALSE
)

for(name in names(WT_model_output)){
  model <- WT_model_output[[name]]
  
  coef_info <- tidy(model)
  beta_row <- coef_info[coef_info$term == "beta", ]

  WT_results <- rbind(
    WT_results,
    data.frame(
      root = name,
      A = coef(model)["A"],
      B = coef(model)["B"],
      xmid = coef(model)["xmid"],
      scal = coef(model)["scal"],
      beta = coef(model)["beta"],
      beta_p = beta_row$p.value,
      beta_sig = ifelse(beta_row$p.value < 0.05, "*", "ns")
    )
  )
}

table1 <- kable(WT_results, digits = 5, caption = "Wild Type Logistic Model Results")
```

```{r}
MU_results <- data.frame(
  root = character(),
  A = numeric(),
  B = numeric(),
  xmid = numeric(),
  scal = numeric(),
  beta = numeric(),
  beta_p = numeric(),
  beta_sig = character(),
  stringsAsFactors = FALSE
)

for(name in names(MU_model_output)){
  model <- MU_model_output[[name]]
  
  coef_info <- tidy(model)
  beta_row <- coef_info[coef_info$term == "beta", ]

  MU_results <- rbind(
    MU_results,
    data.frame(
      root = name,
      A = coef(model)["A"],
      B = coef(model)["B"],
      xmid = coef(model)["xmid"],
      scal = coef(model)["scal"],
      beta = coef(model)["beta"],
      beta_p = beta_row$p.value,
      beta_sig = ifelse(beta_row$p.value < 0.05, "*", "ns")
    )
  )
}

table2 <- kable(MU_results, digits = 5, caption = "Mutant Logistic Model Results")
```

The frequentist model performed well on the wild type with only two roots, root two and root seven, having insignificant beta values as seen in Table 1. All wild type roots have negative beta values besides wild type root two indicating that the outer cell growth starts earlier than the inner cell growth. The insignificant beta values for wild type roots two and seven indicate that there is no significant measurable difference in the inner and outer cell length for these samples.

On the other hand, the frequentist model on the mutant type had less statistically significant beta estimates for the roots, as seen in Table 2. Mutant type roots 1, 4, 5, 6, 7, and 9 have statistically significant beta estimates that are both positive and negative values, showing that the inner or outer measurements might not necessarily be larger than the other but that there is a difference. The rest of the mutant type roots had insignificant beta values, suggesting that there is not a measurable difference between the inner and outer cell lengths.

Our results ultimately give statistically significant evidence that the outer cell measurements are larger for the wild type and that there is not convincing evidence for a clear one-sided discrepancy for mutant type.

## Bayesian - Posterior summary

```{r}
post_summary <-summary(sim)
ps=post_summary[[1]]
```

```{r}
library(DT)

# Mean
means_WT = c()
means_MU = c()

# A, alpha, B, scal
for(i in 0:3){
  s = i*20
  if(i == 3){
    s = s + 10
  }
  means_WT = c(means_WT, mean(ps[s+1], ps[s+12], ps[s+14], ps[s+15], ps[s+16], ps[s+17], ps[s+18], ps[s+19], ps[s+20]))
  means_MU = c(means_MU, mean(ps[s+2], ps[s+3], ps[s+4], ps[s+5], ps[s+6], ps[s+7], ps[s+8], ps[s+9], ps[s+10], ps[s+11], ps[s+13]))
}

# SD
sd_WT = c()
sd_MU = c()

# A, alpha, B, scal
for(i in 0:3){
  s = i*20 + 93
  if(i == 3){
    s = s + 10
  }
  sd_WT = c(sd_WT, mean(ps[s+1], ps[s+12], ps[s+14], ps[s+15], ps[s+16], ps[s+17], ps[s+18], ps[s+19], ps[s+20]))
  sd_MU = c(sd_MU, mean(ps[s+2], ps[s+3], ps[s+4], ps[s+5], ps[s+6], ps[s+7], ps[s+8], ps[s+9], ps[s+10], ps[s+11], ps[s+13]))
}

ps_df <- data.table(
  type = c("WT", "MU"),
  A_mean = c(means_WT[1], means_MU[1]), 
  A_sd = c(sd_WT[1], sd_MU[1]),
  alpha_mean = c(means_WT[2], means_MU[2]),
  alpha_sd = c(sd_WT[2], sd_MU[2]),
  B_mean = c(means_WT[3], means_MU[3]),
  B_sd = c(sd_WT[3], sd_MU[3]),
  scal_mean = c(means_WT[4], means_MU[4]),
  scal_sd = c(sd_WT[4], sd_MU[4])
)

indices = c(61, 62, 63, 64, 65, 66, 67, 68, 69, 70, 91)
means_alt = c()
sd_alt = c()
naive_se_alt = c()
ts_se_alt = c()

for(idx in indices){
  means_alt = c(means_alt, ps[idx])
  sd_alt = c(sd_alt, ps[91 + idx])
  naive_se_alt = c(naive_se_alt, ps[91*2 + idx])
  ts_se_alt = c(ts_se_alt, ps[91*3 + idx])
}

ps_q = post_summary[[2]]

quant_2_5_alt = c()
quant_25_alt = c()
quant_50_alt = c()
quant_75_alt = c()
quant_97_5_alt = c()

for(idx in indices){
  quant_2_5_alt = c(quant_2_5_alt, ps_q[idx])
  quant_25_alt = c(quant_25_alt, ps_q[91 + idx])
  quant_50_alt = c(quant_50_alt, ps_q[91*2 + idx])
  quant_75_alt = c(quant_75_alt, ps_q[91*3 + idx])
  quant_97_5_alt = c(quant_97_5_alt, ps_q[91*4 + idx])
}

alternate_params_df <- data.table(
  parameter = c("Gamma1", "Gamma2", "Mu_A", "Mu_D", "Mu_Eta_S", "Mu_Z0", "Prec_A", "Prec_S", "Prec_Eta_S", "Prec_Z0", "Sigma"),
  mean = means_alt, 
  sd = sd_alt,
  time_series_se = ts_se_alt,
  quantile_2_5 = quant_2_5_alt,
  quantile_97_5 = quant_97_5_alt
)

table7 <- kable(alternate_params_df)

table8 <- kable(ps_df)
```

## Compare freq vs bayes

```{r}
get_freq_preds <- function(model_list, data, geno_label) {
  map2_dfr(
    .x = model_list,
    .y = names(model_list),
    ~ {
      root_id <- str_replace(.y, "root_", "")
      d <- data |> filter(root == root_id)
      d |>
        mutate(Freq_Pred = fitted(.x),
               Genotype = geno_label)
    }
  )
}

freq_preds <- bind_rows(
  get_freq_preds(WT_model_output, rootData_WT, "WT"),
  get_freq_preds(MU_model_output, rootData_MU, "MU")
)

pm <- jags.out$BUGSoutput$mean

A_vec     <- pm$A
B_vec     <- pm$B
alpha_vec <- pm$alpha
scal_vec  <- pm$scal

root_idx <- as.numeric(as.factor(rootData$root))              # 1..n root
side_num <- ifelse(rootData$side == "Outer", 1, 0)            # 0 = inner, 1 = outer
geno_num <- ifelse(rootData$Genotype == "MU", 1, 0)           # match bdat: MU = 1

A_i     <- A_vec[root_idx]
B_i     <- B_vec[root_idx]
alpha_i <- alpha_vec[root_idx]
scal_i  <- scal_vec[root_idx]
gamma1 <- as.numeric(pm$gamma1)
gamma2 <- as.numeric(pm$gamma2)

xmid_eff <- alpha_i + gamma1 * side_num + gamma2 * geno_num

# rebuild Bayesian prediction: note we use x01, not Midline
Bayes_Pred <- A_i + (B_i - A_i) /
  (1 + exp((xmid_eff - rootData$x01) / scal_i))

bayes_preds <- rootData |>
  mutate(Bayes_Pred = Bayes_Pred) |>
  select(root, Midline, side, Genotype, x01, Bayes_Pred)

# join both prediction sources to original data
plot_data <- rootData |>
  left_join(
    freq_preds |> select(root, Midline, side, Genotype, Freq_Pred),
    by = c("root", "Midline", "side", "Genotype")
  ) |>
  left_join(
    bayes_preds,
    by = c("root", "Midline", "side", "Genotype", "x01")
  )

plot_long <- plot_data |>
  select(root, x01, Length, side, Genotype, Freq_Pred, Bayes_Pred) |>
  pivot_longer(
    cols = c(Freq_Pred, Bayes_Pred),
    names_to = "Model",
    values_to = "Pred"
  ) |>
  mutate(Model = recode(Model,
                        "Freq_Pred" = "Frequentist",
                        "Bayes_Pred" = "Bayesian"),
         Model = factor(Model, levels = c("Frequentist", "Bayesian")))

plot_long_avg <- plot_long |>
  group_by(Genotype, Model, side, x01) |>
  summarise(
    Length_mean = mean(Length, na.rm = TRUE),
    Pred_mean   = mean(Pred, na.rm = TRUE),
    Pred_sd     = sd(Pred, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(Genotype, Model, side, x01)
```

```{r}
p_MU_facet <- plot_long %>%
  filter(Genotype == "MU") %>%
  ggplot(aes(x = x01, color = side)) +
  # show actual data 
  stat_summary(aes(y = Length), fun = mean, geom = "point", alpha = 0.25, size = 0.3) +
  # show model predictions
  stat_summary(aes(y = Pred), fun = mean, geom = "line", size = 0.7) +
  facet_grid(root ~ Model) +   # rows = roots, cols = Frequentist/Bayesian
  labs(
    title = "Predicted Growth Curves by Root – Mutant (MU)",
    x = "Normalized Position (x01)",
    y = "Cell Length (µm)",
    color = "Side"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    strip.text = element_text(size = 8),
    panel.spacing = unit(0.5, "lines")
  )

p_WT_facet <- plot_long %>%
  filter(Genotype == "WT") %>%
  ggplot(aes(x = x01, color = side)) +
  stat_summary(aes(y = Length), fun = mean, geom = "point", alpha = 0.25, size = 0.3) +
  stat_summary(aes(y = Pred), fun = mean, geom = "line", size = 0.7) +
  facet_grid(root ~ Model) +
  labs(
    title = "Predicted Growth Curves by Root – Wild Type (WT)",
    x = "Normalized Position (x01)",
    y = "Cell Length (µm)",
    color = "Side"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    strip.text = element_text(size = 8),
    panel.spacing = unit(0.5, "lines")
  )

p_MU_facet_small <- p_MU_facet +
  scale_x_continuous(breaks = c(0, 0.5, 1)) +
  theme(
    axis.text.x = element_text(size = 6),
    axis.title.x = element_text(size = 8)
  )

p_WT_facet_small <- p_WT_facet +
  scale_x_continuous(breaks = c(0, 0.5, 1)) +
  theme(
    axis.text.x = element_text(size = 6),
    axis.title.x = element_text(size = 8)
  )

# put them next to each other
combined <- p_WT_facet_small + p_MU_facet_small +
  plot_layout(ncol = 2, widths = c(1, 1))

combined
```

```{r}
# Compute Bayesian residuals
Bayes_residuals <- rootData |>
  mutate(
    Bayes_Fitted = Bayes_Pred,
    Bayes_Residual = Length - Bayes_Pred
  ) |>
  select(root, side, Genotype, x01, Midline, Bayes_Fitted, Bayes_Residual)

# Filter by genotype within each plot
plot_bayes_fit_wt <- Bayes_residuals %>%
  filter(Genotype == "WT") %>%
  ggplot(aes(x = Bayes_Fitted, y = Bayes_Residual, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (WT)",
       x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

plot_bayes_mid_wt <- Bayes_residuals %>%
  filter(Genotype == "WT") %>%
  ggplot(aes(x = Midline, y = Bayes_Residual, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (WT)",
       x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()

plot_bayes_fit_mu <- Bayes_residuals %>%
  filter(Genotype == "MU") %>%
  ggplot(aes(x = Bayes_Fitted, y = Bayes_Residual, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (MU)",
       x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

plot_bayes_mid_mu <- Bayes_residuals %>%
  filter(Genotype == "MU") %>%
  ggplot(aes(x = Midline, y = Bayes_Residual, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (MU)",
       x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()
```

```{r}
Bayes_residuals <- Bayes_residuals %>%
  mutate(root = factor(root, 
                       levels = c(paste0("WT", 1:9), paste0("MU", 1:11))))

p_all_box_bayes <- ggplot(Bayes_residuals, aes(x = root, y = Bayes_Residual, fill = side)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(title = "Residuals by Root (Bayesian Model)",
       x = "Root ID", y = "ResidualS", fill = "Side") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )
```

# Limitations and Conclusion

# Appendix

```{r}
#| tbl-cap: "Summary of wild type frequentist results."
table1
```

```{r}
#| tbl-cap: "Summary of mutant type frequentist results."
table2
```

```{r}
#| tbl-cap: "Top 10 Raftery-Lewis Burn-In and Dependence Factor Results."
table5
```

```{r}
#| tbl-cap: "Effective Sample Size Hyperparameters."
table6
```

```{r}
#| tbl-cap: "Posterior summary of parameters"
table7
```

```{r}
#| tbl-cap: "Posterior summary for Wildtype and Mutant Roots"
table8
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-dist
#| fig-cap: "Distribution of cell length for inner and outer cells."
plot1
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-wt
#| fig-cap: "Distribution of cell length for each wild root."
plot2
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-mu
#| fig-cap: "Distribution of cell length for each mutant root."
plot3
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-smooth-wt
#| fig-cap: "Distribution of cell length for each wild root."
plot4
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-smooth-mu
#| fig-cap: "Distribution of cell length for each mutant root."
plot5
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-smooth-wt
#| fig-cap: "Distribution of cell length for wild type."
plot6
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-smooth-mu
#| fig-cap: "Distribution of cell length for mutant type."
plot7
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-mean-cell-length-per-root-wt
#| fig-cap: "Distribution of mean cell length for each wild root."
plot10
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-mean-cell-length-per-root-mu
#| fig-cap: "Distribution of mean cell length for each mutant root."
plot11
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-resid-freq
#| fig-cap: "Residuals Plots for Frequentist Model"

(plot12 + plot13) / (plot15 + plot16) +
  plot_layout(guides = "collect") +       
  theme(legend.position = "right")
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-resid-by-root
#| fig-cap: "Residuals by Root for Frequentist Model."
p_all_box
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-freq-acf
#| fig-cap: "Frequentist Model acf plot."
acf(resid(m0))
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-gelman-rubin
#| fig-cap: "Gelman-Rubin top 20 plot."
plot19
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-ess
#| fig-cap: "Effective Sample Size for A, B, Alpha, and Scal"
plot20
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-trace-plots
#| fig-cap: "Traceplots for hyperparameters."
par(mfrow = c(3, 4), mar = c(4, 4, 2, 1))
for (param in c("sigma", "prec_a", "prec_d", "prec_z0", "prec_eta_s",
                "mu_a", "mu_d", "mu_z0", "mu_eta_s",
                "gamma1", "gamma2")) {
  traceplot(sim[, param], main = param, col = 1:4)
}
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-post-vs-prior-hyper
#| fig-cap: "Posterior vs. Prior Distributions for hyperparameters."
plot22
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-resid-bayes
#| fig-cap: "Residuals Plots for Bayesian Model."

(plot_bayes_fit_wt + plot_bayes_mid_wt) /
(plot_bayes_fit_mu + plot_bayes_mid_mu) +
  plot_layout(guides = "collect") +
  theme(legend.position = "right")
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-resid-by-root-bayes
#| fig-cap: "Residuals by Root for Bayesian Model."

p_all_box_bayes
```
