---
title: "case4_code"
format: html
editor: visual
---

```{r}
load("./rootData.RData")
```

```{r}
library(ggplot2)
library(dplyr)
library(nlme)
library(broom)
library(knitr)
library(stringr)
library(tidyverse)
```

# EDA

```{r}
ggplot(rootData, aes(x = side, y = Length, fill = side)) +
  geom_boxplot() +
  facet_wrap(~ Genotype) +
  labs(title = "Cell Length Distribution by Inner vs Outer",
       y = "Cell Length") +
  theme_bw()
```

```{r}
rootData_WT <- rootData |> filter(Genotype == "WT")

ggplot(rootData_WT, aes(x = Midline, y = Length, 
                        color = side, group = interaction(root, side))) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "Cell Length Along Midline – Wild Type Roots",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
rootData_MU <- rootData |> filter(Genotype == "MU")
desired_order <- c("MU1", "MU2", "MU3", "MU4", "MU5", "MU6", "MU7", "MU8", "MU9", "MU10", "MU11")
rootData_MU <- rootData_MU |>
  mutate(root = factor(root, levels = desired_order))

ggplot(rootData_MU, aes(x = Midline, y = Length, 
                        color = side, group = interaction(root, side))) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "Cell Length Along Midline – Mutant Roots",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
ggplot(rootData_WT,
       aes(x = Midline, y = Length, color = side, group = interaction(root, side))) +
  geom_smooth(method = "loess", se = FALSE, span = 0.75, size = 0.8) +
  geom_point(alpha = 0.4, size = 0.5) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "WT: Inner vs Outer Smoothed Trends for Each Root",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(rootData_MU,
       aes(x = Midline, y = Length, color = side, group = interaction(root, side))) +
  geom_smooth(method = "loess", se = FALSE, span = 0.75, size = 0.8) +
  geom_point(alpha = 0.4, size = 0.5) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "MT: Inner vs Outer Smoothed Trends for Each Root",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
ggplot(rootData_WT, aes(x = Midline, y = Length, color = side)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.8, size = 1.2) +
  labs(
    title = "WT: Inner vs Outer Overall Cell Length Trends",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw()

ggplot(rootData_MU, aes(x = Midline, y = Length, color = side)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.8, size = 1.2) +
  labs(
    title = "MU: Inner vs Outer Overall Cell Length Trends",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw()
```

Logistic

```{r warning=FALSE}
logis_smooth <- function(df) {
  ggplot(
    df,
    aes(x = Midline, y = Length, color = side, group = interaction(root, side))
  ) +
    geom_point(alpha = 0.4, size = 0.5) +
    geom_smooth(
      method = "nls",
      formula = y ~ SSlogis(x, Asym, xmid, scal),
      method.args = list(control = nls.control(maxiter = 200, warnOnly = TRUE)),
      se = FALSE,
      size = 0.8
    ) +
    facet_wrap(~ root, scales = "free_y") +
    labs(
      x = "Distance Along Midline",
      y = "Cell Length",
      color = "Side"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Wild type
logis_smooth(rootData_WT) + ggtitle("WT: Inner vs Outer Logistic Fits for Each Root")

# Mutant
logis_smooth(rootData_MU) + ggtitle("MU: Inner vs Outer Logistic Fits for Each Root")
```

```{r}
# Mean Length Difference 

# MU
diffDf <- rootData_MU |>
  mutate(diff_group = as.character(as.integer(Midline / 100) * 100)) |>
  group_by(diff_group, side) |>
  mutate(mean_length = mean(Length))

diffDf |>
  ggplot(aes(x=as.integer(diff_group), y = mean_length, color=side)) +
  geom_point() +
  labs(
    x = "Midline (Grouped)",
    y = "Mean Length",
    title = "Mean Length in Groups of 100 for Midline (Mutant)"
  )

# WT 
diffDf <- rootData_WT |>
  mutate(diff_group = as.character(as.integer(Midline / 100) * 100)) |>
  group_by(diff_group, side) |>
  mutate(mean_length = mean(Length))

diffDf |>
  ggplot(aes(x=as.integer(diff_group), y = mean_length, color=side)) +
  geom_point() +
  labs(
    x = "Midline (Grouped)",
    y = "Mean Length",
    title = "Mean Length in Groups of 100 for Midline (Wildtype)"
  )
```

```{r}
# Plot Average Length Per Genotype
rootData_WT |>
  ggplot(aes(x=root, y=Length, fill=side)) +
  geom_bar(stat="summary", fun="mean", position="dodge")

rootData_MU |>
  ggplot(aes(x=root, y=Length, fill=side)) +
  geom_bar(stat="summary", fun="mean", position="dodge")
```

# Frequentist Model

### Fit nls with customized function

$$
\text{Length ~} A+\frac{(B-A)}{1+e^{((x_{mid} + \beta * \text{side_ind} - x)/\text{scal})}}
$$

```{r}
fpl_customized <- function(x, A, B, xmid, scal, beta, side_ind){
  xmid_eff <- xmid + beta * side_ind
  A + (B - A) / (1 + exp((xmid_eff - x) / scal))
}
```

```{r}
WT_model_output <- list()

for (root_num in unique(rootData_WT$root)) {
  curr_root <- rootData_WT |> filter(root == root_num)
  curr_root$side_ind <- ifelse(curr_root$side == "Outer", 1, 0)

  y_min <- min(curr_root$Length)
  y_max <- max(curr_root$Length)

  start_vals <- list(
    A = y_min * 0.9,
    B = y_max * 1.1,
    xmid = median(curr_root$Midline),
    scal = diff(range(curr_root$Midline)) / 4,
    beta = -50
  )

  lower_bounds <- c(A = y_min * 0.5,
                    B = y_min,
                    xmid = min(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)) / 10,
                    beta = -500)

  upper_bounds <- c(A = y_max,
                    B = y_max * 2.0,
                    xmid = max(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)),
                    beta = 500)

  fit <- try(
    nls(
      Length ~ fpl_customized(Midline, A, B, xmid, scal, beta, side_ind),
      data = curr_root,
      start = start_vals,
      algorithm = "port",
      lower = lower_bounds,
      upper = upper_bounds,
      control = nls.control(maxiter = 1000, warnOnly = TRUE, minFactor = 1e-10)
    ),
    silent = TRUE
  )

  if (!inherits(fit, "try-error")) {
    WT_model_output[[paste0("root_", root_num)]] <- fit
  } else {
    message(paste("Model failed for root", root_num))
  }
}
```

```{r}
WT_results <- data.frame(
  root = character(),
  A = numeric(),
  B = numeric(),
  xmid = numeric(),
  scal = numeric(),
  beta = numeric(),
  beta_p = numeric(),
  beta_sig = character(),
  stringsAsFactors = FALSE
)

for(name in names(WT_model_output)){
  model <- WT_model_output[[name]]
  
  coef_info <- tidy(model)
  beta_row <- coef_info[coef_info$term == "beta", ]

  WT_results <- rbind(
    WT_results,
    data.frame(
      root = name,
      A = coef(model)["A"],
      B = coef(model)["B"],
      xmid = coef(model)["xmid"],
      scal = coef(model)["scal"],
      beta = coef(model)["beta"],
      beta_p = beta_row$p.value,
      beta_sig = ifelse(beta_row$p.value < 0.05, "*", "ns")
    )
  )
}

kable(WT_results, digits = 5, caption = "Wild Type Logistic Model Results")
```

```{r}
MU_model_output <- list()

for (root_num in unique(rootData_MU$root)) {
  curr_root <- rootData_MU |> filter(root == root_num)
  curr_root$side_ind <- ifelse(curr_root$side == "Outer", 1, 0)

  y_min <- min(curr_root$Length)
  y_max <- max(curr_root$Length)

  start_vals <- list(
    A = y_min * 0.9,
    B = y_max * 1.1,
    xmid = median(curr_root$Midline),
    scal = diff(range(curr_root$Midline)) / 4,
    beta = 0
  )

  lower_bounds <- c(A = y_min * 0.5,
                    B = y_min,
                    xmid = min(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)) / 10,
                    beta = -500)

  upper_bounds <- c(A = y_max,
                    B = y_max * 2.0,
                    xmid = max(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)),
                    beta = 500)

  fit <- try(
    nls(
      Length ~ fpl_customized(Midline, A, B, xmid, scal, beta, side_ind),
      data = curr_root,
      start = start_vals,
      algorithm = "port",
      lower = lower_bounds,
      upper = upper_bounds,
      control = nls.control(maxiter = 1000, warnOnly = TRUE, minFactor = 1e-10)
    ),
    silent = TRUE
  )

  if (!inherits(fit, "try-error")) {
    MU_model_output[[paste0("root_", root_num)]] <- fit
  } else {
    message(paste("Model failed for root", root_num))
  }
}
```

```{r}
MU_results <- data.frame(
  root = character(),
  A = numeric(),
  B = numeric(),
  xmid = numeric(),
  scal = numeric(),
  beta = numeric(),
  beta_p = numeric(),
  beta_sig = character(),
  stringsAsFactors = FALSE
)

for(name in names(MU_model_output)){
  model <- MU_model_output[[name]]
  
  coef_info <- tidy(model)
  beta_row <- coef_info[coef_info$term == "beta", ]

  MU_results <- rbind(
    MU_results,
    data.frame(
      root = name,
      A = coef(model)["A"],
      B = coef(model)["B"],
      xmid = coef(model)["xmid"],
      scal = coef(model)["scal"],
      beta = coef(model)["beta"],
      beta_p = beta_row$p.value,
      beta_sig = ifelse(beta_row$p.value < 0.05, "*", "ns")
    )
  )
}

kable(MU_results, digits = 5, caption = "Mutant Logistic Model Results")
```

### Evaluation

```{r}
WT_residuals <- data.frame()

for (name in names(WT_model_output)) {
  model <- WT_model_output[[name]]
  root_label <- str_replace(name, "root_", "")
    curr_data <- rootData_WT |> filter(root == root_label)
  
  curr_data <- curr_data |>
    mutate(
      Fitted = fitted(model),
      Residuals = resid(model)
    )
  
  WT_residuals <- bind_rows(WT_residuals, curr_data)
}
```

```{r}
ggplot(WT_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (WT)", x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

ggplot(WT_residuals, aes(x = Midline, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (WT)", x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()

ggplot(WT_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ root, scales = "free_x") +
  labs(title = "Residuals vs Fitted by Root (WT)", x = "Fitted Length", y = "Residuals") +
  theme_minimal()
```

```{r}
MU_residuals <- data.frame()

for (name in names(MU_model_output)) {
  model <- MU_model_output[[name]]
  root_label <- str_replace(name, "root_", "")
    curr_data <- rootData_MU |> filter(root == root_label)
  
  curr_data <- curr_data |>
    mutate(
      Fitted = fitted(model),
      Residuals = resid(model)
    )
  
  MU_residuals <- bind_rows(MU_residuals, curr_data)
}
```

```{r}
ggplot(MU_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (MU)", x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

ggplot(MU_residuals, aes(x = Midline, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (MU)", x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()

ggplot(MU_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ root, scales = "free_x") +
  labs(title = "Residuals vs Fitted by Root (MU)", x = "Fitted Length", y = "Residuals") +
  theme_minimal()
```

```{r}
combined_residuals <- bind_rows(
  WT_residuals %>% mutate(Genotype = "WT"),
  MU_residuals %>% mutate(Genotype = "MU")
)

ggplot(combined_residuals, aes(x = Fitted, y = Residuals, color = Genotype)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Residuals vs Fitted by Genotype",
    x = "Fitted Cell Length",
    y = "Residuals"
  ) +
  theme_minimal()
```

```{r}
# ACF plot (autocorr check)
m0 <- lm(Length ~ side * poly(Midline, 2), data = rootData)
acf(resid(m0))  # if this shows notable autocorr, ARMA in the next step is justified
```

# Bayesian model

```{r}
library(rjags)
library(R2jags)
library(coda)
```

```{r}
# Normalize Midline to the [0,1] range
rootData <- rootData |>
  group_by(root) |>
  mutate(x01 = (Midline - min(Midline)) / (max(Midline) - min(Midline))) |>
  ungroup()
```

```{r}
# bdat <- list(
#   Length = rootData$Length,
#   N = nrow(rootData),
#   x01 = rootData$x01,
#   side = as.numeric(rootData$side == "Outer"),
#   geno = as.numeric(rootData$Genotype == "WT"),
#   root = as.numeric(as.factor(rootData$root)),
#   n.root = length(unique(rootData$root))
# )
```

```{r}
# fun.params <- c("A", "B", "scal", "alpha", 
#                 "gamma1", "gamma2", "gamma3",
#                 "sigma", "tau")
```

```{r}
# fun.model.file <- "model_root.R"
# mcmc.pars <- list(iter = 50000, burn = 10000, thin = 10)
# 
# fun.inits <- function() {
#   list(
#     A = rep(5, bdat$n.root),
#     B = rep(30, bdat$n.root),
#     scal = rep(0.2, bdat$n.root),
#     alpha = rep(0, bdat$n.root),
#     gamma1 = 0,
#     gamma2 = 0,
#     gamma3 = 0,
#     sigma = 5,
#     tau = 2
#   )
# }
# 
# jags.out <- jags(
#   data = bdat,
#   inits = fun.inits,
#   parameters.to.save = fun.params,
#   model.file = fun.model.file,
#   n.chains = 1,
#   n.iter = mcmc.pars$iter,
#   n.burnin = mcmc.pars$burn,
#   n.thin = mcmc.pars$thin,
#   DIC = FALSE,
#   progress.bar = "none"
# )
```

# MCMC version 2

-   chain 1-\> 4
-   A,B, scal, alpha starting values from frequentist

```{r}
bdat <- list(
  Length = rootData$Length,
  N = nrow(rootData),
  x01 = rootData$x01,
  side = as.numeric(rootData$side == "Outer"),
  geno = as.numeric(rootData$Genotype == "WT"),
  root = as.numeric(as.factor(rootData$root)),
  n.root = length(unique(rootData$root))
)

fun.params <- c("A", "B", "scal", "alpha", 
                "gamma1", "gamma2", "gamma3",
                "sigma", "tau")
```

```{r}
get_nls_summary <- function(model) {
  s <- summary(model)
  est <- s$coefficients[, 1]
  se  <- s$coefficients[, 2]
  tibble(A = est["A"], A_se = se["A"],
         B = est["B"], B_se = se["B"],
         scal = est["scal"], scal_se = se["scal"],
         xmid = est["xmid"], xmid_se = se["xmid"])
}

WT_results_full <- map_dfr(WT_model_output, get_nls_summary, .id = "root")
MU_results_full <- map_dfr(MU_model_output, get_nls_summary, .id = "root")
all_results <- bind_rows(WT_results_full, MU_results_full)
```

```{r}
root_levels <- levels(as.factor(rootData$root))
nroot <- length(root_levels)
A_init <- B_init <- scal_init <- alpha_init <- numeric(nroot)
A_sd <- B_sd <- scal_sd <- alpha_sd <- numeric(nroot)

for (i in seq_along(root_levels)) {
  this_name <- paste0("root_", root_levels[i])
  row_i <- all_results[all_results$root == this_name, ]
  rd_i <- filter(rootData, root == root_levels[i])

  if (nrow(row_i) == 1) {
    A_init[i] <- row_i$A
    A_sd[i]   <- row_i$A_se
    B_init[i] <- row_i$B
    B_sd[i]   <- row_i$B_se

    # Get min and max for this root
    mid_min <- min(rd_i$Midline)
    mid_max <- max(rd_i$Midline)
    span    <- mid_max - mid_min

    # Normalize both xmid and scal
    alpha_init[i] <- (row_i$xmid - mid_min) / span
    alpha_sd[i]   <- row_i$xmid_se / span
    scal_init[i]  <- row_i$scal / span
    scal_sd[i]    <- row_i$scal_se / span
  } else {
    A_init[i] <- 5;  A_sd[i] <- 1
    B_init[i] <- 30; B_sd[i] <- 2
    scal_init[i] <- 0.2; scal_sd[i] <- 0.05
    alpha_init[i] <- 0.5; alpha_sd[i] <- 0.05
  }
}
```

```{r}
fun.inits <- function(chain_id = 1) {
  sigma.start <- runif(1, 2, 5)
  tau.start   <- runif(1, 0, 3)
  
  # “Overdisperse” A, B, scal, alpha around empirical estimates
  A.start = pmax(0, rnorm(bdat$n.root, mean = A_init, sd = 5 * A_sd))
  B.start = pmax(0,rnorm(bdat$n.root, mean = B_init, sd = 5 * B_sd))
  scal.start = pmax(1e-3, rnorm(bdat$n.root, mean = scal_init, sd = 5 * scal_sd))
  alpha.start = rnorm(bdat$n.root, mean = alpha_init, sd = 5 * alpha_sd)

  # Fixed effects (gamma terms) — random small perturbations
  gamma1.start <- rnorm(1, mean = 0, sd = 0.01)
  gamma2.start <- rnorm(1, mean = 0, sd = 0.01)
  gamma3.start <- rnorm(1, mean = 0, sd = 0.01)
  
  # Return inits list (matching model parameters)
  return(list(
    A = A.start,
    B = B.start,
    scal = scal.start,
    alpha = alpha.start,
    gamma1 = gamma1.start,
    gamma2 = gamma2.start,
    gamma3 = gamma3.start,
    sigma = sigma.start,
    tau = tau.start
  ))
}

init_list <- list(
  fun.inits(1),
  fun.inits(2),
  fun.inits(3),
  fun.inits(4)
)
```

```{r}
muA_hat      <- mean(A_init,      na.rm = TRUE)
sdA_hat      <- sd(A_init,        na.rm = TRUE)
muB_hat      <- mean(B_init,      na.rm = TRUE)
sdB_hat      <- sd(B_init,        na.rm = TRUE)
muAlpha_hat  <- mean(alpha_init,  na.rm = TRUE)
sdAlpha_hat  <- sd(alpha_init,    na.rm = TRUE)
muScal_hat   <- mean(scal_init,   na.rm = TRUE)
sdScal_hat   <- sd(scal_init,     na.rm = TRUE)

bdat <- c(
  bdat,
  list(
    muA_hat = muA_hat,
    prior_sd_A = 3*sdA_hat,
    muB_hat = muB_hat,
    prior_sd_B = 3*sdB_hat,
    muAlpha_hat = muAlpha_hat,
    prior_sd_alpha = 3*sdAlpha_hat,
    muScal_hat = muScal_hat,
    prior_sd_scal = 3*sdScal_hat
  )
)
```

```{r}
fun.model.file <- "model_root.R" 
mcmc.pars <- list(iter = 50000, burn = 10000, nchain=4, thin=1)

jags.out <- jags(
  data = bdat,
  inits = init_list,
  parameters.to.save = fun.params,
  model.file = fun.model.file,
  n.chains = mcmc.pars$nchain,
  n.iter = mcmc.pars$iter,
  n.burnin = mcmc.pars$burn,
  n.thin = mcmc.pars$thin,
  DIC = FALSE,
  progress.bar = "none"
)
```

```{r}
jags.out$BUGSoutput$summary[c("gamma1", "gamma2", "gamma3"), ]
```

```{r}
S <- as.mcmc(jags.out)
traceplot(S[, c("sigma", "tau")])
densplot(S[, c("sigma", "tau")])
```

### Gelman-Rubin Diagnostics

Provides an upper bound on variation - this provides confidence that the sampler is not still moving towards the posterior mode (can provide evidence that the poster is multi-modal and highlight an inefficient (slow mixing) sampler)

```{r}
# Gelman Plot
sim <- as.mcmc(jags.out)
pdf("grPlots.pdf", width=10, height=8)
par(mfrow=c(3,2), las=1)
gelman.plot(sim, las=1, transform=TRUE, ask=FALSE)
dev.off()
```

```{r}
gelman.diag(sim, transform = TRUE)
```

### Effective Size Results

Answers: How many independent samples are roughly equivalent in information content to this correlated sample? Variation value at zero is used to estimate effective sample size

```{r}
ess <- effectiveSize(sim)
```

```{r}
ess[1:9]
```

```{r}
ess[10:18]
```

```{r}
ess[19:27]
```

```{r}
ess[28:36]
```

```{r}
ess[37:45]
```

```{r}
ess[46:55]
```

```{r}
ess[56:63]
```

```{r}
ess[64:73]
```

```{r}
ess[73:81]
```

```{r}
ess[82:85]
```

### Raftery-Lewis Diagnostic

Number of iterations required to estimate a specified quantity q of the marginal posterior distribution of each parameter within a given accuracy (+- r) with specified probability (s)

```{r}
rl <- raftery.diag(sim)
```

```{r}
# Raftery-Lewis Diagnostic 
rl[[1]]
```

### ACF Plots (Autocorrelation)

```{r}
pdf("acfPlots.pdf", width=10, height=8)
par(mfrow=c(3, 3), las=1)
autocorr.plot(sim, lag.max=10, auto.layout=FALSE, ask=FALSE)
dev.off()
```

```{r}
autocorr.diag(sim)
```

### CODA Output Analysis Functions

```{r}
comboHC <- as.mcmc(jags.out$BUGSoutput$sims.matrix)
summary(comboHC)
```
