---
title: "Sta 440 Case 4"
author: "Olivia Fu, Yura Heo, Jack Steel"
date: "November 2025"
format: 
  pdf:
    documentclass: article
    fontsize: 11pt
geometry: margin=1in
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
load("./rootData.RData")

library(ggplot2)
library(dplyr)
library(nlme)
library(broom)
library(knitr)
library(stringr)
library(tidyverse)
```


# Background

Root circumnutation, the helical or oscillatory growth movement of roots, has been hypothesized to result from differential cell elongation on opposing sides of the root tip. This growth pattern enables roots to navigate the soil, penetrate obstacles, and optimize anchorage and nutrient uptake. To explore this mechanism, we compare spatial growth patterns between wild-type and mutant genotypes that differ in their circumnutation behavior. Specifically, we model root elongation along the midline across inner and outer sides to test whether growth asymmetry drives circumnutation and whether this asymmetry is reduced in mutants.   

This analysis aims to quantify the extent of asymmetry in cell length along the inner and outer curves of individual roots and to assess how these patterns differ between genotypes. By fitting both frequentist and hierarchical Bayesian nonlinear regression models, we seek to:  
(1) identify within-root growth differences linked to circumnutation  
(2) characterize variability across roots, and 
(3) evaluate whether the mutant genotype exhibits reduced differential growth consistent with its impaired movement phenotype.    

# Exploratory Data Analysis


```{r}
plot1 <- ggplot(rootData, aes(x = side, y = Length, fill = side)) +
  geom_boxplot() +
  facet_wrap(~ Genotype) +
  labs(title = "Cell Length Distribution by Inner vs Outer",
       y = "Cell Length") +
  theme_bw()
```

The roots are classified as mutant or wild type, and their cell lengths on the inner side and outer side of the roots are measured at various distances along the midline. The distribution for the cell length is right skewed for both sides and both root types, with the mutant type being slightly stronger right skewed (@fig-cell-length-dist). The inner and outer side for the mutant type have a nearly identical distribution, with a median cell length around 9. On the other hand, the distributions for the cell length for the wild type differ between the inner measurements and outer measurements, where the outer cell length measurements are slightly larger than the inner measurements. 

```{r}
rootData_WT <- rootData |> filter(Genotype == "WT")

plot2 <- ggplot(rootData_WT, aes(x = Midline, y = Length, 
                        color = side, group = interaction(root, side))) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "Cell Length Along Midline – Wild Type Roots",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

rootData_MU <- rootData |> filter(Genotype == "MU")
desired_order <- c("MU1", "MU2", "MU3", "MU4", "MU5", "MU6", "MU7", "MU8", "MU9", "MU10", "MU11")
rootData_MU <- rootData_MU |>
  mutate(root = factor(root, levels = desired_order))

plot3 <- ggplot(rootData_MU, aes(x = Midline, y = Length, 
                        color = side, group = interaction(root, side))) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "Cell Length Along Midline – Mutant Roots",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The trends for the wild type roots in @fig-cell-length-per-root-wt and @fig-cell-length-per-root-smooth-wt show a large gap between the lines for the inner and outer measurements, while the trends for the mutant type roots in @fig-cell-length-per-root-mu and @fig-cell-length-per-root-smooth-mu show a small gap between the lines for the inner and outer measurements. Wild type roots one, three, and eight particularly have higher inner measurements than outer measurements. On the other hand, mutant type roots have intersecting lines for inner and outer measurements as distance along the midline increases. This is further emphasized by the smooth line plots in @fig-cell-length-smooth-wt and @fig-cell-length-smooth-mu - the wild type has noticeably different inner and outer measurements for cell length. 


```{r}
plot4 <- ggplot(rootData_WT,
       aes(x = Midline, y = Length, color = side, group = interaction(root, side))) +
  geom_smooth(method = "loess", se = FALSE, span = 0.75, size = 0.8) +
  geom_point(alpha = 0.4, size = 0.5) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "WT: Inner vs Outer Smoothed Trends for Each Root",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot5 <- ggplot(rootData_MU,
       aes(x = Midline, y = Length, color = side, group = interaction(root, side))) +
  geom_smooth(method = "loess", se = FALSE, span = 0.75, size = 0.8) +
  geom_point(alpha = 0.4, size = 0.5) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "MT: Inner vs Outer Smoothed Trends for Each Root",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
plot6 <- ggplot(rootData_WT, aes(x = Midline, y = Length, color = side)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.8, size = 1.2) +
  labs(
    title = "WT: Inner vs Outer Overall Cell Length Trends",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw()

plot7 <- ggplot(rootData_MU, aes(x = Midline, y = Length, color = side)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.8, size = 1.2) +
  labs(
    title = "MU: Inner vs Outer Overall Cell Length Trends",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw()
```


```{r}
logis_smooth <- function(df) {
  ggplot(
    df,
    aes(x = Midline, y = Length, color = side, group = interaction(root, side))
  ) +
    geom_point(alpha = 0.4, size = 0.5) +
    geom_smooth(
      method = "nls",
      formula = y ~ SSlogis(x, Asym, xmid, scal),
      method.args = list(control = nls.control(maxiter = 200, warnOnly = TRUE)),
      se = FALSE,
      size = 0.8
    ) +
    facet_wrap(~ root, scales = "free_y") +
    labs(
      x = "Distance Along Midline",
      y = "Cell Length",
      color = "Side"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Wild type
plot8 <- logis_smooth(rootData_WT) + ggtitle("WT: Inner vs Outer Logistic Fits for Each Root")

# Mutant
plot9 <- logis_smooth(rootData_MU) + ggtitle("MU: Inner vs Outer Logistic Fits for Each Root")
```

```{r}
plot10 <- rootData_WT |>
  ggplot(aes(x=root, y=Length, fill=side)) +
  geom_bar(stat="summary", fun="mean", position="dodge") +
  labs(
    x = "Root", 
    y = "Mean Cell Length",
    title = "Mean Cell Length for Each Wild Type Root"
  )

plot11 <- rootData_MU |>
  ggplot(aes(x=root, y=Length, fill=side)) +
  geom_bar(stat="summary", fun="mean", position="dodge") +
  labs(
    x = "Root", 
    y = "Mean Cell Length",
    title = "Mean Cell Length for Each Mutant Type Root"
  )
```

The mean cell length across each root for mutant type has similar measurements for the inside and the outside cells, besides for mutant root nine, where the mean inner measurement seems to be larger than the mean outer cell length measurement (@fig-mean-cell-length-per-root-mu). On the other hand, the mean outer cell length is larger for every wild type root besides root seven, where the mean inner measurement is slightly larger than the mean outer measurement (@fig-mean-cell-length-per-root-wt). 

# Model Rationale and Implementation 

## Frequentist Model 

### Model Rationale 
Our EDA showed that cell length increases nonlinearly from the root tip toward the base, consistent with a sigmoidal growth pattern. These findings motivated a restricted logistic model structure that captures both the saturating nature of growth and the side-specific shift in elongation patterns through the inclusion of a side indicator term ($\beta$) to quantify asymmetry in elongation between the inner and outer sides. Because the data were collected separately for each side of the root and for two genotypes (wild type and mutant), this formulation directly reflects the experimental design and allows side- and genotype-level comparisons. The logistic form also aligns with biological constraints: cell lengths cannot be negative and are expected to plateau as elongation slows near the root base.
$$
\text{Length ~} A+\frac{(B-A)}{1+e^{((x_{mid} + \beta * \text{side_ind} - x)/\text{scal})}}
$$
Here, A and B are lower and upper asymptotes of cell length, $x_{mid}$ is the inflection point, $scal$ controls curve steepness, and $\beta$ shifts the midpoint between inner (0) and outer (1) sides, indicating differential growth.   


### Model Implementation

Nonlinear least squares (NLS) models were fit separately for each root with a custom logistic function. Parameter bounds were set to ensure biologically realistic ranges, and starting values were derived from each root's observed cell length distribution. A total of 9 wild-type and 11 mutant roots were models independently to assess within-root and between-genotype differences in elongation patterns.   

```{r}
fpl_customized <- function(x, A, B, xmid, scal, beta, side_ind){
  xmid_eff <- xmid + beta * side_ind
  A + (B - A) / (1 + exp((xmid_eff - x) / scal))
}
```

```{r}
fpl_customized <- function(x, A, B, xmid, scal, beta, side_ind){
  xmid_eff <- xmid + beta * side_ind
  A + (B - A) / (1 + exp((xmid_eff - x) / scal))
}

WT_model_output <- list()

for (root_num in unique(rootData_WT$root)) {
  curr_root <- rootData_WT |> filter(root == root_num)
  curr_root$side_ind <- ifelse(curr_root$side == "Outer", 1, 0)

  y_min <- min(curr_root$Length)
  y_max <- max(curr_root$Length)

  start_vals <- list(
    A = y_min * 0.9,
    B = y_max * 1.1,
    xmid = median(curr_root$Midline),
    scal = diff(range(curr_root$Midline)) / 4,
    beta = -50
  )

  lower_bounds <- c(A = y_min * 0.5,
                    B = y_min,
                    xmid = min(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)) / 10,
                    beta = -500)

  upper_bounds <- c(A = y_max,
                    B = y_max * 2.0,
                    xmid = max(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)),
                    beta = 500)

  fit <- try(
    nls(
      Length ~ fpl_customized(Midline, A, B, xmid, scal, beta, side_ind),
      data = curr_root,
      start = start_vals,
      algorithm = "port",
      lower = lower_bounds,
      upper = upper_bounds,
      control = nls.control(maxiter = 1000, warnOnly = TRUE, minFactor = 1e-10)
    ),
    silent = TRUE
  )

  if (!inherits(fit, "try-error")) {
    WT_model_output[[paste0("root_", root_num)]] <- fit
  } else {
    message(paste("Model failed for root", root_num))
  }
}
```

```{r}
MU_model_output <- list()

for (root_num in unique(rootData_MU$root)) {
  curr_root <- rootData_MU |> filter(root == root_num)
  curr_root$side_ind <- ifelse(curr_root$side == "Outer", 1, 0)

  y_min <- min(curr_root$Length)
  y_max <- max(curr_root$Length)

  start_vals <- list(
    A = y_min * 0.9,
    B = y_max * 1.1,
    xmid = median(curr_root$Midline),
    scal = diff(range(curr_root$Midline)) / 4,
    beta = 0
  )

  lower_bounds <- c(A = y_min * 0.5,
                    B = y_min,
                    xmid = min(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)) / 10,
                    beta = -500)

  upper_bounds <- c(A = y_max,
                    B = y_max * 2.0,
                    xmid = max(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)),
                    beta = 500)

  fit <- try(
    nls(
      Length ~ fpl_customized(Midline, A, B, xmid, scal, beta, side_ind),
      data = curr_root,
      start = start_vals,
      algorithm = "port",
      lower = lower_bounds,
      upper = upper_bounds,
      control = nls.control(maxiter = 1000, warnOnly = TRUE, minFactor = 1e-10)
    ),
    silent = TRUE
  )

  if (!inherits(fit, "try-error")) {
    MU_model_output[[paste0("root_", root_num)]] <- fit
  } else {
    message(paste("Model failed for root", root_num))
  }
}
```

### Model Evaluation
```{r}
WT_results <- data.frame(
  root = character(),
  A = numeric(),
  B = numeric(),
  xmid = numeric(),
  scal = numeric(),
  beta = numeric(),
  beta_p = numeric(),
  beta_sig = character(),
  stringsAsFactors = FALSE
)

for(name in names(WT_model_output)){
  model <- WT_model_output[[name]]
  
  coef_info <- tidy(model)
  beta_row <- coef_info[coef_info$term == "beta", ]

  WT_results <- rbind(
    WT_results,
    data.frame(
      root = name,
      A = coef(model)["A"],
      B = coef(model)["B"],
      xmid = coef(model)["xmid"],
      scal = coef(model)["scal"],
      beta = coef(model)["beta"],
      beta_p = beta_row$p.value,
      beta_sig = ifelse(beta_row$p.value < 0.05, "*", "ns")
    )
  )
}

table1 <- kable(WT_results, digits = 5, caption = "Wild Type Logistic Model Results")
```

```{r}
WT_residuals <- data.frame()

for (name in names(WT_model_output)) {
  model <- WT_model_output[[name]]
  root_label <- str_replace(name, "root_", "")
    curr_data <- rootData_WT |> filter(root == root_label)
  
  curr_data <- curr_data |>
    mutate(
      Fitted = fitted(model),
      Residuals = resid(model)
    )
  
  WT_residuals <- bind_rows(WT_residuals, curr_data)
}

plot12 <- ggplot(WT_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (WT)", x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

plot13 <- ggplot(WT_residuals, aes(x = Midline, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (WT)", x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()

plot14 <- ggplot(WT_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ root, scales = "free_x") +
  labs(title = "Residuals vs Fitted by Root (WT)", x = "Fitted Length", y = "Residuals") +
  theme_minimal()
```

```{r}
MU_results <- data.frame(
  root = character(),
  A = numeric(),
  B = numeric(),
  xmid = numeric(),
  scal = numeric(),
  beta = numeric(),
  beta_p = numeric(),
  beta_sig = character(),
  stringsAsFactors = FALSE
)

for(name in names(MU_model_output)){
  model <- MU_model_output[[name]]
  
  coef_info <- tidy(model)
  beta_row <- coef_info[coef_info$term == "beta", ]

  MU_results <- rbind(
    MU_results,
    data.frame(
      root = name,
      A = coef(model)["A"],
      B = coef(model)["B"],
      xmid = coef(model)["xmid"],
      scal = coef(model)["scal"],
      beta = coef(model)["beta"],
      beta_p = beta_row$p.value,
      beta_sig = ifelse(beta_row$p.value < 0.05, "*", "ns")
    )
  )
}

table2 <- kable(MU_results, digits = 5, caption = "Mutant Logistic Model Results")
```

```{r}
MU_residuals <- data.frame()

for (name in names(MU_model_output)) {
  model <- MU_model_output[[name]]
  root_label <- str_replace(name, "root_", "")
    curr_data <- rootData_MU |> filter(root == root_label)
  
  curr_data <- curr_data |>
    mutate(
      Fitted = fitted(model),
      Residuals = resid(model)
    )
  
  MU_residuals <- bind_rows(MU_residuals, curr_data)
}

plot15 <- ggplot(MU_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (MU)", x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

plot16 <- ggplot(MU_residuals, aes(x = Midline, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (MU)", x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()

plot17 <- ggplot(MU_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ root, scales = "free_x") +
  labs(title = "Residuals vs Fitted by Root (MU)", x = "Fitted Length", y = "Residuals") +
  theme_minimal()
```

```{r}
# ACF plot (autocorr check)
m0 <- lm(Length ~ side * poly(Midline, 2), data = rootData)
```

The frequentist model performed well on the wild type with only two roots, root two and root seven, having insignificant beta values as seen in Table One. All wild type roots have negative beta values besides wild type root two indicating that the outer cell growth starts earlier than the inner cell growth. The insignificant beta values for wild type roots two and seven indicate that there is no significant measurable difference in the inner and outer cell length for these samples.

On the other hand, the frequentist model on the mutant type had less statistically significant beta estimates for the roots, as seen in Table Two. Mutant type roots 1, 4, 5, 6, 7, and 9 have statistically significant beta estimates that are both positive and negative values, showing that the inner or outer measurements might not necessarily be larger than the other but that there is a difference. The rest of the mutant type roots had insignificant beta values, suggesting that there is not a measurable difference between the inner and outer cell lengths. 

Our results ultimately give statistically significant evidence that the outer cell measurements are larger for the wild type and that there is not convincing evidence for a clear one-sided discrepancy for mutant type. 

The residual versus fitted plots for the wild type (@fig-resid-fitted-wt-freq) and mutant type roots (@fig-resid-fitted-mu-freq) show a fanning pattern while the residuals vs. midline plots for the wild type (@fig-resid-midline-wt-freq) and mutant type roots (fig-resid-midline-mu-freq) are scattered and evenly distanced along the midline. This shows that our frequentist model was a good fit for both the mutant type root data and the wild type root data. 

The ACF plot (@fig-freq-acf) suggests that there is autocorrelation that our model might not necessarily account for and that our model may have not captured the patterns in the data, as the bars exceed the bounds. 

## Bayesian Model 

## Model Rationale

## Model Implementation

```{r}
library(rjags)
library(R2jags)
library(coda)
```

```{r}
# Normalize Midline to the [0,1] range
rootData <- rootData |>
  group_by(root) |>
  mutate(x01 = (Midline - min(Midline)) / (max(Midline) - min(Midline))) |>
  ungroup()
```


```{r}
L_lower <- min(rootData$Length, na.rm = TRUE)
L_upper <- max(rootData$Length, na.rm = TRUE)
L_span  <- L_upper - L_lower

bdat <- list(
  Length = rootData$Length,
  N = nrow(rootData),
  x01 = rootData$x01,
  side = as.numeric(rootData$side == "Outer"),
  geno = as.numeric(rootData$Genotype == "MU"),
  root = as.numeric(as.factor(rootData$root)),
  n.root = length(unique(rootData$root)),
  L_lower = L_lower,
  L_span  = L_span
)

fun.params <- c("A", "B", "alpha", "scal",
                "gamma1", "gamma2", "gamma3",
                "sigma",
                "mu_a", "mu_d", "mu_z0", "mu_eta_s",
                "prec_a", "prec_d", "prec_z0", "prec_eta_s")
```

```{r}
get_nls_summary <- function(model) {
  s <- summary(model)
  est <- s$coefficients[, 1]
  se  <- s$coefficients[, 2]
  tibble(A = est["A"], A_se = se["A"],
         B = est["B"], B_se = se["B"],
         scal = est["scal"], scal_se = se["scal"],
         xmid = est["xmid"], xmid_se = se["xmid"])
}

WT_results_full <- map_dfr(WT_model_output, get_nls_summary, .id = "root")
MU_results_full <- map_dfr(MU_model_output, get_nls_summary, .id = "root")
all_results <- bind_rows(WT_results_full, MU_results_full)
```

```{r}
root_levels <- levels(as.factor(rootData$root))
nroot <- length(root_levels)
A_init <- B_init <- scal_init <- alpha_init <- numeric(nroot)
A_sd <- B_sd <- scal_sd <- alpha_sd <- numeric(nroot)

for (i in seq_along(root_levels)) {
  this_name <- paste0("root_", root_levels[i])
  row_i <- all_results[all_results$root == this_name, ]
  rd_i <- filter(rootData, root == root_levels[i])

  # we have fits for every root
  A_init[i] <- row_i$A
  A_sd[i]   <- row_i$A_se
  B_init[i] <- row_i$B
  B_sd[i]   <- row_i$B_se

  # Get min and max for this root
  mid_min <- min(rd_i$Midline)
  mid_max <- max(rd_i$Midline)
  span    <- mid_max - mid_min

  # Normalize both xmid and scal
  alpha_init[i] <- (row_i$xmid - mid_min) / span
  alpha_sd[i]   <- row_i$xmid_se / span
  scal_init[i]  <- row_i$scal / span
  scal_sd[i]    <- row_i$scal_se / span
}
```

```{r}
fun.inits <- function(chain_id = 1) {
  sigma.start <- runif(1, 1, 5)

  # Overdisperse latent parameters around NLS estimates
  a.start     <- log(pmax(1e-3, rnorm(bdat$n.root, mean = A_init, sd = 5 * A_sd)))
  d.start     <- log(pmax(1e-3, rnorm(bdat$n.root, mean = B_init - A_init, sd = 5 * B_sd)))
  z0.start    <- qlogis(
    pmin(0.99, pmax(0.01, rnorm(bdat$n.root, mean = alpha_init, sd = 5 * alpha_sd))))
  eta_s.start <- log(pmax(1e-3, rnorm(bdat$n.root, mean = scal_init, sd = 5 * scal_sd)))

  # Fixed effects (gamma terms) — random small perturbations
  gamma1.start <- rnorm(1, mean = 0, sd = 0.01)
  gamma2.start <- rnorm(1, mean = 0, sd = 0.01)
  gamma3.start <- rnorm(1, mean = 0, sd = 0.01)
  
  return(list(
    a = a.start,
    d = d.start,
    z0 = z0.start,
    eta_s = eta_s.start,
    gamma1 = gamma1.start,
    gamma2 = gamma2.start,
    gamma3 = gamma3.start,
    sigma = sigma.start
  ))
}

init_list <- list(
  fun.inits(1),
  fun.inits(2),
  fun.inits(3),
  fun.inits(4)
)
```

```{r}
fun.model.file <- "model_root.R" 
mcmc.pars <- list(iter = 50000, burn = 10000, nchain=4, thin=1)

jags.out <- jags(
  data = bdat,
  inits = init_list,
  parameters.to.save = fun.params,
  model.file = fun.model.file,
  n.chains = mcmc.pars$nchain,
  n.iter = mcmc.pars$iter,
  n.burnin = mcmc.pars$burn,
  n.thin = mcmc.pars$thin,
  DIC = FALSE,
  progress.bar = "none"
)
```



# Results

## Frequentist Model Evaluation
```{r}
WT_residuals <- data.frame()

for (name in names(WT_model_output)) {
  model <- WT_model_output[[name]]
  root_label <- str_replace(name, "root_", "")
    curr_data <- rootData_WT |> filter(root == root_label)
  
  curr_data <- curr_data |>
    mutate(
      Fitted = fitted(model),
      Residuals = resid(model)
    )
  
  WT_residuals <- bind_rows(WT_residuals, curr_data)
}
```

```{r}
ggplot(WT_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (WT)", x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

ggplot(WT_residuals, aes(x = Midline, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (WT)", x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()

ggplot(WT_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ root, scales = "free_x") +
  labs(title = "Residuals vs Fitted by Root (WT)", x = "Fitted Length", y = "Residuals") +
  theme_minimal()
```

```{r}
MU_residuals <- data.frame()

for (name in names(MU_model_output)) {
  model <- MU_model_output[[name]]
  root_label <- str_replace(name, "root_", "")
    curr_data <- rootData_MU |> filter(root == root_label)
  
  curr_data <- curr_data |>
    mutate(
      Fitted = fitted(model),
      Residuals = resid(model)
    )
  
  MU_residuals <- bind_rows(MU_residuals, curr_data)
}
```

```{r}
ggplot(MU_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (MU)", x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

ggplot(MU_residuals, aes(x = Midline, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (MU)", x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()

ggplot(MU_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ root, scales = "free_x") +
  labs(title = "Residuals vs Fitted by Root (MU)", x = "Fitted Length", y = "Residuals") +
  theme_minimal()
```

```{r}
combined_residuals <- bind_rows(
  WT_residuals %>% mutate(Genotype = "WT"),
  MU_residuals %>% mutate(Genotype = "MU")
)

ggplot(combined_residuals, aes(x = Fitted, y = Residuals, color = Genotype)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Residuals vs Fitted by Genotype",
    x = "Fitted Cell Length",
    y = "Residuals"
  ) +
  theme_minimal()
```

```{r}
# ACF plot (autocorr check)
m0 <- lm(Length ~ side * poly(Midline, 2), data = rootData)
acf(resid(m0))  # if this shows notable autocorr, ARMA in the next step is justified
```


## Bayesian Model Evaluation 

```{r}
# Normalize Midline to the [0,1] range
rootData <- rootData |>
  group_by(root) |>
  mutate(x01 = (Midline - min(Midline)) / (max(Midline) - min(Midline))) |>
  ungroup()
```

```{r}
L_lower <- min(rootData$Length, na.rm = TRUE)
L_upper <- max(rootData$Length, na.rm = TRUE)
L_span  <- L_upper - L_lower

bdat <- list(
  Length = rootData$Length,
  N = nrow(rootData),
  x01 = rootData$x01,
  side = as.numeric(rootData$side == "Outer"),
  geno = as.numeric(rootData$Genotype == "MU"),
  root = as.numeric(as.factor(rootData$root)),
  n.root = length(unique(rootData$root)),
  L_lower = L_lower,
  L_span  = L_span
)

fun.params <- c("A", "B", "alpha", "scal",
                "gamma1", "gamma2", "gamma3",
                "sigma",
                "mu_a", "mu_d", "mu_z0", "mu_eta_s",
                "prec_a", "prec_d", "prec_z0", "prec_eta_s")
```

```{r}
get_nls_summary <- function(model) {
  s <- summary(model)
  est <- s$coefficients[, 1]
  se  <- s$coefficients[, 2]
  tibble(A = est["A"], A_se = se["A"],
         B = est["B"], B_se = se["B"],
         scal = est["scal"], scal_se = se["scal"],
         xmid = est["xmid"], xmid_se = se["xmid"])
}

WT_results_full <- map_dfr(WT_model_output, get_nls_summary, .id = "root")
MU_results_full <- map_dfr(MU_model_output, get_nls_summary, .id = "root")
all_results <- bind_rows(WT_results_full, MU_results_full)
```

```{r}
root_levels <- levels(as.factor(rootData$root))
nroot <- length(root_levels)
A_init <- B_init <- scal_init <- alpha_init <- numeric(nroot)
A_sd <- B_sd <- scal_sd <- alpha_sd <- numeric(nroot)

for (i in seq_along(root_levels)) {
  this_name <- paste0("root_", root_levels[i])
  row_i <- all_results[all_results$root == this_name, ]
  rd_i <- filter(rootData, root == root_levels[i])

  # we have fits for every root
  A_init[i] <- row_i$A
  A_sd[i]   <- row_i$A_se
  B_init[i] <- row_i$B
  B_sd[i]   <- row_i$B_se

  # Get min and max for this root
  mid_min <- min(rd_i$Midline)
  mid_max <- max(rd_i$Midline)
  span    <- mid_max - mid_min

  # Normalize both xmid and scal
  alpha_init[i] <- (row_i$xmid - mid_min) / span
  alpha_sd[i]   <- row_i$xmid_se / span
  scal_init[i]  <- row_i$scal / span
  scal_sd[i]    <- row_i$scal_se / span
}
```

```{r}
fun.inits <- function(chain_id = 1) {
  sigma.start <- runif(1, 1, 5)

  # Overdisperse latent parameters around NLS estimates
  a.start     <- log(pmax(1e-3, rnorm(bdat$n.root, mean = A_init, sd = 5 * A_sd)))
  d.start     <- log(pmax(1e-3, rnorm(bdat$n.root, mean = B_init - A_init, sd = 5 * B_sd)))
  z0.start    <- qlogis(
    pmin(0.99, pmax(0.01, rnorm(bdat$n.root, mean = alpha_init, sd = 5 * alpha_sd))))
  eta_s.start <- log(pmax(1e-3, rnorm(bdat$n.root, mean = scal_init, sd = 5 * scal_sd)))

  # Fixed effects (gamma terms) — random small perturbations
  gamma1.start <- rnorm(1, mean = 0, sd = 0.01)
  gamma2.start <- rnorm(1, mean = 0, sd = 0.01)
  gamma3.start <- rnorm(1, mean = 0, sd = 0.01)
  
  return(list(
    a = a.start,
    d = d.start,
    z0 = z0.start,
    eta_s = eta_s.start,
    gamma1 = gamma1.start,
    gamma2 = gamma2.start,
    gamma3 = gamma3.start,
    sigma = sigma.start
  ))
}

init_list <- list(
  fun.inits(1),
  fun.inits(2),
  fun.inits(3),
  fun.inits(4)
)
```

```{r}
fun.model.file <- "model_root.R" 
mcmc.pars <- list(iter = 50000, burn = 10000, nchain=4, thin=1)

jags.out <- jags(
  data = bdat,
  inits = init_list,
  parameters.to.save = fun.params,
  model.file = fun.model.file,
  n.chains = mcmc.pars$nchain,
  n.iter = mcmc.pars$iter,
  n.burnin = mcmc.pars$burn,
  n.thin = mcmc.pars$thin,
  DIC = FALSE,
  progress.bar = "none"
)
```

## Model Evaluation 

### Gelman-Rubin Diagnostics

Provides an upper bound on variation - this provides confidence that the sampler is not still moving towards the posterior mode (can provide evidence that the poster is multi-modal and highlight an inefficient (slow mixing) sampler)

```{r}
# Gelman Plot
sim <- as.mcmc(jags.out)
pdf("grPlots.pdf", width=10, height=8)
par(mfrow=c(3,2), las=1)
gelman.plot(sim, las=1, transform=TRUE, ask=FALSE)
dev.off()
```

### Effective Size Results

Answers: How many independent samples are roughly equivalent in information content to this correlated sample? Variation value at zero is used to estimate effective sample size

```{r}
library(data.table)
ess <- effectiveSize(sim)
```

```{r}
A_vals = c()
for(idx in 1:20){
  A_vals = c(A_vals, ess[[idx]])
}

alpha_vals = c()
for(idx in 21:40){
  print(ess[idx])
  alpha_vals = c(alpha_vals, ess[[idx]])
}

B_vals = c()
for(idx in 41:60){
  B_vals = c(B_vals, ess[[idx]])
}

gamma_vals = c()
for(idx in 61:63){
  gamma_vals = c(gamma_vals, ess[[idx]])
}

scal_vals = c()
for(idx in 64:83){
  scal_vals = c(scal_vals, ess[[idx]])
}

sigma = ess[[84]]
tau = ess[[85]]

effective_dt <- data.table(
  x = c(1, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 2, 20, 3, 4, 5, 6, 7, 8, 9),
  type = c("WT", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "WT", "MU", "WT", "WT", "WT", "WT", "WT", "WT", "WT"),
  A = A_vals,
  alpha = alpha_vals,
  B = B_vals,
  scal = scal_vals
)
```

```{r}
library(patchwork)

p1 <- ggplot(data=effective_dt, aes(x=x, y=A, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "A Effective Sample Size",
    title = "A Effective Sample Size\nfor each Root",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 600)) +
  annotate("label", x=7, y=600, label=as.character(as.integer(max(effective_dt$A))), size=2)

p2 <- effective_dt |>
  ggplot(aes(x=x, y=B, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "B Effective Sample Size",
    title = "B Effective Sample Size\nfor each Root",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 1000)) +
  annotate("label", x=7, y=1000, label=as.character(as.integer(max(effective_dt$B))), size=2)

p3 <- effective_dt |>
  ggplot(aes(x=x, y=alpha, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Alpha Effective Sample Size",
    title = "Alpha Effective Sample Size\nfor each Root",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 200)) +
  annotate("label", x=15, y=200, label=as.character(as.integer(max(effective_dt$alpha))), size=2)

p4 <- effective_dt |>
  ggplot(aes(x=x, y=scal, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Scale Effective Sample Size",
    title = "Scale Effective Sample Size\nfor each Root",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 600)) +
  annotate("label", x=7, y=600, label=as.character(as.integer(max(effective_dt$scal))), size=2)

plot19 <- (p1 + p2) / (p3 + p4)
```

```{r}
gamma_vals
sigma
tau
```

### Raftery-Lewis Diagnostic

Number of iterations required to estimate a specified quantity q of the marginal posterior distribution of each parameter within a given accuracy (+- r) with specified probability (s)

```{r}
rl <- raftery.diag(sim)
```

```{r}
# Raftery-Lewis Diagnostic 
# Burn-in: Number of initial iterations to discard for the chain to reach a stationary distribution (small = fast convergence)
# Total N: Number of iterations needed to estimate specific quantile (Q=0.025) with desired precision (a value closer to lower bound indicates good efficiency)
# Lower Bound Nmin: theoretical minimum number of samples needed if the samples were independent (zero autocorrelation)
# Dependence Factor: extent to which autocorrelation inflates the required sample size (N - N/N_min) - a value close to 1 is valuable, indicating low autocorrelation and efficient sampling 
```

#### Burn-In

```{r}
# Burn-In for RL - Mean Across Chains 
A_rl = c()
A_df_i = c()
for(idx in 1:20){
  A_rl = c(A_rl, mean(c(rl[[1]][2][[1]][idx], rl[[2]][2][[1]][idx], rl[[3]][2][[1]][idx], rl[[4]][2][[1]][idx])))
  new_idx = idx + (85*3)
  A_df_i = c(A_df_i, mean(c(rl[[1]][2][[1]][new_idx], rl[[2]][2][[1]][new_idx], rl[[3]][2][[1]][new_idx], rl[[4]][2][[1]][new_idx])))
}

alpha_rl = c()
alpha_df_i = c()
for(idx in 21:40){
  alpha_rl = c(alpha_rl, mean(c(rl[[1]][2][[1]][idx], rl[[2]][2][[1]][idx], rl[[3]][2][[1]][idx], rl[[4]][2][[1]][idx])))
  new_idx = idx + (85*3)
  alpha_df_i = c(alpha_df_i, mean(c(rl[[1]][2][[1]][new_idx], rl[[2]][2][[1]][new_idx], rl[[3]][2][[1]][new_idx], rl[[4]][2][[1]][new_idx])))
}

B_rl = c()
B_df_i = c()
for(idx in 41:60){
  B_rl = c(B_rl, mean(c(rl[[1]][2][[1]][idx], rl[[2]][2][[1]][idx], rl[[3]][2][[1]][idx], rl[[4]][2][[1]][idx])))
  new_idx = idx + (85*3)
  B_df_i = c(B_df_i, mean(c(rl[[1]][2][[1]][new_idx], rl[[2]][2][[1]][new_idx], rl[[3]][2][[1]][new_idx], rl[[4]][2][[1]][new_idx])))
}

gamma_rl = c()
gamma_df_i = c()
for(idx in 61:62){
  gamma_rl = c(gamma_rl, mean(c(rl[[1]][2][[1]][idx], rl[[2]][2][[1]][idx], rl[[3]][2][[1]][idx], rl[[4]][2][[1]][idx])))
  new_idx = idx + (85*3)
  gamma_df_i = c(gamma_df_i, mean(c(rl[[1]][2][[1]][new_idx], rl[[2]][2][[1]][new_idx], rl[[3]][2][[1]][new_idx], rl[[4]][2][[1]][new_idx])))
}

mu_a_rl = mean(c(rl[[1]][2][[1]][63], rl[[2]][2][[1]][63], rl[[3]][2][[1]][63], rl[[4]][2][[1]][63]))
mu_a_df_i = mean(c(rl[[1]][2][[1]][63 + (91 * 3)], rl[[2]][2][[1]][63 + (91 * 3)], rl[[3]][2][[1]][63 + (91 * 3)], rl[[4]][2][[1]][63 + (91 * 3)]))

mu_d_rl = mean(c(rl[[1]][2][[1]][64], rl[[2]][2][[1]][64], rl[[3]][2][[1]][64], rl[[4]][2][[1]][64]))
mu_d_df_i = mean(c(rl[[1]][2][[1]][64 + (91 * 3)], rl[[2]][2][[1]][64 + (91 * 3)], rl[[3]][2][[1]][64 + (91 * 3)], rl[[4]][2][[1]][64 + (91 * 3)]))

mu_eta_s_rl = mean(c(rl[[1]][2][[1]][65], rl[[2]][2][[1]][65], rl[[3]][2][[1]][65], rl[[4]][2][[1]][65]))
mu_eta_s_df_i = mean(c(rl[[1]][2][[1]][65 + (91 * 3)], rl[[2]][2][[1]][65 + (91 * 3)], rl[[3]][2][[1]][65 + (91 * 3)], rl[[4]][2][[1]][65 + (91 * 3)]))

mu_z0_rl = mean(c(rl[[1]][2][[1]][66], rl[[2]][2][[1]][66], rl[[3]][2][[1]][66], rl[[4]][2][[1]][66]))
mu_z0_df_i = mean(c(rl[[1]][2][[1]][66 + (91 * 3)], rl[[2]][2][[1]][66 + (91 * 3)], rl[[3]][2][[1]][66 + (91 * 3)], rl[[4]][2][[1]][66 + (91 * 3)]))

prec_a_rl = mean(c(rl[[1]][2][[1]][67], rl[[2]][2][[1]][67], rl[[3]][2][[1]][67], rl[[4]][2][[1]][67]))
prec_a_df_i = mean(c(rl[[1]][2][[1]][67 + (91 * 3)], rl[[2]][2][[1]][67 + (91 * 3)], rl[[3]][2][[1]][67 + (91 * 3)], rl[[4]][2][[1]][67 + (91 * 3)]))

prec_d_rl = mean(c(rl[[1]][2][[1]][68], rl[[2]][2][[1]][68], rl[[3]][2][[1]][68], rl[[4]][2][[1]][68]))
prec_d_df_i = mean(c(rl[[1]][2][[1]][68 + (91 * 3)], rl[[2]][2][[1]][68 + (91 * 3)], rl[[3]][2][[1]][68 + (91 * 3)], rl[[4]][2][[1]][68 + (91 * 3)]))

prec_eta_s_rl = mean(c(rl[[1]][2][[1]][69], rl[[2]][2][[1]][69], rl[[3]][2][[1]][69], rl[[4]][2][[1]][69]))
prec_eta_s_df_i = mean(c(rl[[1]][2][[1]][69 + (91 * 3)], rl[[2]][2][[1]][69 + (91 * 3)], rl[[3]][2][[1]][69 + (91 * 3)], rl[[4]][2][[1]][69 + (91 * 3)]))

prec_z0_rl = mean(c(rl[[1]][2][[1]][70], rl[[2]][2][[1]][70], rl[[3]][2][[1]][70], rl[[4]][2][[1]][70]))
prec_z0_df_i = mean(c(rl[[1]][2][[1]][70 + (91 * 3)], rl[[2]][2][[1]][70 + (91 * 3)], rl[[3]][2][[1]][70 + (91 * 3)], rl[[4]][2][[1]][70 + (91 * 3)]))

scal_rl = c()
scal_df_i = c()
for(idx in 71:90){
  scal_rl = c(scal_rl, mean(c(rl[[1]][2][[1]][idx], rl[[2]][2][[1]][idx], rl[[3]][2][[1]][idx], rl[[4]][2][[1]][idx])))
  new_idx = idx + (85*3)
  scal_df_i = c(scal_df_i, mean(c(rl[[1]][2][[1]][new_idx], rl[[2]][2][[1]][new_idx], rl[[3]][2][[1]][new_idx], rl[[4]][2][[1]][new_idx])))
}

sigma_rl = mean(c(rl[[1]][2][[1]][91], rl[[2]][2][[1]][91], rl[[3]][2][[1]][91], rl[[4]][2][[1]][91]))
sigma_df_i = mean(c(rl[[1]][2][[1]][(91 * 4)], rl[[2]][2][[1]][(91 * 4)], rl[[3]][2][[1]][(91 * 4)], rl[[4]][2][[1]][(91 * 3)]))

rl_dt <- data.table(
  x = c(1, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 2, 20, 3, 4, 5, 6, 7, 8, 9),
  type = c("WT", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "WT", "MU", "WT", "WT", "WT", "WT", "WT", "WT", "WT"),
  A_rl = A_rl,
  alpha_rl = alpha_rl,
  B_rl = B_rl,
  scal_rl = scal_rl,
  A_df_i = A_df_i,
  alpha_df_i = alpha_df_i,
  B_df_i = B_df_i,
  scal_df_i = scal_df_i
)
```

```{r}
# Plots for Burn-In RL for Mean of Chains
p1 <- rl_dt |>
  ggplot(aes(x=x, y=A_rl, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "A Burn-In",
    title = "A Burn-In\nfor each Root",
    fill = "Root Type"
  ) 

p2 <- rl_dt |>
  ggplot(aes(x=x, y=B_rl, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "B Burn-In",
    title = "B Burn-In\nfor each Root",
    fill = "Root Type"
  ) 

p3 <- rl_dt |>
  ggplot(aes(x=x, y=alpha_rl, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Alpha Burn-In",
    title = "Alpha Burn-In\nfor each Root",
    fill = "Root Type"
  )  

p4 <- rl_dt |>
  ggplot(aes(x=x, y=scal_rl, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Scale Burn-In",
    title = "Scale Burn-In\nfor each Root",
    fill = "Root Type"
  ) 

plot20 <- (p1 + p2) / (p3 + p4)
```

```{r}
# Gamma, Sigma, Tau for Burn-In 
rl_dt2 <- data.table(
  gamma1 = gamma_rl[1],
  gamma2 = gamma_rl[2],
  mu_a = mu_a_rl,
  mu_d = mu_d_rl,
  mu_eta_s = mu_eta_s_rl,
  mu_z0 = mu_z0_rl,
  prec_a = prec_a_rl,
  prec_d = prec_d_rl,
  prec_eta_s = prec_eta_s_rl,
  prec_z0 = prec_z0_rl
)

table3 <- kable(rl_dt2)
```

#### Dependence Factor

```{r}
# Plots for Dependence Factor RL for Mean of Chains
p1 <- rl_dt |>
  ggplot(aes(x=x, y=A_df_i, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "A Dependence Factor",
    title = "A Dependence Factor\nfor each Root",
    fill = "Root Type"
  ) 

p2 <- rl_dt |>
  ggplot(aes(x=x, y=B_df_i, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "B Dependence Factor",
    title = "B Dependence Factor\nfor each Root",
    fill = "Root Type"
  ) 

p3 <- rl_dt |>
  ggplot(aes(x=x, y=alpha_df_i, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Alpha Dependence Factor",
    title = "Alpha Dependence Factor\nfor each Root",
    fill = "Root Type"
  )  

p4 <- rl_dt |>
  ggplot(aes(x=x, y=scal_df_i, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Scale Dependence Factor",
    title = "Scale Dependence Factor\nfor each Root",
    fill = "Root Type"
  ) 

plot21 <- (p1 + p2) / (p3 + p4)
```

```{r}
df_i_dt2 <- data.table(
  gamma1 = gamma_df_i[1],
  gamma2 = gamma_df_i[2],
  mu_a = mu_a_df_i,
  mu_d = mu_d_df_i,
  mu_eta_s = mu_eta_s_df_i,
  mu_z0 = mu_z0_df_i,
  prec_a = prec_a_df_i,
  prec_d = prec_d_df_i,
  prec_eta_s = prec_eta_s_df_i,
  prec_z0 = prec_z0_df_i
)

table4 <- kable(df_i_dt2)
```

### ACF Plots (Autocorrelation)

```{r}
pdf("acfPlots.pdf", width=10, height=8)
par(mfrow=c(3, 3), las=1)
autocorr.plot(sim, lag.max=10, auto.layout=FALSE, ask=FALSE)
dev.off()
```

### posterior summary: CODA Output Analysis Functions

```{r}
post_summary <-summary(sim)
post_summary
```

```{r}
library(DT)

# Mean
means_WT = c()
means_MU = c()

# A, alpha, B
for(i in 0:3){
  s = i*20
  if(s == 3){
    s = s + 10
  }
  means_WT = c(means_WT, mean(ps[s+1], ps[s+12], ps[s+14], ps[s+15], ps[s+16], ps[s+17], ps[s+18], ps[s+19], ps[s+20]))
  means_MU = c(means_MU, mean(ps[s+2], ps[s+3], ps[s+4], ps[s+5], ps[s+6], ps[s+7], ps[s+8], ps[s+9], ps[s+10], ps[s+11], ps[s+13]))
}

# SD
sd_WT = c()
sd_MU = c()

# A, alpha, B
for(i in 0:3){
  s = i*20 + 93
  if(s == 3){
    s = s + 10
  }
  sd_WT = c(sd_WT, mean(ps[s+1], ps[s+12], ps[s+14], ps[s+15], ps[s+16], ps[s+17], ps[s+18], ps[s+19], ps[s+20]))
  sd_MU = c(sd_MU, mean(ps[s+2], ps[s+3], ps[s+4], ps[s+5], ps[s+6], ps[s+7], ps[s+8], ps[s+9], ps[s+10], ps[s+11], ps[s+13]))
}

ps_df <- data.table(
  type = c("WT", "MU"),
  A_mean = c(means_WT[1], means_MU[1]), 
  A_sd = c(sd_WT[1], sd_MU[1]),
  alpha_mean = c(means_WT[2], means_MU[2]),
  alpha_sd = c(sd_WT[2], sd_MU[2]),
  B_mean = c(means_WT[3], means_MU[3]),
  B_sd = c(sd_WT[3], sd_MU[3]),
  scal_mean = c(means_WT[4], means_MU[4]),
  scal_sd = c(sd_WT[4], sd_MU[4])
)

indices = c(61, 62, 63, 64, 65, 66, 67, 68, 69, 70, 91)
means_alt = c()
sd_alt = c()
naive_se_alt = c()
ts_se_alt = c()

for(idx in indices){
  means_alt = c(means_alt, ps[idx])
  sd_alt = c(sd_alt, ps[91 + idx])
  naive_se_alt = c(naive_se_alt, ps[91*2 + idx])
  ts_se_alt = c(ts_se_alt, ps[91*3 + idx])
}

ps_q = post_summary[[2]]

quant_2_5_alt = c()
quant_25_alt = c()
quant_50_alt = c()
quant_75_alt = c()
quant_97_5_alt = c()

for(idx in indices){
  quant_2_5_alt = c(quant_2_5_alt, ps_q[idx])
  quant_25_alt = c(quant_25_alt, ps_q[91 + idx])
  quant_50_alt = c(quant_50_alt, ps_q[91*2 + idx])
  quant_75_alt = c(quant_75_alt, ps_q[91*3 + idx])
  quant_97_5_alt = c(quant_97_5_alt, ps_q[91*4 + idx])
}

alternate_params_df <- data.table(
  parameter = c("Gamma1", "Gamma2", "Mu_A", "Mu_D", "Mu_Eta_S", "Mu_Z0", "Prec_A", "Prec_S", "Prec_Eta_S", "Prec_Z0", "Sigma"),
  mean = means_alt, 
  sd = sd_alt,
  naive_se = naive_se_alt,
  time_series_se = ts_se_alt,
  quantile_2_5 = quant_2_5_alt,
  quantile_25 = quant_25_alt,
  quantile_50 = quant_50_alt,
  quantile_75 = quant_75_alt,
  quantile_97_5 = quant_97_5_alt
)

kable(alternate_params_df)

kable(ps_df)
```

## Results

# Limitations and Conclusion

# Appendix 
```{r}
#| tbl-cap: "Summary of wild type frequentist results."
table1
```

```{r}
#| tbl-cap: "Summary of mutant type frequentist results."
table2
```

```{r}
#| tbl-cap: "Raftery-Lewis Burn-In Results."
table3
```

```{r}
#| tbl-cap: "Raftery-Lewis Dependence Factor Results."
table4
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-dist
#| fig-cap: "Distribution of cell length for inner and outer cells."
plot1
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-wt
#| fig-cap: "Distribution of cell length for each wild root."
plot2
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-mu
#| fig-cap: "Distribution of cell length for each mutant root."
plot3
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-smooth-wt
#| fig-cap: "Distribution of cell length for each wild root."
plot4
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-smooth-mu
#| fig-cap: "Distribution of cell length for each mutant root."
plot5
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-smooth-wt
#| fig-cap: "Distribution of cell length for wild type."
plot6
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-smooth-wt
#| fig-cap: "Distribution of cell length for mutant type."
plot7
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-smooth-wt
#| fig-cap: "Logistic fit of cell length for wild type."
plot8
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-smooth-wu
#| fig-cap: "Logistic fit of cell length for mutant type."
plot9
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-mean-cell-length-per-root-wt
#| fig-cap: "Distribution of mean cell length for each wild root."
plot10
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-smooth-mu
#| fig-cap: "Distribution of mean cell length for each mutant root."
plot11
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-resid-fitted-wt-freq
#| fig-cap: "Residuals vs. Fitted for Wild Type Frequentist Model."
plot12
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-resid-midline-wt-freq
#| fig-cap: "Residuals vs. Midline for Wild Type Frequentist Model."
plot13
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-resid-fitted-by-root-wt-freq
#| fig-cap: "Residuals vs. Fitted by Root for Wild Type Frequentist Model."
plot14
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-resid-fitted-mu-freq
#| fig-cap: "Residuals vs. Fitted for Mutant Type Frequentist Model."
plot15
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-resid-midline-mu-freq
#| fig-cap: "Residuals vs. Midline for Mutant Type Frequentist Model."
plot16
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-resid-fitted-by-root-mu-freq
#| fig-cap: "Residuals vs. Fitted by Root for Mutant Type Frequentist Model."
plot17
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-freq-acf
#| fig-cap: "Frequentist Model acf plot."
acf(resid(m0))
```

