---
title: "case4_code"
format: html
editor: visual
---

```{r}
load("./rootData.RData")
```

```{r}
library(ggplot2)
library(dplyr)
library(nlme)
library(broom)
library(knitr)
library(stringr)
library(tidyverse)
```

# EDA

```{r}
ggplot(rootData, aes(x = side, y = Length, fill = side)) +
  geom_boxplot() +
  facet_wrap(~ Genotype) +
  labs(title = "Cell Length Distribution by Inner vs Outer",
       y = "Cell Length") +
  theme_bw()
```

```{r}
rootData_WT <- rootData |> filter(Genotype == "WT")

ggplot(rootData_WT, aes(x = Midline, y = Length, 
                        color = side, group = interaction(root, side))) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "Cell Length Along Midline – Wild Type Roots",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
rootData_MU <- rootData |> filter(Genotype == "MU")
desired_order <- c("MU1", "MU2", "MU3", "MU4", "MU5", "MU6", "MU7", "MU8", "MU9", "MU10", "MU11")
rootData_MU <- rootData_MU |>
  mutate(root = factor(root, levels = desired_order))

ggplot(rootData_MU, aes(x = Midline, y = Length, 
                        color = side, group = interaction(root, side))) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "Cell Length Along Midline – Mutant Roots",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
ggplot(rootData_WT,
       aes(x = Midline, y = Length, color = side, group = interaction(root, side))) +
  geom_smooth(method = "loess", se = FALSE, span = 0.75, size = 0.8) +
  geom_point(alpha = 0.4, size = 0.5) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "WT: Inner vs Outer Smoothed Trends for Each Root",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(rootData_MU,
       aes(x = Midline, y = Length, color = side, group = interaction(root, side))) +
  geom_smooth(method = "loess", se = FALSE, span = 0.75, size = 0.8) +
  geom_point(alpha = 0.4, size = 0.5) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "MT: Inner vs Outer Smoothed Trends for Each Root",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
ggplot(rootData_WT, aes(x = Midline, y = Length, color = side)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.8, size = 1.2) +
  labs(
    title = "WT: Inner vs Outer Overall Cell Length Trends",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw()

ggplot(rootData_MU, aes(x = Midline, y = Length, color = side)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.8, size = 1.2) +
  labs(
    title = "MU: Inner vs Outer Overall Cell Length Trends",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw()
```

Logistic

```{r warning=FALSE}
logis_smooth <- function(df) {
  ggplot(
    df,
    aes(x = Midline, y = Length, color = side, group = interaction(root, side))
  ) +
    geom_point(alpha = 0.4, size = 0.5) +
    geom_smooth(
      method = "nls",
      formula = y ~ SSlogis(x, Asym, xmid, scal),
      method.args = list(control = nls.control(maxiter = 200, warnOnly = TRUE)),
      se = FALSE,
      size = 0.8
    ) +
    facet_wrap(~ root, scales = "free_y") +
    labs(
      x = "Distance Along Midline",
      y = "Cell Length",
      color = "Side"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Wild type
logis_smooth(rootData_WT) + ggtitle("WT: Inner vs Outer Logistic Fits for Each Root")

# Mutant
logis_smooth(rootData_MU) + ggtitle("MU: Inner vs Outer Logistic Fits for Each Root")
```

```{r}
# Mean Length Difference 

# MU
diffDf <- rootData_MU |>
  mutate(diff_group = as.character(as.integer(Midline / 100) * 100)) |>
  group_by(diff_group, side) |>
  mutate(mean_length = mean(Length))

diffDf |>
  ggplot(aes(x=as.integer(diff_group), y = mean_length, color=side)) +
  geom_point() +
  labs(
    x = "Midline (Grouped)",
    y = "Mean Length",
    title = "Mean Length in Groups of 100 for Midline (Mutant)"
  )

# WT 
diffDf <- rootData_WT |>
  mutate(diff_group = as.character(as.integer(Midline / 100) * 100)) |>
  group_by(diff_group, side) |>
  mutate(mean_length = mean(Length))

diffDf |>
  ggplot(aes(x=as.integer(diff_group), y = mean_length, color=side)) +
  geom_point() +
  labs(
    x = "Midline (Grouped)",
    y = "Mean Length",
    title = "Mean Length in Groups of 100 for Midline (Wildtype)"
  )
```

```{r}
# Plot Average Length Per Genotype
rootData_WT |>
  ggplot(aes(x=root, y=Length, fill=side)) +
  geom_bar(stat="summary", fun="mean", position="dodge")

rootData_MU |>
  ggplot(aes(x=root, y=Length, fill=side)) +
  geom_bar(stat="summary", fun="mean", position="dodge")
```

# Frequentist Model

### Fit nls with customized function

$$
\text{Length ~} A+\frac{(B-A)}{1+e^{((x_{mid} + \beta * \text{side_ind} - x)/\text{scal})}}
$$

```{r}
fpl_customized <- function(x, A, B, xmid, scal, beta, side_ind){
  xmid_eff <- xmid + beta * side_ind
  A + (B - A) / (1 + exp((xmid_eff - x) / scal))
}
```

```{r}
WT_model_output <- list()

for (root_num in unique(rootData_WT$root)) {
  curr_root <- rootData_WT |> filter(root == root_num)
  curr_root$side_ind <- ifelse(curr_root$side == "Outer", 1, 0)

  y_min <- min(curr_root$Length)
  y_max <- max(curr_root$Length)

  start_vals <- list(
    A = y_min * 0.9,
    B = y_max * 1.1,
    xmid = median(curr_root$Midline),
    scal = diff(range(curr_root$Midline)) / 4,
    beta = -50
  )

  lower_bounds <- c(A = y_min * 0.5,
                    B = y_min,
                    xmid = min(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)) / 10,
                    beta = -500)

  upper_bounds <- c(A = y_max,
                    B = y_max * 2.0,
                    xmid = max(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)),
                    beta = 500)

  fit <- try(
    nls(
      Length ~ fpl_customized(Midline, A, B, xmid, scal, beta, side_ind),
      data = curr_root,
      start = start_vals,
      algorithm = "port",
      lower = lower_bounds,
      upper = upper_bounds,
      control = nls.control(maxiter = 1000, warnOnly = TRUE, minFactor = 1e-10)
    ),
    silent = TRUE
  )

  if (!inherits(fit, "try-error")) {
    WT_model_output[[paste0("root_", root_num)]] <- fit
  } else {
    message(paste("Model failed for root", root_num))
  }
}
```

```{r}
WT_results <- data.frame(
  root = character(),
  A = numeric(),
  B = numeric(),
  xmid = numeric(),
  scal = numeric(),
  beta = numeric(),
  beta_p = numeric(),
  beta_sig = character(),
  stringsAsFactors = FALSE
)

for(name in names(WT_model_output)){
  model <- WT_model_output[[name]]
  
  coef_info <- tidy(model)
  beta_row <- coef_info[coef_info$term == "beta", ]

  WT_results <- rbind(
    WT_results,
    data.frame(
      root = name,
      A = coef(model)["A"],
      B = coef(model)["B"],
      xmid = coef(model)["xmid"],
      scal = coef(model)["scal"],
      beta = coef(model)["beta"],
      beta_p = beta_row$p.value,
      beta_sig = ifelse(beta_row$p.value < 0.05, "*", "ns")
    )
  )
}

kable(WT_results, digits = 5, caption = "Wild Type Logistic Model Results")
```

```{r}
MU_model_output <- list()

for (root_num in unique(rootData_MU$root)) {
  curr_root <- rootData_MU |> filter(root == root_num)
  curr_root$side_ind <- ifelse(curr_root$side == "Outer", 1, 0)

  y_min <- min(curr_root$Length)
  y_max <- max(curr_root$Length)

  start_vals <- list(
    A = y_min * 0.9,
    B = y_max * 1.1,
    xmid = median(curr_root$Midline),
    scal = diff(range(curr_root$Midline)) / 4,
    beta = 0
  )

  lower_bounds <- c(A = y_min * 0.5,
                    B = y_min,
                    xmid = min(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)) / 10,
                    beta = -500)

  upper_bounds <- c(A = y_max,
                    B = y_max * 2.0,
                    xmid = max(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)),
                    beta = 500)

  fit <- try(
    nls(
      Length ~ fpl_customized(Midline, A, B, xmid, scal, beta, side_ind),
      data = curr_root,
      start = start_vals,
      algorithm = "port",
      lower = lower_bounds,
      upper = upper_bounds,
      control = nls.control(maxiter = 1000, warnOnly = TRUE, minFactor = 1e-10)
    ),
    silent = TRUE
  )

  if (!inherits(fit, "try-error")) {
    MU_model_output[[paste0("root_", root_num)]] <- fit
  } else {
    message(paste("Model failed for root", root_num))
  }
}
```

```{r}
MU_results <- data.frame(
  root = character(),
  A = numeric(),
  B = numeric(),
  xmid = numeric(),
  scal = numeric(),
  beta = numeric(),
  beta_p = numeric(),
  beta_sig = character(),
  stringsAsFactors = FALSE
)

for(name in names(MU_model_output)){
  model <- MU_model_output[[name]]
  
  coef_info <- tidy(model)
  beta_row <- coef_info[coef_info$term == "beta", ]

  MU_results <- rbind(
    MU_results,
    data.frame(
      root = name,
      A = coef(model)["A"],
      B = coef(model)["B"],
      xmid = coef(model)["xmid"],
      scal = coef(model)["scal"],
      beta = coef(model)["beta"],
      beta_p = beta_row$p.value,
      beta_sig = ifelse(beta_row$p.value < 0.05, "*", "ns")
    )
  )
}

kable(MU_results, digits = 5, caption = "Mutant Logistic Model Results")
```

### Evaluation

```{r}
WT_residuals <- data.frame()

for (name in names(WT_model_output)) {
  model <- WT_model_output[[name]]
  root_label <- str_replace(name, "root_", "")
    curr_data <- rootData_WT |> filter(root == root_label)
  
  curr_data <- curr_data |>
    mutate(
      Fitted = fitted(model),
      Residuals = resid(model)
    )
  
  WT_residuals <- bind_rows(WT_residuals, curr_data)
}
```

```{r}
ggplot(WT_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (WT)", x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

ggplot(WT_residuals, aes(x = Midline, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (WT)", x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()

ggplot(WT_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ root, scales = "free_x") +
  labs(title = "Residuals vs Fitted by Root (WT)", x = "Fitted Length", y = "Residuals") +
  theme_minimal()
```

```{r}
MU_residuals <- data.frame()

for (name in names(MU_model_output)) {
  model <- MU_model_output[[name]]
  root_label <- str_replace(name, "root_", "")
    curr_data <- rootData_MU |> filter(root == root_label)
  
  curr_data <- curr_data |>
    mutate(
      Fitted = fitted(model),
      Residuals = resid(model)
    )
  
  MU_residuals <- bind_rows(MU_residuals, curr_data)
}
```

```{r}
ggplot(MU_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (MU)", x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

ggplot(MU_residuals, aes(x = Midline, y = Residuals, color = side)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (MU)", x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()

ggplot(MU_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ root, scales = "free_x") +
  labs(title = "Residuals vs Fitted by Root (MU)", x = "Fitted Length", y = "Residuals") +
  theme_minimal()
```

```{r}
combined_residuals <- bind_rows(
  WT_residuals %>% mutate(Genotype = "WT"),
  MU_residuals %>% mutate(Genotype = "MU")
)

ggplot(combined_residuals, aes(x = Fitted, y = Residuals, color = Genotype)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Residuals vs Fitted by Genotype",
    x = "Fitted Cell Length",
    y = "Residuals"
  ) +
  theme_minimal()
```

```{r}
# ACF plot (autocorr check)
m0 <- lm(Length ~ side * poly(Midline, 2), data = rootData)
acf(resid(m0))  # if this shows notable autocorr, ARMA in the next step is justified
```

# Bayesian model

```{r}
library(rjags)
library(R2jags)
library(coda)
```

```{r}
# Normalize Midline to the [0,1] range
rootData <- rootData |>
  group_by(root) |>
  mutate(x01 = (Midline - min(Midline)) / (max(Midline) - min(Midline))) |>
  ungroup()
```

```{r}
# bdat <- list(
#   Length = rootData$Length,
#   N = nrow(rootData),
#   x01 = rootData$x01,
#   side = as.numeric(rootData$side == "Outer"),
#   geno = as.numeric(rootData$Genotype == "WT"),
#   root = as.numeric(as.factor(rootData$root)),
#   n.root = length(unique(rootData$root))
# )
```

```{r}
# fun.params <- c("A", "B", "scal", "alpha", 
#                 "gamma1", "gamma2", "gamma3",
#                 "sigma", "tau")
```

```{r}
# fun.model.file <- "model_root.R"
# mcmc.pars <- list(iter = 50000, burn = 10000, thin = 10)
# 
# fun.inits <- function() {
#   list(
#     A = rep(5, bdat$n.root),
#     B = rep(30, bdat$n.root),
#     scal = rep(0.2, bdat$n.root),
#     alpha = rep(0, bdat$n.root),
#     gamma1 = 0,
#     gamma2 = 0,
#     gamma3 = 0,
#     sigma = 5,
#     tau = 2
#   )
# }
# 
# jags.out <- jags(
#   data = bdat,
#   inits = fun.inits,
#   parameters.to.save = fun.params,
#   model.file = fun.model.file,
#   n.chains = 1,
#   n.iter = mcmc.pars$iter,
#   n.burnin = mcmc.pars$burn,
#   n.thin = mcmc.pars$thin,
#   DIC = FALSE,
#   progress.bar = "none"
# )
```

# MCMC version 2

-   chain 1-\> 4
-   A,B, scal, alpha starting values from frequentist

```{r}
L_lower <- min(rootData$Length, na.rm = TRUE)
L_upper <- max(rootData$Length, na.rm = TRUE)
L_span  <- L_upper - L_lower

bdat <- list(
  Length = rootData$Length,
  N = nrow(rootData),
  x01 = rootData$x01,
  side = as.numeric(rootData$side == "Outer"),
  geno = as.numeric(rootData$Genotype == "MU"),
  root = as.numeric(as.factor(rootData$root)),
  n.root = length(unique(rootData$root)),
  L_lower = L_lower,
  L_span  = L_span
)

fun.params <- c("A", "B", "alpha", "scal",
                "gamma1", "gamma2", "gamma3",
                "sigma",
                "mu_a", "mu_d", "mu_z0", "mu_eta_s",
                "prec_a", "prec_d", "prec_z0", "prec_eta_s")
```

```{r}
get_nls_summary <- function(model) {
  s <- summary(model)
  est <- s$coefficients[, 1]
  se  <- s$coefficients[, 2]
  tibble(A = est["A"], A_se = se["A"],
         B = est["B"], B_se = se["B"],
         scal = est["scal"], scal_se = se["scal"],
         xmid = est["xmid"], xmid_se = se["xmid"])
}

WT_results_full <- map_dfr(WT_model_output, get_nls_summary, .id = "root")
MU_results_full <- map_dfr(MU_model_output, get_nls_summary, .id = "root")
all_results <- bind_rows(WT_results_full, MU_results_full)
```

```{r}
root_levels <- levels(as.factor(rootData$root))
nroot <- length(root_levels)
A_init <- B_init <- scal_init <- alpha_init <- numeric(nroot)
A_sd <- B_sd <- scal_sd <- alpha_sd <- numeric(nroot)

for (i in seq_along(root_levels)) {
  this_name <- paste0("root_", root_levels[i])
  row_i <- all_results[all_results$root == this_name, ]
  rd_i <- filter(rootData, root == root_levels[i])

  # we have fits for every root
  A_init[i] <- row_i$A
  A_sd[i]   <- row_i$A_se
  B_init[i] <- row_i$B
  B_sd[i]   <- row_i$B_se

  # Get min and max for this root
  mid_min <- min(rd_i$Midline)
  mid_max <- max(rd_i$Midline)
  span    <- mid_max - mid_min

  # Normalize both xmid and scal
  alpha_init[i] <- (row_i$xmid - mid_min) / span
  alpha_sd[i]   <- row_i$xmid_se / span
  scal_init[i]  <- row_i$scal / span
  scal_sd[i]    <- row_i$scal_se / span
}
```

```{r}
fun.inits <- function(chain_id = 1) {
  sigma.start <- runif(1, 1, 5)

  # Overdisperse latent parameters around NLS estimates
  a.start     <- log(pmax(1e-3, rnorm(bdat$n.root, mean = A_init, sd = 5 * A_sd)))
  d.start     <- log(pmax(1e-3, rnorm(bdat$n.root, mean = B_init - A_init, sd = 5 * B_sd)))
  z0.start    <- qlogis(
    pmin(0.99, pmax(0.01, rnorm(bdat$n.root, mean = alpha_init, sd = 5 * alpha_sd))))
  eta_s.start <- log(pmax(1e-3, rnorm(bdat$n.root, mean = scal_init, sd = 5 * scal_sd)))

  # Fixed effects (gamma terms) — random small perturbations
  gamma1.start <- rnorm(1, mean = 0, sd = 0.01)
  gamma2.start <- rnorm(1, mean = 0, sd = 0.01)
  gamma3.start <- rnorm(1, mean = 0, sd = 0.01)
  
  return(list(
    a = a.start,
    d = d.start,
    z0 = z0.start,
    eta_s = eta_s.start,
    gamma1 = gamma1.start,
    gamma2 = gamma2.start,
    gamma3 = gamma3.start,
    sigma = sigma.start
  ))
}

init_list <- list(
  fun.inits(1),
  fun.inits(2),
  fun.inits(3),
  fun.inits(4)
)
```

```{r}
fun.model.file <- "model_root.R" 
mcmc.pars <- list(iter = 50000, burn = 10000, nchain=4, thin=1)

jags.out <- jags(
  data = bdat,
  inits = init_list,
  parameters.to.save = fun.params,
  model.file = fun.model.file,
  n.chains = mcmc.pars$nchain,
  n.iter = mcmc.pars$iter,
  n.burnin = mcmc.pars$burn,
  n.thin = mcmc.pars$thin,
  DIC = FALSE,
  progress.bar = "none"
)
```

```{r}
save(jags.out, bdat, mcmc.pars, file = "bayes_model_run1.RData")
```

```{r}
jags.out$BUGSoutput$summary[c("gamma1", "gamma2", "gamma3"), ]
```

```{r}
S <- as.mcmc(jags.out)
traceplot(S[, c("sigma", "prec_a", "prec_d", "prec_z0", "prec_eta_s")])
```

### Gelman-Rubin Diagnostics

Provides an upper bound on variation - this provides confidence that the sampler is not still moving towards the posterior mode (can provide evidence that the poster is multi-modal and highlight an inefficient (slow mixing) sampler)

```{r}
# Gelman Plot
sim <- as.mcmc(jags.out)
pdf("grPlots.pdf", width=10, height=8)
par(mfrow=c(3,2), las=1)
gelman.plot(sim, las=1, transform=TRUE, ask=FALSE)
dev.off()
```

```{r}
# Gelman Diag
gelman.diag(sim, transform=TRUE)
```

### Effective Size Results

Answers: How many independent samples are roughly equivalent in information content to this correlated sample? Variation value at zero is used to estimate effective sample size

```{r}
library(data.table)
ess <- effectiveSize(sim)
```

```{r}
A_vals = c()
for(idx in 1:20){
  A_vals = c(A_vals, ess[[idx]])
}

alpha_vals = c()
for(idx in 21:40){
  print(ess[idx])
  alpha_vals = c(alpha_vals, ess[[idx]])
}

B_vals = c()
for(idx in 41:60){
  B_vals = c(B_vals, ess[[idx]])
}

gamma_vals = c()
for(idx in 61:63){
  gamma_vals = c(gamma_vals, ess[[idx]])
}

scal_vals = c()
for(idx in 64:83){
  scal_vals = c(scal_vals, ess[[idx]])
}

sigma = ess[[84]]
tau = ess[[85]]

effective_dt <- data.table(
  x = c(1, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 2, 20, 3, 4, 5, 6, 7, 8, 9),
  type = c("WT", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "WT", "MU", "WT", "WT", "WT", "WT", "WT", "WT", "WT"),
  A = A_vals,
  alpha = alpha_vals,
  B = B_vals,
  scal = scal_vals
)
```

```{r}
library(patchwork)

p1 <- ggplot(data=effective_dt, aes(x=x, y=A, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "A Effective Sample Size",
    title = "A Effective Sample Size\nfor each Root",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 600)) +
  annotate("label", x=7, y=600, label=as.character(as.integer(max(effective_dt$A))), size=2)

p2 <- effective_dt |>
  ggplot(aes(x=x, y=B, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "B Effective Sample Size",
    title = "B Effective Sample Size\nfor each Root",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 1000)) +
  annotate("label", x=7, y=1000, label=as.character(as.integer(max(effective_dt$B))), size=2)

p3 <- effective_dt |>
  ggplot(aes(x=x, y=alpha, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Alpha Effective Sample Size",
    title = "Alpha Effective Sample Size\nfor each Root",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 200)) +
  annotate("label", x=15, y=200, label=as.character(as.integer(max(effective_dt$alpha))), size=2)

p4 <- effective_dt |>
  ggplot(aes(x=x, y=scal, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Scale Effective Sample Size",
    title = "Scale Effective Sample Size\nfor each Root",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 600)) +
  annotate("label", x=7, y=600, label=as.character(as.integer(max(effective_dt$scal))), size=2)

(p1 + p2) / (p3 + p4)
```

```{r}
gamma_vals
sigma
tau
```

### Raftery-Lewis Diagnostic

Number of iterations required to estimate a specified quantity q of the marginal posterior distribution of each parameter within a given accuracy (+- r) with specified probability (s)

```{r}
rl <- raftery.diag(sim)
```

```{r}
# Raftery-Lewis Diagnostic 
# Burn-in: Number of initial iterations to discard for the chain to reach a stationary distribution (small = fast convergence)
# Total N: Number of iterations needed to estimate specific quantile (Q=0.025) with desired precision (a value closer to lower bound indicates good efficiency)
# Lower Bound Nmin: theoretical minimum number of samples needed if the samples were independent (zero autocorrelation)
# Dependence Factor: extent to which autocorrelation inflates the required sample size (N - N/N_min) - a value close to 1 is valuable, indicating low autocorrelation and efficient sampling 
rl[[1]]
```

#### Burn-In

```{r}
# Burn-In for RL - Mean Across Chains 
A_rl = c()
A_df_i = c()
for(idx in 1:20){
  A_rl = c(A_rl, mean(c(rl[[1]][2][[1]][idx], rl[[2]][2][[1]][idx], rl[[3]][2][[1]][idx], rl[[4]][2][[1]][idx])))
  new_idx = idx + (85*3)
  A_df_i = c(A_df_i, mean(c(rl[[1]][2][[1]][new_idx], rl[[2]][2][[1]][new_idx], rl[[3]][2][[1]][new_idx], rl[[4]][2][[1]][new_idx])))
}

alpha_rl = c()
alpha_df_i = c()
for(idx in 21:40){
  alpha_rl = c(alpha_rl, mean(c(rl[[1]][2][[1]][idx], rl[[2]][2][[1]][idx], rl[[3]][2][[1]][idx], rl[[4]][2][[1]][idx])))
  new_idx = idx + (85*3)
  alpha_df_i = c(alpha_df_i, mean(c(rl[[1]][2][[1]][new_idx], rl[[2]][2][[1]][new_idx], rl[[3]][2][[1]][new_idx], rl[[4]][2][[1]][new_idx])))
}

B_rl = c()
B_df_i = c()
for(idx in 41:60){
  B_rl = c(B_rl, mean(c(rl[[1]][2][[1]][idx], rl[[2]][2][[1]][idx], rl[[3]][2][[1]][idx], rl[[4]][2][[1]][idx])))
  new_idx = idx + (85*3)
  B_df_i = c(B_df_i, mean(c(rl[[1]][2][[1]][new_idx], rl[[2]][2][[1]][new_idx], rl[[3]][2][[1]][new_idx], rl[[4]][2][[1]][new_idx])))
}

gamma_rl = c()
gamma_df_i = c()
for(idx in 61:63){
  gamma_rl = c(gamma_rl, mean(c(rl[[1]][2][[1]][idx], rl[[2]][2][[1]][idx], rl[[3]][2][[1]][idx], rl[[4]][2][[1]][idx])))
  new_idx = idx + (85*3)
  gamma_df_i = c(gamma_df_i, mean(c(rl[[1]][2][[1]][new_idx], rl[[2]][2][[1]][new_idx], rl[[3]][2][[1]][new_idx], rl[[4]][2][[1]][new_idx])))
}

scal_rl = c()
scal_df_i = c()
for(idx in 64:83){
  scal_rl = c(scal_rl, mean(c(rl[[1]][2][[1]][idx], rl[[2]][2][[1]][idx], rl[[3]][2][[1]][idx], rl[[4]][2][[1]][idx])))
  new_idx = idx + (85*3)
  scal_df_i = c(scal_df_i, mean(c(rl[[1]][2][[1]][new_idx], rl[[2]][2][[1]][new_idx], rl[[3]][2][[1]][new_idx], rl[[4]][2][[1]][new_idx])))
}

new_idx =  + (85*3)
sigma_rl = mean(c(rl[[1]][2][[1]][84], rl[[2]][2][[1]][84], rl[[3]][2][[1]][84], rl[[4]][2][[1]][84]))
sigma_df_i = mean(c(rl[[1]][2][[1]][84 + (85 * 3)], rl[[2]][2][[1]][84 + (85 * 3)], rl[[3]][2][[1]][84 + (85 * 3)], rl[[4]][2][[1]][84 + (85 * 3)]))
tau_rl = mean(c(rl[[1]][2][[1]][85], rl[[2]][2][[1]][85], rl[[3]][2][[1]][85], rl[[4]][2][[1]][85]))
tau_df_i = mean(c(rl[[1]][2][[1]][(85 * 4)], rl[[2]][2][[1]][(85 * 4)], rl[[3]][2][[1]][(85 * 4)], rl[[4]][2][[1]][(85 * 4)]))

rl_dt <- data.table(
  x = c(1, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 2, 20, 3, 4, 5, 6, 7, 8, 9),
  type = c("WT", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "WT", "MU", "WT", "WT", "WT", "WT", "WT", "WT", "WT"),
  A_rl = A_rl,
  alpha_rl = alpha_rl,
  B_rl = B_rl,
  scal_rl = scal_rl,
  A_df_i = A_df_i,
  alpha_df_i = alpha_df_i,
  B_df_i = B_df_i,
  scal_df_i = scal_df_i
)
```

```{r}
# Plots for Burn-In RL for Mean of Chains
p1 <- rl_dt |>
  ggplot(aes(x=x, y=A_rl, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "A Burn-In",
    title = "A Burn-In\nfor each Root",
    fill = "Root Type"
  ) 

p2 <- rl_dt |>
  ggplot(aes(x=x, y=B_rl, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "B Burn-In",
    title = "B Burn-In\nfor each Root",
    fill = "Root Type"
  ) 

p3 <- rl_dt |>
  ggplot(aes(x=x, y=alpha_rl, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Alpha Burn-In",
    title = "Alpha Burn-In\nfor each Root",
    fill = "Root Type"
  )  

p4 <- rl_dt |>
  ggplot(aes(x=x, y=scal_rl, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Scale Burn-In",
    title = "Scale Burn-In\nfor each Root",
    fill = "Root Type"
  ) 

(p1 + p2) / (p3 + p4)
```

```{r}
# Gamma, Sigma, Tau for Burn-In 
gamma_rl
sigma_rl
tau_rl
```

#### Dependence Factor

```{r}
# Plots for Dependence Factor RL for Mean of Chains
p1 <- rl_dt |>
  ggplot(aes(x=x, y=A_df_i, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "A Dependence Factor",
    title = "A Dependence Factor\nfor each Root",
    fill = "Root Type"
  ) 

p2 <- rl_dt |>
  ggplot(aes(x=x, y=B_df_i, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "B Dependence Factor",
    title = "B Dependence Factor\nfor each Root",
    fill = "Root Type"
  ) 

p3 <- rl_dt |>
  ggplot(aes(x=x, y=alpha_df_i, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Alpha Dependence Factor",
    title = "Alpha Dependence Factor\nfor each Root",
    fill = "Root Type"
  )  

p4 <- rl_dt |>
  ggplot(aes(x=x, y=scal_df_i, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "Scale Dependence Factor",
    title = "Scale Dependence Factor\nfor each Root",
    fill = "Root Type"
  ) 

(p1 + p2) / (p3 + p4)
```

```{r}
gamma_df_i
sigma_df_i
tau_df_i
```

### ACF Plots (Autocorrelation)

```{r}
pdf("acfPlots.pdf", width=10, height=8)
par(mfrow=c(3, 3), las=1)
autocorr.plot(sim, lag.max=10, auto.layout=FALSE, ask=FALSE)
dev.off()
```

```{r}
autocorr.diag(sim)
```

### posterior summary: CODA Output Analysis Functions

```{r}
post_summary <-summary(sim)
post_summary
```

```{r}
library(DT)
starting_point = 0
means_WT = c()
means_MU = c()

# A, alpha, B
for(i in 0:3){
  s = i*20
  if(s == 3){
    s = s + 11
  }
  means_WT = c(means_WT, mean(ps[s+1], ps[s+12], ps[s+14], ps[s+15], ps[s+16], ps[s+17], ps[s+18], ps[s+19], ps[s+20]))
  means_MU = c(means_MU, mean(ps[s+2], ps[s+3], ps[s+4], ps[s+5], ps[s+6], ps[s+7], ps[s+8], ps[s+9], ps[s+10], ps[s+11], ps[s+13]))
}

ps_df <- data.table(
  type = c("WT", "MU"),
  A_mean = c(means_WT[1], means_MU[1]), 
  alpha_mean = c(means_WT[2], means_MU[2]),
  B_mean = c(means_WT[3], means_MU[3]),
  scal_mean = c(means_WT[4], means_MU[4])
)

kable(ps_df)
```

### Frequentist vs. Bayesian Model Comparison

```{r}
root_levels <- levels(as.factor(rootData$root))

# 2) get posterior means from JAGS
post_stats <- post_summary$statistics
post_mean <- tibble(
  param = rownames(post_stats),
  mean  = post_stats[, "Mean"]
)

# 3) keep only root-level params and pull out index
bayes_roots <- post_mean %>%
  filter(str_detect(param, "^(A|B|alpha|scal)\\[")) %>%
  mutate(
    # extract the number inside brackets, e.g. "A[12]" -> "12"
    idx  = as.integer(str_extract(param, "\\d+")),
    # map numeric index back to actual root label from the data
    root_label = root_levels[idx],             # "WT1", "MU3", ...
    root = paste0("root_", root_label),        # "root_WT1", "root_MU3"
    param_name = str_extract(param, "^[A-Za-z]+")
  ) %>%
  select(root, param_name, mean) %>%
  # wide so we get columns A, B, alpha, scal
  pivot_wider(names_from = param_name, values_from = mean)
```

```{r}
compare_results <- all_results |>
  left_join(bayes_roots, by = "root", suffix = c("_nls", "_bayes"))
```

```{r}
# scatter plots
compare_results |>
  ggplot(aes(x = A_nls, y = A_bayes)) +
  geom_point(color = "#1b9e77") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(x = "Frequentist (NLS) A", y = "Bayesian Posterior Mean A",
       title = "Comparison of Lower Asymptote (A)") +
  theme_minimal()

compare_results |>
  ggplot(aes(x = B_nls, y = B_bayes)) +
  geom_point(color = "#1b9e77") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(x = "Frequentist (NLS) ", y = "Bayesian Posterior Mean B",
       title = "Comparison of Upper Asymptote (B)") +
  theme_minimal()


```

```{r}
# difference plot
compare_results |>
  mutate(A_diff = A_bayes - A_nls) |>
  ggplot(aes(x = root, y = A_diff)) +
  geom_col(fill = "#7570b3") +
  labs(y = "Difference (Bayes – NLS)", title = "Shrinkage in A across roots") +
  theme_minimal()
```

# Predicted Growth Curves by Genotype × Side
Plot the fitted (Frequentist) and posterior mean (Bayesian) logistic curves on the same axes for Inner vs Outer sides, separated by genotype
```{r}
library(purrr)
get_freq_preds <- function(model_list, data, geno_label) {
  map2_dfr(
    .x = model_list,
    .y = names(model_list),
    ~ {
      root_id <- str_replace(.y, "root_", "")
      d <- data %>% filter(root == root_id)
      d %>% 
        mutate(Freq_Pred = fitted(.x),
               Genotype = geno_label)
    }
  )
}

freq_preds <- bind_rows(
  get_freq_preds(WT_model_output, rootData_WT, "WT"),
  get_freq_preds(MU_model_output, rootData_MU, "MU")
)

pm <- jags.out$BUGSoutput$mean

# 1) root index in same order as JAGS
root_idx <- as.numeric(as.factor(rootData$root))

# 2) numeric side / genotype to match model
side_num <- ifelse(rootData$side == "Outer", 1, 0)
geno_num <- ifelse(rootData$Genotype == "WT", 0, 1)  # adjust if you coded differently

# 3) pull root-level params from the list
A_i     <- pm$A[root_idx]
B_i     <- pm$B[root_idx]
alpha_i <- pm$alpha[root_idx]
scal_i  <- pm$scal[root_idx]

# 4) build effective midpoint using scalars from the list
xmid_eff <- alpha_i +
  pm$gamma1 * side_num +
  pm$gamma2 * geno_num +
  pm$gamma3 * side_num * geno_num

# 5) rebuild mu exactly like in JAGS
Bayes_Pred <- A_i + (B_i - A_i) /
  (1 + exp((xmid_eff - rootData$x01) / scal_i))

# 6) attach to your data
bayes_preds <- dplyr::mutate(rootData, Bayes_Pred = Bayes_Pred)
```

```{r}
plot_data <- rootData %>%
  left_join(freq_preds %>% select(root, Midline, side, Genotype, Freq_Pred),
            by = c("root", "Midline", "side", "Genotype")) %>%
  left_join(bayes_preds %>% select(Midline, Bayes_Pred),
            by = "Midline")

plot_long <- plot_data %>%
  select(x01, Length, side, Genotype, Freq_Pred, Bayes_Pred) %>%
  pivot_longer(cols = c(Freq_Pred, Bayes_Pred),
               names_to = "Model",
               values_to = "Pred") %>%
  mutate(Model = recode(Model,
                        "Freq_Pred" = "Frequentist",
                        "Bayes_Pred" = "Bayesian"))

plot_long <- plot_long %>%
  mutate(Model = factor(Model, levels = c("Frequentist", "Bayesian")))
```
```{r}

p_MU <- plot_long %>%
  filter(Genotype == "MU") %>%
  ggplot(aes(x = x01, color = side)) +
  stat_summary(aes(y = Length), fun = mean, geom = "point", alpha = 0.3, size = 0.4) +
  stat_summary(aes(y = Pred), fun = mean, geom = "line", size = 1) +
  facet_wrap(~ Model, ncol = 2, scales = "free_y") +
  labs(
    title = "Predicted Growth Curves – Mutant (MU)",
    subtitle = "Left: Frequentist  |  Right: Bayesian",
    x = "Normalized Position (x01)",
    y = "Mean Cell Length (µm)",
    color = "Side"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    panel.spacing = unit(1.5, "lines")
  )

p_WT <- plot_long %>%
  filter(Genotype == "WT") %>%
  ggplot(aes(x = x01, color = side)) +
  stat_summary(aes(y = Length), fun = mean, geom = "point", alpha = 0.3, size = 0.4) +
  stat_summary(aes(y = Pred), fun = mean, geom = "line", size = 1) +
  facet_wrap(~ Model, ncol = 2, scales = "free_y") +
  labs(
    title = "Predicted Growth Curves – Wild Type (WT)",
    subtitle = "Left: Frequentist  |  Right: Bayesian",
    x = "Normalized Position (x01)",
    y = "Mean Cell Length (µm)",
    color = "Side"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    panel.spacing = unit(1.5, "lines")
  )
```

```{r}
p_MU
p_WT
```
# residual plot
```{r}
bayes_res <- rootData %>%
  mutate(
    Fitted = Bayes_Pred,
    Residuals = Length - Fitted
  )

WT_residuals_bayes <- bayes_res %>% filter(Genotype == "WT")
MU_residuals_bayes <- bayes_res %>% filter(Genotype == "MU")

p_wt_fit <- ggplot(WT_residuals_bayes,
                   aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (WT)",
       x = "Fitted Cell Length",
       y = "Residuals") +
  theme_minimal()

p_wt_mid <- ggplot(WT_residuals_bayes,
                   aes(x = Midline, y = Residuals, color = side)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (WT)",
       x = "Distance Along Midline",
       y = "Residuals") +
  theme_minimal()

# MU residual plots
p_mu_fit <- ggplot(MU_residuals_bayes,
                   aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (MU)",
       x = "Fitted Cell Length",
       y = "Residuals") +
  theme_minimal()

p_mu_mid <- ggplot(MU_residuals_bayes,
                   aes(x = Midline, y = Residuals, color = side)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (MU)",
       x = "Distance Along Midline",
       y = "Residuals") +
  theme_minimal()

library(patchwork)

(p_wt_fit + p_wt_mid) /
(p_mu_fit + p_mu_mid) 

```


