---
title: "Sta 440 Case 4"
author: "Olivia Fu, Yura Heo, Jack Steel"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format: 
  pdf:
    documentclass: article
    fontsize: 11pt
geometry: margin=1in
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
load("./rootData.RData")

library(ggplot2)
library(patchwork)
library(dplyr)
library(nlme)
library(broom)
library(knitr)
library(stringr)
library(tidyverse)
library(purrr)
library(kableExtra)
library(tidyr)
```

# Background

Root circumnutation, the helical or oscillatory growth movement of roots, has been hypothesized to result from differential cell elongation on opposing sides of the root tip. This asymmetry helps roots navigate soil, bypass obstacles, and stabilize themselves. To investigate this mechanism, we compare spatial growth patterns in wild-type and mutant genotypes that differ in circumnutation behavior. Our analysis models elongation along the midline for inner and outer sides to test whether asymmetric growth drives circumnutation and whether this asymmetry is diminished in mutants.

This analysis aims to quantify the extent of asymmetry in cell length along the inner and outer curves of individual roots and to assess how these patterns differ between genotypes. By fitting both frequentist and hierarchical Bayesian nonlinear regression models, we seek to: (1) identify within-root growth differences linked to circumnutation, (2) characterize variability across roots, and (3) evaluate whether the mutant genotype exhibits reduced differential growth consistent with its impaired movement phenotype.

# Exploratory Data Analysis

```{r}
plot1 <- ggplot(rootData, aes(x = side, y = Length, fill = side)) +
  geom_boxplot() +
  facet_wrap(~ Genotype) +
  labs(title = "Cell Length Distribution by Inner vs Outer",
       y = "Cell Length") +
  theme_bw()
```

The roots are classified as mutant or wild type, and their cell lengths on the inner side and outer side of the roots are measured at various distances along the midline. The distribution for the cell length is right skewed for both sides and both root types, with the mutant type being slightly stronger right skewed (@fig-cell-length-dist). The inner and outer side for the mutant type have a nearly identical distribution, with a median cell length around 9. On the other hand, the distributions for the cell length for the wild type differ between the inner measurements and outer measurements, where the outer cell length measurements are slightly larger than the inner measurements.

```{r}
rootData_WT <- rootData |> filter(Genotype == "WT")

plot2 <- ggplot(rootData_WT, aes(x = Midline, y = Length, 
                        color = side, group = interaction(root, side))) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "Cell Length Along Midline – Wild Type Roots",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

rootData_MU <- rootData |> filter(Genotype == "MU")
desired_order <- c("MU1", "MU2", "MU3", "MU4", "MU5", "MU6", "MU7", "MU8", "MU9", "MU10", "MU11")
rootData_MU <- rootData_MU |>
  mutate(root = factor(root, levels = desired_order))

plot3 <- ggplot(rootData_MU, aes(x = Midline, y = Length, 
                        color = side, group = interaction(root, side))) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "Cell Length Along Midline – Mutant Roots",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The wild type roots (@fig-cell-length-per-root-wt, @fig-cell-length-per-root-smooth-wt) show distinct inner and outer measurement trends, while the mutant roots (@fig-cell-length-per-root-mu, @fig-cell-length-per-root-smooth-mu) show nearly identical trends. Wild type roots one, three, and eight particularly have higher inner measurements than outer measurements. This is further emphasized by the smooth line plots in @fig-cell-length-smooth-wt and @fig-cell-length-smooth-mu - the wild type has noticeably different inner and outer measurements for cell length.

```{r}
plot4 <- ggplot(rootData_WT,
       aes(x = Midline, y = Length, color = side, group = interaction(root, side))) +
  geom_smooth(method = "loess", se = FALSE, span = 0.75, size = 0.8) +
  geom_point(alpha = 0.4, size = 0.5) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "WT: Inner vs Outer Smoothed Trends for Each Root",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot5 <- ggplot(rootData_MU,
       aes(x = Midline, y = Length, color = side, group = interaction(root, side))) +
  geom_smooth(method = "loess", se = FALSE, span = 0.75, size = 0.8) +
  geom_point(alpha = 0.4, size = 0.5) +
  facet_wrap(~ root, scales = "free_y") +
  labs(
    title = "MT: Inner vs Outer Smoothed Trends for Each Root",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
plot6 <- ggplot(rootData_WT, aes(x = Midline, y = Length, color = side)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.8, size = 1.2) +
  labs(
    title = "WT: Inner vs Outer Overall Cell Length Trends",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw()

plot7 <- ggplot(rootData_MU, aes(x = Midline, y = Length, color = side)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.8, size = 1.2) +
  labs(
    title = "MU: Inner vs Outer Overall Cell Length Trends",
    x = "Distance Along Midline",
    y = "Cell Length",
    color = "Side"
  ) +
  theme_bw()
```

```{r}
plot10 <- rootData_WT |>
  ggplot(aes(x=root, y=Length, fill=side)) +
  geom_bar(stat="summary", fun="mean", position="dodge") +
  labs(
    x = "Root", 
    y = "Mean Cell Length",
    title = "Mean Cell Length for Each Wild Type Root"
  ) +
  theme_minimal()

plot11 <- rootData_MU |>
  ggplot(aes(x=root, y=Length, fill=side)) +
  geom_bar(stat="summary", fun="mean", position="dodge") +
  labs(
    x = "Root", 
    y = "Mean Cell Length",
    title = "Mean Cell Length for Each Mutant Type Root"
  ) +
  theme_minimal()
```

For the mutant roots, mean cell lengths are similar between the inner and outer sides, except for root nine, where the mean inner length is slightly larger (@fig-mean-cell-length-per-root-mu). In contrast, the wild type roots generally show larger mean outer cell lengths, except for root seven, where the inner mean is slightly higher (@fig-mean-cell-length-per-root-wt).

# Model Rationale and Implementation

## Frequentist Model

Our EDA showed that cell length increases nonlinearly from the root tip toward the base, consistent with a sigmoidal growth pattern. These findings motivated a restricted logistic model structure that captures both the saturating nature of growth and the side-specific shift in elongation patterns through the inclusion of a side indicator term ($\beta$) to quantify asymmetry in elongation between the inner and outer sides. Because the data were collected separately for each side of the root and for two genotypes, this formulation directly reflects the experimental design and allows side- and genotype-level comparisons. The logistic form also aligns with biological constraints: cell lengths cannot be negative and are expected to plateau as elongation slows near the root base. $$
\text{Length} \sim A + \frac{(B - A)}{1 + \exp\left(\frac{x_{\text{mid}} + \beta \cdot \text{side} - x}{\text{scal}}\right)}
$$Here, A and B are lower and upper asymptotes of cell length, $x_{mid}$ is the inflection point, $scal$ controls curve steepness, and $\beta$ shifts the midpoint between inner (0) and outer (1) sides, indicating differential growth.

Using this setup, nonlinear least squares (NLS) models were fit separately for each root using a custom logistic function. Parameter bounds were defined to ensure biologically realistic values, and starting points were initialized from each root’s observed cell length distribution to aid convergence. In total, 9 wild type and 11 mutant roots were modeled independently to evaluate within-root and between-genotype differences in elongation patterns.

```{r}
fpl_customized <- function(x, A, B, xmid, scal, beta, side_ind){
  xmid_eff <- xmid + beta * side_ind
  A + (B - A) / (1 + exp((xmid_eff - x) / scal))
}
```

```{r}
WT_model_output <- list()

for (root_num in unique(rootData_WT$root)) {
  curr_root <- rootData_WT |> filter(root == root_num)
  curr_root$side_ind <- ifelse(curr_root$side == "Outer", 1, 0)

  y_min <- min(curr_root$Length)
  y_max <- max(curr_root$Length)

  start_vals <- list(
    A = y_min * 0.9,
    B = y_max * 1.1,
    xmid = median(curr_root$Midline),
    scal = diff(range(curr_root$Midline)) / 4,
    beta = -50
  )

  lower_bounds <- c(A = y_min * 0.5,
                    B = y_min,
                    xmid = min(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)) / 10,
                    beta = -500)

  upper_bounds <- c(A = y_max,
                    B = y_max * 2.0,
                    xmid = max(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)),
                    beta = 500)

  fit <- try(
    nls(
      Length ~ fpl_customized(Midline, A, B, xmid, scal, beta, side_ind),
      data = curr_root,
      start = start_vals,
      algorithm = "port",
      lower = lower_bounds,
      upper = upper_bounds,
      control = nls.control(maxiter = 1000, warnOnly = TRUE, minFactor = 1e-10)
    ),
    silent = TRUE
  )

  if (!inherits(fit, "try-error")) {
    WT_model_output[[paste0("root_", root_num)]] <- fit
  } else {
    message(paste("Model failed for root", root_num))
  }
}
```

```{r}
MU_model_output <- list()

for (root_num in unique(rootData_MU$root)) {
  curr_root <- rootData_MU |> filter(root == root_num)
  curr_root$side_ind <- ifelse(curr_root$side == "Outer", 1, 0)

  y_min <- min(curr_root$Length)
  y_max <- max(curr_root$Length)

  start_vals <- list(
    A = y_min * 0.9,
    B = y_max * 1.1,
    xmid = median(curr_root$Midline),
    scal = diff(range(curr_root$Midline)) / 4,
    beta = 0
  )

  lower_bounds <- c(A = y_min * 0.5,
                    B = y_min,
                    xmid = min(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)) / 10,
                    beta = -500)

  upper_bounds <- c(A = y_max,
                    B = y_max * 2.0,
                    xmid = max(curr_root$Midline),
                    scal = diff(range(curr_root$Midline)),
                    beta = 500)

  fit <- try(
    nls(
      Length ~ fpl_customized(Midline, A, B, xmid, scal, beta, side_ind),
      data = curr_root,
      start = start_vals,
      algorithm = "port",
      lower = lower_bounds,
      upper = upper_bounds,
      control = nls.control(maxiter = 1000, warnOnly = TRUE, minFactor = 1e-10)
    ),
    silent = TRUE
  )

  if (!inherits(fit, "try-error")) {
    MU_model_output[[paste0("root_", root_num)]] <- fit
  } else {
    message(paste("Model failed for root", root_num))
  }
}
```

```{r}
WT_residuals <- data.frame()

for (name in names(WT_model_output)) {
  model <- WT_model_output[[name]]
  root_label <- str_replace(name, "root_", "")
    curr_data <- rootData_WT |> filter(root == root_label)
  
  curr_data <- curr_data |>
    mutate(
      Fitted = fitted(model),
      Residuals = resid(model)
    )
  
  WT_residuals <- bind_rows(WT_residuals, curr_data)
}

plot12 <- ggplot(WT_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(size = 0.6, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (WT)", x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

plot13 <- ggplot(WT_residuals, aes(x = Midline, y = Residuals, color = side)) +
  geom_point(size = 0.6, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (WT)", x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()
```

```{r}
MU_residuals <- data.frame()

for (name in names(MU_model_output)) {
  model <- MU_model_output[[name]]
  root_label <- str_replace(name, "root_", "")
    curr_data <- rootData_MU |> filter(root == root_label)
  
  curr_data <- curr_data |>
    mutate(
      Fitted = fitted(model),
      Residuals = resid(model)
    )
  
  MU_residuals <- bind_rows(MU_residuals, curr_data)
}

plot15 <- ggplot(MU_residuals, aes(x = Fitted, y = Residuals, color = side)) +
  geom_point(size = 0.6, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (MU)", x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

plot16 <- ggplot(MU_residuals, aes(x = Midline, y = Residuals, color = side)) +
  geom_point(size = 0.6, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (MU)", x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()
```

```{r}
all_residuals <- bind_rows(WT_residuals, MU_residuals)

# Ensure proper root order: WT1..WT9, MU1..MU11
all_residuals <- all_residuals %>%
  mutate(root = factor(root, 
                       levels = c(paste0("WT", 1:9), paste0("MU", 1:11))))

# Plot combined residual boxplot
p_all_box <- ggplot(all_residuals, aes(x = root, y = Residuals, fill = side)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(title = "Residuals by Root (Frequentist Model)",
       x = "Root ID", y = "ResidualS", fill = "Side") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )
```

```{r}
# ACF plot (autocorr check)
m0 <- lm(Length ~ side * poly(Midline, 2), data = rootData)
```

For model evaluation, the residuals versus fitted plots (@fig-resid-freq) for both wild type and mutant roots show a slight fanning pattern, indicating minor heteroscedasticity and a mild violation of the constant variance assumption. The residuals versus midline plots (@fig-resid-freq) appear evenly scattered along the midline for both genotypes. Examining residuals by root (@fig-resid-by-root) shows that most are centered around zero with no outlying roots. These results indicate that the frequentist model provided a good fit for both the mutant and wild type root data. However, the ACF plot (@fig-freq-acf) suggests that there is autocorrelation that our model might not necessarily account for and that our model may have not captured the patterns in the data, as the bars exceed the bounds.

## Bayesian Model

### Model Rationale

We modeled cell length as normally distributed around a logistic growth curve, with root-specific random effects and covariate effects for genotype and side:

$$
\text{Length}_i \sim \mathcal{N}(\mu_i, \sigma^2),
$$

where $\mu_i$ represents the expected cell length for observation $i$. The distance along the root midline was normalized to $[0,1]$ to ensure comparability across roots of different lengths.

Guided by our EDA and frequentist results, we modeled the mean cell length with a four-parameter logistic curve, where for each root $r$, $A_r$ denotes the lower asymptote, $B_r$ the upper asymptote, $x_{\text{mid}}$ the effective inflection point, and $\text{scal}_r$ the steepness parameter:

$$
\mu_i = A_{r[i]} + \frac{B_{r[i]} - A_{r[i]}}{1 + \exp\left(\frac{x_{\text{mid},i} - x_{01,i}}{\text{scal}_{r[i]}}\right)}.
$$

To capture differences between wildtype vs mutant and inner vs outer root sides, we modeled the midpoint as a function of side and genotype effects, where $\alpha_r$ represents the baseline midpoint for wildtype inner and mutant inner sides, $\gamma_1$ captures the shift between outer and inner sides in wildtype roots, and $\gamma_2$ captures the additional shift for mutant outer roots. This structure allows side and genotype effects to interact, enabling direct evaluation of the difference between inner and outer sides in wildtype roots and whether this difference diminishes in mutant roots:

$$
x_{\text{mid},i} = \alpha_{r[i]} + \gamma_1 \cdot \text{side}_i + \gamma_2 \cdot \text{side}_i \cdot \text{geno}_i.
$$

To represent uncertainty about side and genotype effects, we placed normal priors centered at zero, $\gamma_1, \gamma_2 \sim \mathcal{N}(0, 0.5^2)$, corresponding to a standard deviation of 0.5 on the normalized $x$-axis. Residual variation was modeled via a half-$t$ prior, $\sigma \sim \text{Half-}t(0, 3, \nu = 3)$, which provides heavier tails than a uniform prior – allowing large residuals if supported but concentrating most mass around moderate values.

To account for biological variability in the four logistic parameters, we included root-level random effects, $a_j \sim \mathcal{N}(\mu_a, \sigma_a^2)$, $d_j \sim \mathcal{N}(\mu_d, \sigma_d^2)$, $z_{0,j} \sim \mathcal{N}(\mu_{z0}, \sigma_{z0}^2)$, and $\eta_{s,j} \sim \mathcal{N}(\mu_{\eta_s}, \sigma_{\eta_s}^2)$. These were transformed to ensure biologically meaningful constraints: $A_j = e^{a_j} > 0$, $B_j = A_j + e^{d_j} > A_j$, $\alpha_j = \text{logit}^{-1}(z_{0,j}) \in (0,1)$, and $\text{scal}_j = e^{\eta_{s,j}} > 0$.

We used weakly informative, data-anchored hyperpriors to guide estimation: $\mu_a \sim \mathcal{N}(\log L_{\text{lower}}, 1)$, $\mu_d \sim \mathcal{N}(\log L_{\text{span}}, 1)$, $\mu_{z0} \sim \mathcal{N}(0, 1)$, and $\mu_{\eta_s} \sim \mathcal{N}(\log 0.2, 0.5)$, where $L_{\text{lower}}$ and $L_{\text{span}}$ are empirical lower and span length scales from the data. To allow moderate between-root variability, we set the precisions following $\text{Gamma}(1, 1)$ priors.

### Model Implementation

We initialized the Bayesian model using estimates from the frequentist nonlinear least squares (NLS) fits. For each root, the NLS point estimates and standard errors for the four logistic parameters ($A$, $B$, $x_{\text{mid}}$, and $\text{scal}$) were used to generate dispersed starting values. Specifically, for each chain, we drew initial values from Normal distributions centered at the NLS estimates with standard deviations set to five times the corresponding NLS standard errors. This overdispersion ensured that the chains started from diverse points in the parameter space rather than being overly constrained by the frequentist results. The starting values were then transformed to the latent parameters ($a$, $d$, $z_0$, $\eta_s$) as described previously. The fixed effects ($\gamma_1$, $\gamma_2$) were initialized near zero with small random noise ($\text{sd} = 0.01$) to maintain variability without imposing strong directional bias. The residual scale parameter ($t_\sigma$) was initialized from a $\text{Normal}(0, 3)$ distribution, consistent with its half-$t$ prior specification. We ran four MCMC chains with 50,000 iterations each, discarding the first 10,000 as burn-in and using no thinning.

```{r}
library(rjags)
library(R2jags)
library(coda)
```

```{r}
# Normalize Midline to the [0,1] range
rootData <- rootData |>
  group_by(root) |>
  mutate(x01 = (Midline - min(Midline)) / (max(Midline) - min(Midline))) |>
  ungroup()
```

```{r}
L_lower <- min(rootData$Length, na.rm = TRUE)
L_upper <- max(rootData$Length, na.rm = TRUE)
L_span  <- L_upper - L_lower

bdat <- list(
  Length = rootData$Length,
  N = nrow(rootData),
  x01 = rootData$x01,
  side = as.numeric(rootData$side == "Outer"),
  geno = as.numeric(rootData$Genotype == "MU"),
  root = as.numeric(as.factor(rootData$root)),
  n.root = length(unique(rootData$root)),
  L_lower = L_lower,
  L_span  = L_span
)

fun.params <- c("A", "B", "alpha", "scal",
                "gamma1", "gamma2",
                "sigma",
                "mu_a", "mu_d", "mu_z0", "mu_eta_s",
                "prec_a", "prec_d", "prec_z0", "prec_eta_s")
```

```{r}
get_nls_summary <- function(model) {
  s <- summary(model)
  est <- s$coefficients[, 1]
  se  <- s$coefficients[, 2]
  tibble(A = est["A"], A_se = se["A"],
         B = est["B"], B_se = se["B"],
         scal = est["scal"], scal_se = se["scal"],
         xmid = est["xmid"], xmid_se = se["xmid"])
}

WT_results_full <- map_dfr(WT_model_output, get_nls_summary, .id = "root")
MU_results_full <- map_dfr(MU_model_output, get_nls_summary, .id = "root")
all_results <- bind_rows(WT_results_full, MU_results_full)
```

```{r}
root_levels <- levels(as.factor(rootData$root))
nroot <- length(root_levels)
A_init <- B_init <- scal_init <- alpha_init <- numeric(nroot)
A_sd <- B_sd <- scal_sd <- alpha_sd <- numeric(nroot)

for (i in seq_along(root_levels)) {
  this_name <- paste0("root_", root_levels[i])
  row_i <- all_results[all_results$root == this_name, ]
  rd_i <- filter(rootData, root == root_levels[i])

  # we have fits for every root
  A_init[i] <- row_i$A
  A_sd[i]   <- row_i$A_se
  B_init[i] <- row_i$B
  B_sd[i]   <- row_i$B_se

  # Get min and max for this root
  mid_min <- min(rd_i$Midline)
  mid_max <- max(rd_i$Midline)
  span    <- mid_max - mid_min

  # Normalize both xmid and scal
  alpha_init[i] <- (row_i$xmid - mid_min) / span
  alpha_sd[i]   <- row_i$xmid_se / span
  scal_init[i]  <- row_i$scal / span
  scal_sd[i]    <- row_i$scal_se / span
}
```

```{r}
fun.inits <- function(chain_id = 1) {
  t_sigma.start <- rnorm(1, mean = 0, sd = 3)

  # Overdisperse latent parameters around NLS estimates
  a.start     <- log(pmax(1e-3, rnorm(bdat$n.root, mean = A_init, sd = 5 * A_sd)))
  d.start     <- log(pmax(1e-3, rnorm(bdat$n.root, mean = B_init - A_init, sd = 5 * B_sd)))
  z0.start    <- qlogis(
    pmin(0.99, pmax(0.01, rnorm(bdat$n.root, mean = alpha_init, sd = 5 * alpha_sd))))
  eta_s.start <- log(pmax(1e-3, rnorm(bdat$n.root, mean = scal_init, sd = 5 * scal_sd)))

  # Fixed effects (gamma terms) — random small perturbations
  gamma1.start <- rnorm(1, mean = 0, sd = 0.01)
  gamma2.start <- rnorm(1, mean = 0, sd = 0.01)

  return(list(
    a = a.start,
    d = d.start,
    z0 = z0.start,
    eta_s = eta_s.start,
    gamma1 = gamma1.start,
    gamma2 = gamma2.start,
    t_sigma = t_sigma.start
  ))
}

init_list <- list(
  fun.inits(1),
  fun.inits(2),
  fun.inits(3),
  fun.inits(4)
)
```

```{r}
# fun.model.file <- "model_root.R" 
# mcmc.pars <- list(iter = 50000, burn = 10000, nchain=4, thin=1)
# 
# jags.out <- jags(
#   data = bdat,
#   inits = init_list,
#   parameters.to.save = fun.params,
#   model.file = fun.model.file,
#   n.chains = mcmc.pars$nchain,
#   n.iter = mcmc.pars$iter,
#   n.burnin = mcmc.pars$burn,
#   n.thin = mcmc.pars$thin,
#   DIC = FALSE,
#   progress.bar = "none"
# )
```

```{r}
# load the result instead of running jags everytime
load("./bayes_model_run_final.RData")
```

### Model Evaluation

```{r}
sim <- as.mcmc(jags.out)
```

```{r}
gelman_results <- gelman.diag(sim)
psrf_df <- as.data.frame(gelman_results$psrf)
psrf_df$Parameter <- rownames(psrf_df)
colnames(psrf_df) <- c("Point_Est", "Upper_CI", "Parameter")
```

```{r}
psrf_top20 <- psrf_df |>
  arrange(desc(Point_Est)) |>
  slice(1:20) |>
  mutate(Flag = ifelse(Point_Est > 1.01, "Above 1.01", "Below 1.01"))

# Plot
plot19 <-  ggplot(psrf_top20, aes(x = Point_Est, y = reorder(Parameter, Point_Est))) +
  geom_point(size = 1.8, color = "steelblue") +
  geom_errorbarh(aes(xmin = Point_Est, xmax = Upper_CI),
                 height = 0.1, color = "gray50") +
  geom_vline(xintercept = 1, linetype = "dotted", color = "gray40") +
  geom_vline(xintercept = 1.01, linetype = "dashed", color = "orange") +
  geom_vline(xintercept = 1.05, linetype = "dashed", color = "red") +
  coord_cartesian(xlim = c(0.999, 1.06)) +
  scale_color_manual(values = c("Below 1.01" = "steelblue",
                                "Above 1.01" = "firebrick")) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 7),
    panel.grid.major.y = element_blank()
  ) +
  labs(
    y = "Parameters (Top 20 by R-hat)",
    x = expression(hat(R)),
    title = "Gelman–Rubin Diagnostic (R-hat)",
    subtitle = "Dashed lines at 1.05 indicate potential non-convergence"
  )
```

```{r}
library(data.table)
ess <- effectiveSize(sim)

A_vals     <- ess[grep("^A\\[", names(ess))]
alpha_vals <- ess[grep("^alpha\\[", names(ess))]
B_vals     <- ess[grep("^B\\[", names(ess))]
scal_vals  <- ess[grep("^scal\\[", names(ess))]
gamma_vals <- ess[grep("^gamma", names(ess))]

sigma       <- ess["sigma"]
mu_a        <- ess["mu_a"]
mu_d        <- ess["mu_d"]
mu_eta_s    <- ess["mu_eta_s"]
mu_z0       <- ess["mu_z0"]
prec_a      <- ess["prec_a"]
prec_d      <- ess["prec_d"]
prec_eta_s  <- ess["prec_eta_s"]
prec_z0     <- ess["prec_z0"]

effective_dt <- data.table(
  x = c(1, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 2, 20, 3, 4, 5, 6, 7, 8, 9),
  type = c("WT", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU", "MU",
           "MU", "WT", "MU", "WT", "WT", "WT", "WT", "WT", "WT", "WT"),
  A = as.numeric(A_vals),
  alpha = as.numeric(alpha_vals),
  B = as.numeric(B_vals),
  scal = as.numeric(scal_vals)
)

hyper_dt <- data.table(
  parameter = c(
    "gamma1 (Outer effect)",
    "gamma2 (Interaction)",
    "sigma (Residual SD)",
    "mu_a",
    "mu_d",
    "mu_eta_s",
    "mu_z0",
    "prec_a",
    "prec_d",
    "prec_eta_s",
    "prec_z0"
  ),
  ESS = as.numeric(c(
    gamma_vals,
    sigma,
    mu_a, mu_d, mu_eta_s, mu_z0,
    prec_a, prec_d, prec_eta_s, prec_z0
  ))
)


hyper_dt[, category := fifelse(grepl("^gamma", parameter), "Fixed effects",
                        fifelse(grepl("^mu_", parameter), "Hyper means",
                        fifelse(grepl("^prec_", parameter), "Precisions",
                        "Residual variance")))]

setorder(hyper_dt, category)


table6 <- kable(hyper_dt, digits = 1, caption = "Effective Sample Size (ESS) for Hyperparameters")
```

```{r}
library(patchwork)

p1 <- ggplot(data=effective_dt, aes(x=x, y=A, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "ESS",
    title = "A Effective Sample Size",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 8000)) +
  annotate("label", x=7, y=8000, label=as.character(as.integer(max(effective_dt$A))), size=2) +
  theme_minimal()

p2 <- effective_dt |>
  ggplot(aes(x=x, y=B, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "ESS",
    title = "B Effective Sample Size",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 8000)) +
  annotate("label", x=7, y=8000, label=as.character(as.integer(max(effective_dt$B))), size=2) +
  theme_minimal()

p3 <- effective_dt |>
  ggplot(aes(x=x, y=alpha, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "ESS",
    title = "Alpha Effective Sample Size",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 8000)) +
  annotate("label", x=7, y=8000, label=as.character(as.integer(max(effective_dt$alpha))), size=2) +
  theme_minimal()

p4 <- effective_dt |>
  ggplot(aes(x=x, y=scal, fill=type)) +
  geom_bar(stat="identity") +
  labs(
    x = "Root Number",
    y = "ESS",
    title = "Scale Effective Sample Size",
    fill = "Root Type"
  ) +
  coord_cartesian(ylim = c(0, 8000)) +
  annotate("label", x=7, y=8000, label=as.character(as.integer(max(effective_dt$scal))), size=2) +
  theme_minimal()

plot20 <- (p1 + p2) / (p3 + p4)
```

```{r}
rl <- raftery.diag(sim)
```

```{r}
# Combine the resmatrix data for all chains
df_list <- lapply(rl, function(x) as.data.frame(x$resmatrix))

# Get common parameter names
common_params <- Reduce(intersect, lapply(df_list, rownames))

# Average across chains
df_mean <- Reduce("+", lapply(df_list, function(df) df[common_params, ])) / length(df_list)

# Add parameter names
df_mean$Parameter <- common_params

# Clean column names
colnames(df_mean) <- c("Burn_in_M", "Total_N", "Lower_bound_Nmin", "Dependence_I", "Parameter")

# Drop Lower_bound_Nmin
df_mean <- subset(df_mean, select = -Lower_bound_Nmin)

# Sort and select top 10 by Dependence_I
df_top10 <- df_mean[order(-df_mean$Dependence_I), ][1:10, ]

# Nicely formatted table
table5 <- kable(df_top10, digits = 2, align = "c",
      caption = "Top 10 Parameters by Mean Dependence Factor (Averaged Across Chains)")
```

The Gelman-Rubin Plot (@fig-gelman-rubin) indicates that all of our parameters converged with 95% confidence, as each parameter has an R-hat upper confidence interval boundary less than 1.03, which is less than potential non-convergence boundary of 1.05. The traceplots (@fig-trace-plots) also point towards convergence for each of the hyperparameters in our model. Each traceplot appears to have a “fuzzy caterpillar” shape and seem to oscillate around a value that the model converged on for the duration of the iterations.

The effective sample sizes for parameters A, B, Alpha, and Scale (@fig-ess) are high for both wild and mutant roots, with root seven showing particularly large values. Overall, the ESS values for wild type roots are slightly higher than those for mutant roots. The hyperparameters also exhibit high ESS (Table 6), especially for $\gamma_1$ (outer effect) and $\gamma_2$ (genotype–side interaction).

In the Raftery-Lewis diagnostic table (Table 5), the parameters with the top ten burn-in and dependence factors are displayed. The dependence factor is highest for various wild and mutant type A and B parameters, indicating that high autocorrelation in the samples increases the sample size needed to converge. This is correlated with the burn in value, as a high burn means there must be a large number of discarded initial iterations before the chain reached stationarity.

The residuals versus fitted and residuals versus midline plots (@fig-resid-bayes) show a random scatter of points with only slight fanning for both wild and mutant roots, indicating that the Bayesian model fits the data well and that variance is approximately constant. Residuals for the wild type roots exhibit slightly greater variability than those for the mutant roots (@fig-resid-by-root-bayes), and most are centered around zero, except for the inner sides of WT7, MU9, and MU10.

```{r}
posterior <- as.data.frame(do.call(rbind, sim))
```

```{r}
# Number of draws for prior
n_samp <- nrow(posterior)

priors <- data.frame(
  gamma1 = rnorm(n_samp, 0, 1/sqrt(4)),
  gamma2 = rnorm(n_samp, 0, 1/sqrt(4)),
  mu_a = rnorm(n_samp, log(L_lower), 1),
  mu_d = rnorm(n_samp, log(L_span), 1),
  mu_z0 = rnorm(n_samp, 0, 1),
  mu_eta_s = rnorm(n_samp, log(0.2), 1/sqrt(2)),
  sigma = abs(rt(n_samp, df = 3)) * 3,
  prec_a = rgamma(n_samp, shape = 1, rate = 1),
  prec_d = rgamma(n_samp, shape = 1, rate = 1),
  prec_z0 = rgamma(n_samp, shape = 1, rate = 1),
  prec_eta_s = rgamma(n_samp, shape = 1, rate = 1)
)
```

```{r}
params_to_plot <- c("gamma1", "gamma2", 
                    "mu_a", "mu_d", "mu_z0","mu_eta_s", 
                    "prec_a", "prec_d", "prec_z0", "prec_eta_s", "sigma")

# Make sure we only use parameters actually present
params_to_plot <- intersect(params_to_plot, names(posterior))

# Convert wide → long format
priors_long <- priors %>%
  select(all_of(params_to_plot)) %>%
  pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") %>%
  mutate(type = "Prior")

posterior_long <- posterior %>%
  select(all_of(params_to_plot)) %>%
  pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") %>%
  mutate(type = "Posterior")

df_all <- bind_rows(priors_long, posterior_long)

# Limit sigma range for better visualization
df_all <- df_all |>
  mutate(value = ifelse(parameter == "sigma" & value > 10, 10, value))
```

```{r}
plot22 <- ggplot(df_all, aes(x = value, fill = type)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~parameter, scales = "free", ncol = 4) +
  theme_bw(base_size = 13) +
  labs(title = "Prior vs Posterior Distributions",
       x = "Parameter value", y = "Density") +
  scale_fill_manual(values = c("gray70", "steelblue"))
```

The prior and posterior distributions for the hyperparameters are shown in @fig-post-vs-prior-hyper. The priors for $\gamma_1$ and $\gamma_2$ are relatively flat across their ranges, while the posterior distributions are sharply peaked around values slightly below and above zero, respectively, indicating that the model estimated these parameters precisely. A similar pattern is observed for the $\mu$ hyperparameters, where the posterior distributions are much narrower than their priors, suggesting strong information gained from the data. On the other hand, the posterior distributions for the precision hyperparameters are right-skewed with higher means than the priors, implying that the model inferred greater precision (lower variance) than initially assumed.

# Results

```{r}
WT_results <- data.frame(
  root = character(),
  A = numeric(),
  B = numeric(),
  xmid = numeric(),
  scal = numeric(),
  beta = numeric(),
  beta_p = numeric(),
  beta_sig = character(),
  stringsAsFactors = FALSE
)

for(name in names(WT_model_output)){
  model <- WT_model_output[[name]]
  
  coef_info <- tidy(model)
  beta_row <- coef_info[coef_info$term == "beta", ]

  WT_results <- rbind(
    WT_results,
    data.frame(
      root = name,
      A = coef(model)["A"],
      B = coef(model)["B"],
      xmid = coef(model)["xmid"],
      scal = coef(model)["scal"],
      beta = coef(model)["beta"],
      beta_p = beta_row$p.value,
      beta_sig = ifelse(beta_row$p.value < 0.05, "*", "ns")
    )
  )
}

table1 <- kable(WT_results, digits = 5, caption = "Wild Type Logistic Model Results")
```

```{r}
MU_results <- data.frame(
  root = character(),
  A = numeric(),
  B = numeric(),
  xmid = numeric(),
  scal = numeric(),
  beta = numeric(),
  beta_p = numeric(),
  beta_sig = character(),
  stringsAsFactors = FALSE
)

for(name in names(MU_model_output)){
  model <- MU_model_output[[name]]
  
  coef_info <- tidy(model)
  beta_row <- coef_info[coef_info$term == "beta", ]

  MU_results <- rbind(
    MU_results,
    data.frame(
      root = name,
      A = coef(model)["A"],
      B = coef(model)["B"],
      xmid = coef(model)["xmid"],
      scal = coef(model)["scal"],
      beta = coef(model)["beta"],
      beta_p = beta_row$p.value,
      beta_sig = ifelse(beta_row$p.value < 0.05, "*", "ns")
    )
  )
}

table2 <- kable(MU_results, digits = 5, caption = "Mutant Logistic Model Results")
```

The frequentist model results for the wild type (Table 1) show that only roots 2 and 7 have insignificant $\beta$ estimates, indicating no measurable difference between inner and outer cell lengths for these samples. All other roots have significant negative $\beta$ values, suggesting that outer cell elongation begins earlier than inner cell elongation. In contrast, the mutant roots (Table 2) show fewer statistically significant $\beta$ estimates. Roots 1, 4, 5, 6, 7, and 9 exhibit significant positive or negative $\beta$ values, indicating that while differences exist between inner and outer measurements, the direction of these differences varies across roots. The remaining mutant roots have insignificant $\beta$ estimates, implying little to no measurable difference between sides. The frequentist results ultimately give statistically significant evidence that the outer cell measurements are larger for the wild type and that there is not convincing evidence for a clear one-sided discrepancy for mutant type.

```{r}
post_summary <-summary(sim)
ps=post_summary[[1]]
```

```{r}
library(DT)

# Mean
means_WT = c()
means_MU = c()

# A, alpha, B, scal
for(i in 0:3){
  s = i*20
  if(i == 3){
    s = s + 10
  }
  means_WT = c(means_WT, mean(ps[s+1], ps[s+12], ps[s+14], ps[s+15], ps[s+16], ps[s+17], ps[s+18], ps[s+19], ps[s+20]))
  means_MU = c(means_MU, mean(ps[s+2], ps[s+3], ps[s+4], ps[s+5], ps[s+6], ps[s+7], ps[s+8], ps[s+9], ps[s+10], ps[s+11], ps[s+13]))
}

# SD
sd_WT = c()
sd_MU = c()

# A, alpha, B, scal
for(i in 0:3){
  s = i*20 + 93
  if(i == 3){
    s = s + 10
  }
  sd_WT = c(sd_WT, mean(ps[s+1], ps[s+12], ps[s+14], ps[s+15], ps[s+16], ps[s+17], ps[s+18], ps[s+19], ps[s+20]))
  sd_MU = c(sd_MU, mean(ps[s+2], ps[s+3], ps[s+4], ps[s+5], ps[s+6], ps[s+7], ps[s+8], ps[s+9], ps[s+10], ps[s+11], ps[s+13]))
}

ps_df <- data.table(
  type = c("WT", "MU"),
  A_mean = c(means_WT[1], means_MU[1]), 
  A_sd = c(sd_WT[1], sd_MU[1]),
  alpha_mean = c(means_WT[2], means_MU[2]),
  alpha_sd = c(sd_WT[2], sd_MU[2]),
  B_mean = c(means_WT[3], means_MU[3]),
  B_sd = c(sd_WT[3], sd_MU[3]),
  scal_mean = c(means_WT[4], means_MU[4]),
  scal_sd = c(sd_WT[4], sd_MU[4])
)

indices = c(61, 62, 63, 64, 65, 66, 67, 68, 69, 70, 91)
means_alt = c()
sd_alt = c()
naive_se_alt = c()
ts_se_alt = c()

for(idx in indices){
  means_alt = c(means_alt, ps[idx])
  sd_alt = c(sd_alt, ps[91 + idx])
  naive_se_alt = c(naive_se_alt, ps[91*2 + idx])
  ts_se_alt = c(ts_se_alt, ps[91*3 + idx])
}

ps_q = post_summary[[2]]

quant_2_5_alt = c()
quant_25_alt = c()
quant_50_alt = c()
quant_75_alt = c()
quant_97_5_alt = c()

for(idx in indices){
  quant_2_5_alt = c(quant_2_5_alt, ps_q[idx])
  quant_25_alt = c(quant_25_alt, ps_q[91 + idx])
  quant_50_alt = c(quant_50_alt, ps_q[91*2 + idx])
  quant_75_alt = c(quant_75_alt, ps_q[91*3 + idx])
  quant_97_5_alt = c(quant_97_5_alt, ps_q[91*4 + idx])
}

alternate_params_df <- data.table(
  parameter = c("gamma1", "gamma2", "mu_A", "mu_d", "mu_eta_s", "mu_z0", "prec_a", "prec_d", "prec_eta_s", "prec_z0", "sigma"),
  mean = means_alt, 
  sd = sd_alt,
  time_series_se = ts_se_alt,
  quantile_2_5 = quant_2_5_alt,
  quantile_97_5 = quant_97_5_alt
)

table7 <- kable(alternate_params_df)

table8 <- kable(ps_df)
```

From the Bayesian model, the posterior summary for the hyperparameters (Table 7) shows mean estimates of −0.1494 for $\gamma_1$ and 0.1738 for $\gamma_2$. On the \[0, 1\] scale, this implies that outer cells in wild type roots have an xmid value 0.1494 lower than inner cells—indicating earlier elongation onset—while in mutant roots, the outer cells have only a 0.0244 higher xmid than the inner cells. Thus, outer cells begin elongation earlier in wild types, but this difference is greatly reduced in mutants.

The posterior summary for the per-root parameters (Table 8) indicates that wild type cells begin elongation closer to the root tip than mutant cells, with mean A values of 8.63 and 5.80, respectively. Wild type roots also reach greater final cell lengths, as reflected in higher mean B values (30.83 vs. 25.37). The mean scale parameter is slightly larger for mutants (0.232 vs. 0.221), suggesting a more gradual elongation transition.

Overall, the Bayesian model indicates that wild type roots begin elongation at higher initial values, transition more sharply, and reach greater final lengths, while mutant roots grow more slowly and smoothly. Notably, outer cells in wild type roots elongate earlier than inner cells, but this inner–outer contrast is largely reduced in mutants.

```{r}
get_freq_preds <- function(model_list, data, geno_label) {
  map2_dfr(
    .x = model_list,
    .y = names(model_list),
    ~ {
      root_id <- str_replace(.y, "root_", "")
      d <- data |> filter(root == root_id)
      d |>
        mutate(Freq_Pred = fitted(.x),
               Genotype = geno_label)
    }
  )
}

freq_preds <- bind_rows(
  get_freq_preds(WT_model_output, rootData_WT, "WT"),
  get_freq_preds(MU_model_output, rootData_MU, "MU")
)

pm <- jags.out$BUGSoutput$mean

A_vec     <- pm$A
B_vec     <- pm$B
alpha_vec <- pm$alpha
scal_vec  <- pm$scal

root_idx <- as.numeric(as.factor(rootData$root))              # 1..n root
side_num <- ifelse(rootData$side == "Outer", 1, 0)            # 0 = inner, 1 = outer
geno_num <- ifelse(rootData$Genotype == "MU", 1, 0)           # match bdat: MU = 1

A_i     <- A_vec[root_idx]
B_i     <- B_vec[root_idx]
alpha_i <- alpha_vec[root_idx]
scal_i  <- scal_vec[root_idx]
gamma1 <- as.numeric(pm$gamma1)
gamma2 <- as.numeric(pm$gamma2)

xmid_eff <- alpha_i + gamma1 * side_num + gamma2 * geno_num

# rebuild Bayesian prediction: note we use x01, not Midline
Bayes_Pred <- A_i + (B_i - A_i) /
  (1 + exp((xmid_eff - rootData$x01) / scal_i))

bayes_preds <- rootData |>
  mutate(Bayes_Pred = Bayes_Pred) |>
  select(root, Midline, side, Genotype, x01, Bayes_Pred)

# join both prediction sources to original data
plot_data <- rootData |>
  left_join(
    freq_preds |> select(root, Midline, side, Genotype, Freq_Pred),
    by = c("root", "Midline", "side", "Genotype")
  ) |>
  left_join(
    bayes_preds,
    by = c("root", "Midline", "side", "Genotype", "x01")
  )

plot_long <- plot_data |>
  select(root, x01, Length, side, Genotype, Freq_Pred, Bayes_Pred) |>
  pivot_longer(
    cols = c(Freq_Pred, Bayes_Pred),
    names_to = "Model",
    values_to = "Pred"
  ) |>
  mutate(Model = recode(Model,
                        "Freq_Pred" = "Frequentist",
                        "Bayes_Pred" = "Bayesian"),
         Model = factor(Model, levels = c("Frequentist", "Bayesian")))

plot_long_avg <- plot_long |>
  group_by(Genotype, Model, side, x01) |>
  summarise(
    Length_mean = mean(Length, na.rm = TRUE),
    Pred_mean   = mean(Pred, na.rm = TRUE),
    Pred_sd     = sd(Pred, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(Genotype, Model, side, x01)
```

```{r}
p_MU_facet <- plot_long %>%
  filter(Genotype == "MU") %>%
  ggplot(aes(x = x01, color = side)) +
  # show actual data 
  stat_summary(aes(y = Length), fun = mean, geom = "point", alpha = 0.25, size = 0.3) +
  # show model predictions
  stat_summary(aes(y = Pred), fun = mean, geom = "line", size = 0.4) +
  facet_grid(root ~ Model) +   # rows = roots, cols = Frequentist/Bayesian
  labs(
    title = "Predicted Growth Curves by Root – Mutant (MU)",
    x = "Normalized Position (x01)",
    y = "Cell Length (µm)",
    color = "Side"
  ) +
  theme_minimal(base_size = 6) +
  theme(
    strip.text = element_text(size = 6),
    panel.spacing = unit(0.5, "lines")
  )

p_WT_facet <- plot_long %>%
  filter(Genotype == "WT") %>%
  ggplot(aes(x = x01, color = side)) +
  stat_summary(aes(y = Length), fun = mean, geom = "point", alpha = 0.25, size = 0.3) +
  stat_summary(aes(y = Pred), fun = mean, geom = "line", size = 0.4) +
  facet_grid(root ~ Model) +
  labs(
    title = "Predicted Growth Curves by Root – Wild Type (WT)",
    x = "Normalized Position (x01)",
    y = "Cell Length (µm)",
    color = "Side"
  ) +
  theme_minimal(base_size = 6) +
  theme(
    strip.text = element_text(size = 6),
    panel.spacing = unit(0.5, "lines")
  )

p_MU_facet_small <- p_MU_facet +
  scale_x_continuous(breaks = c(0, 0.5, 1)) +
  theme(
    axis.text.x = element_text(size = 5),
    axis.title.x = element_text(size = 6)
    
  )

p_WT_facet_small <- p_WT_facet +
  scale_x_continuous(breaks = c(0, 0.5, 1)) +
  theme(
    axis.text.x = element_text(size = 5),
    axis.title.x = element_text(size = 6),
    legend.position = "none"
  )

# put them next to each other
combined <- p_WT_facet_small + plot_spacer() + p_MU_facet_small +
  plot_layout(ncol = 3, widths = c(1, 0.07, 1))
```

```{r}
# Compute Bayesian residuals
Bayes_residuals <- rootData |>
  mutate(
    Bayes_Fitted = Bayes_Pred,
    Bayes_Residual = Length - Bayes_Pred
  ) |>
  select(root, side, Genotype, x01, Midline, Bayes_Fitted, Bayes_Residual)

# Filter by genotype within each plot
plot_bayes_fit_wt <- Bayes_residuals %>%
  filter(Genotype == "WT") %>%
  ggplot(aes(x = Bayes_Fitted, y = Bayes_Residual, color = side)) +
  geom_point(size= 0.6, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (WT)",
       x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

plot_bayes_mid_wt <- Bayes_residuals %>%
  filter(Genotype == "WT") %>%
  ggplot(aes(x = Midline, y = Bayes_Residual, color = side)) +
  geom_point(size= 0.6, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (WT)",
       x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()

plot_bayes_fit_mu <- Bayes_residuals %>%
  filter(Genotype == "MU") %>%
  ggplot(aes(x = Bayes_Fitted, y = Bayes_Residual, color = side)) +
  geom_point(size= 0.6, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted (MU)",
       x = "Fitted Cell Length", y = "Residuals") +
  theme_minimal()

plot_bayes_mid_mu <- Bayes_residuals %>%
  filter(Genotype == "MU") %>%
  ggplot(aes(x = Midline, y = Bayes_Residual, color = side)) +
  geom_point(size= 0.6, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Midline (MU)",
       x = "Distance Along Midline", y = "Residuals") +
  theme_minimal()
```

```{r}
Bayes_residuals <- Bayes_residuals %>%
  mutate(root = factor(root, 
                       levels = c(paste0("WT", 1:9), paste0("MU", 1:11))))

p_all_box_bayes <- ggplot(Bayes_residuals, aes(x = root, y = Bayes_Residual, fill = side)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  labs(title = "Residuals by Root (Bayesian Model)",
       x = "Root ID", y = "ResidualS", fill = "Side") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )
```

The frequentist and Bayesian models performed similarly in their ability to predict cell length for both mutant and wild type roots (@fig-freq-vs-bayes). Both models captured the lack of discrepancy between inner and outer side measurements for the mutant type, while a clearer separation between the two sides was observed for the wild type. In particular, for wild type roots such as WT4 and WT8, the distance between the outer and inner side predicted cell length curves is slightly larger in the frequentist model compared to the Bayesian model.

# Conclusion

Overall, both the frequentist and hierarchical Bayesian analyses consistently support the hypothesis that wild type roots exhibit greater asymmetry in cell elongation compared to the mutant roots. The frequentist models revealed that most WT roots had significantly negative $\beta$ estimates, indicating earlier elongation on the other side relative to the inner side. In contrast, MU roots displayed fewer significant $\beta$ effects, with both positive and negative signs, suggesting that elongation differences between inner and outer sides are weaker and less consistent.

The Bayesian hierarchical model reinforced these trends at the population level. Posterior summaries of $\gamma_1$ and $\gamma_2$ showed that in wild type roots, outer cells have smaller inflection points and therefore begin elongation earlier, whereas this contrast is diminished in mutant roots. Additionally, wild type roots exhibited higher baseline (A) and asymptotic (B) cell lengths and smaller scale (scal) values than mutant roots, indicating longer initial and final cell lengths and a sharper elongation transition.

# Limitations

The logistic model assumes symmetric sigmoidal growth, which may oversimplify local variations along the root. Additionally, the current models treat sides as fixed effects, ignoring potential spatial correlations among adjacent cells and temporal changes in elongation. We also allowed only the midpoint (xmid) to vary between inner vs. outer and wild type vs. mutant roots, whereas other parameters may differ as well. Future analyses could incorporate mixed-effects structures to better capture within-root dependence, and spline-based models to represent curvature-driven or non-sigmoidal growth dynamics. Expanding the sample size and integrating imaging-based covariates (e.g., curvature angle, local growth rate) would further strengthen inference on genotype-specific elongation patterns.

## TODO: limitations regarding mixing, ess, priors??

\pagebreak

# Appendix

```{r}
#| tbl-cap: "Summary of wild type frequentist results."
table1
```

```{r}
#| tbl-cap: "Summary of mutant type frequentist results."
table2
```

```{r}
#| tbl-cap: "Top 10 Raftery-Lewis Burn-In and Dependence Factor Results."
table5
```

```{r}
#| tbl-cap: "Effective Sample Size Hyperparameters."
table6
```

```{r}
#| tbl-cap: "Posterior summary of parameters"
table7
```

```{r}
#| tbl-cap: "Posterior summary for Wildtype and Mutant Roots"
table8
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-dist
#| fig-cap: "Distribution of cell length for inner and outer cells."
plot1
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-wt
#| fig-cap: "Distribution of cell length for each wild root."
plot2
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-mu
#| fig-cap: "Distribution of cell length for each mutant root."
plot3
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-smooth-wt
#| fig-cap: "Distribution of cell length for each wild root."
plot4
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-per-root-smooth-mu
#| fig-cap: "Distribution of cell length for each mutant root."
plot5
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-smooth-wt
#| fig-cap: "Distribution of cell length for wild type."
plot6
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-cell-length-smooth-mu
#| fig-cap: "Distribution of cell length for mutant type."
plot7
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-mean-cell-length-per-root-wt
#| fig-cap: "Distribution of mean cell length for each wild root."
plot10
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-mean-cell-length-per-root-mu
#| fig-cap: "Distribution of mean cell length for each mutant root."
plot11
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-resid-freq
#| fig-cap: "Residuals Plots for Frequentist Model"

(plot12 + plot13) / (plot15 + plot16) +
  plot_layout(guides = "collect") +       
  theme(legend.position = "right")
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-resid-by-root
#| fig-cap: "Residuals by Root for Frequentist Model."
p_all_box
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-freq-acf
#| fig-cap: "Frequentist Model acf plot."
acf(resid(m0))
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-gelman-rubin
#| fig-cap: "Gelman-Rubin top 20 plot."
plot19
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-ess
#| fig-cap: "Effective Sample Size for A, B, Alpha, and Scal"
plot20
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-trace-plots
#| fig-cap: "Traceplots for hyperparameters."
par(mfrow = c(3, 4), mar = c(4, 4, 2, 1))
for (param in c("sigma", "prec_a", "prec_d", "prec_z0", "prec_eta_s",
                "mu_a", "mu_d", "mu_z0", "mu_eta_s",
                "gamma1", "gamma2")) {
  traceplot(sim[, param], main = param, col = 1:4)
}
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-post-vs-prior-hyper
#| fig-cap: "Posterior vs. Prior Distributions for hyperparameters."
plot22
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-resid-bayes
#| fig-cap: "Residuals Plots for Bayesian Model."

(plot_bayes_fit_wt + plot_bayes_mid_wt) /
(plot_bayes_fit_mu + plot_bayes_mid_mu) +
  plot_layout(guides = "collect") +
  theme(legend.position = "right")
```

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-align: 'center'
#| label: fig-resid-by-root-bayes
#| fig-cap: "Residuals by Root for Bayesian Model."

p_all_box_bayes
```

```{r}
#| fig-width: 6
#| fig-height: 6
#| fig-align: 'center'
#| label: fig-freq-vs-bayes
#| fig-cap: "Predicted Growth Curves by Root "

combined +
  plot_layout(guides = "collect") +
  theme(legend.position = "right")
```
